### 2025-03-02 paper/main.tex
set up paper according to JMLR template, adjusted it to fit the seminar

\documentclass[twoside,11pt]{article}

\usepackage{blindtext}

% Any additional packages needed should be included after jmlr2e.
% Note that jmlr2e.sty includes epsfig, amssymb, natbib and graphicx,
% and defines many common macros, such as 'proof' and 'example'.
%
% It also sets the bibliographystyle to plainnat; for more information on
% natbib citation styles, see the natbib documentation, a copy of which
% is archived at http://www.jmlr.org/format/natbib.pdf

% Available options for package jmlr2e are:
%
%   - abbrvbib : use abbrvnat for the bibliography style
%   - nohyperref : do not load the hyperref package
%   - preprint : remove JMLR specific information from the template,
%         useful for example for posting to preprint servers.
%
% Example of using the package with custom options:
%
% \usepackage[abbrvbib, preprint]{jmlr2e}

\usepackage{jmlr2e}

% Definitions of handy macros can go here

\newcommand{\dataset}{{\cal D}}
\newcommand{\fracpartial}[2]{\frac{\partial #1}{\partial  #2}}

% Heading arguments are {volume}{year}{pages}{date submitted}{date published}{paper id}{author-full-names}

\usepackage{lastpage}
\tabmlheading{WS 2024/25}{1-\pageref{LastPage}}{15.03.2025}{}{Simon Stürzebecher}

% Short headings should be running head and authors last names

\ShortHeadings{Extending EAGGA to Neural Networks}{Simon Stürzebecher}
\firstpageno{1}

\begin{document}

\title{Extending EAGGA to Neural Networks}

\author{\name Simon Stürzebecher \email simon.stuerzebecher@campus.lmu.de}

%\editor{My editor}

\maketitle

\begin{abstract}%   <- trailing '%' for backward compatibility of .sty file
\blindtext
\end{abstract}

\begin{keywords}
  keyword one, keyword two, keyword three
\end{keywords}

\section{Introduction}

\blindmathpaper

Here is a citation \cite{EAGGA}.

% Manual newpage inserted to improve layout of sample file - not
% needed in general before appendices/bibliography.

\newpage

\appendix
\section{}
\label{app:theorem}

% Note: in this sample, the section number is hard-coded in. Following
% proper LaTeX conventions, it should properly be coded as a reference:

%In this appendix we prove the following theorem from
%Section~\ref{sec:textree-generalization}:

In this appendix we prove the following theorem from
Section~6.2:

\noindent
{\bf Theorem} {\it Let $u,v,w$ be discrete variables such that $v, w$ do
not co-occur with $u$ (i.e., $u\neq0\;\Rightarrow \;v=w=0$ in a given
dataset $\dataset$). Let $N_{v0},N_{w0}$ be the number of data points for
which $v=0, w=0$ respectively, and let $I_{uv},I_{uw}$ be the
respective empirical mutual information values based on the sample
$\dataset$. Then
\[
	N_{v0} \;>\; N_{w0}\;\;\Rightarrow\;\;I_{uv} \;\leq\;I_{uw}
\]
with equality only if $u$ is identically 0.} \hfill\BlackBox

\section{}

\noindent
{\bf Proof}. We use the notation:
\[
P_v(i) \;=\;\frac{N_v^i}{N},\;\;\;i \neq 0;\;\;\;
P_{v0}\;\equiv\;P_v(0)\; = \;1 - \sum_{i\neq 0}P_v(i).
\]
These values represent the (empirical) probabilities of $v$
taking value $i\neq 0$ and 0 respectively.  Entropies will be denoted
by $H$. We aim to show that $\fracpartial{I_{uv}}{P_{v0}} < 0$....\\

{\noindent \em Remainder omitted in this sample. See http://www.jmlr.org/papers/ for full paper.}

\vskip 0.2in
\bibliography{references}

\end{document}

### 2025-03-02 code/classes.py
add crowding distance as tie breaker in population sorting

@@ -586,18 +586,26 @@ class EAGGA:
                 if datetime.now() >= time_start + timedelta(seconds=self.secs_total):
                     break  # don't train any more individuals if time ran out
             
+            # create metrics list for computing non-dom-sort ranks + crowding distances
             metrics = [(
                 np.mean(individual['metrics']['performance']['auc']).item(),
                 individual['metrics']['performance']['nf'],
                 individual['metrics']['performance']['ni'],
                 individual['metrics']['performance']['nnm'],
-            ) for individual in self.population] + [self.performance_majority_predictor]  # majority class predictor only used for non dominated sorting, won't even be looked at in loop over self.population below where we add pareto front ranks to the population because we add it at position len(self.population) in the performances to be ranked here
-            metrics = [(1 - metric[0], *metric[1:]) for metric in metrics]  # compute pareto fronts w.r.t. reference (worst) point (0, 1, 1, 1)
-            ranks_nds = self.nds.do(np.array(metrics), return_rank=True)[1]
-            for i in range(len(self.population)):  # majority class predictor would be at ranks_nds[len(self.population)] -> no impact here, as intended
-                self.population[i]['rank_nds'] = ranks_nds[i].item()
-            self.population = sorted(self.population, key=lambda individual: individual['rank_nds'])[:self.hps['mu']]  # ascending order, lower ranks are better, choose best mu individuals
+            ) for individual in self.population]
+            metrics_nds = metrics + [self.performance_majority_predictor]  # majority class predictor only used for non dominated sorting, won't be assigned to an individual in loop over self.population below where we add front ranks + cds to the population
+            metrics_nds = [(1 - metric[0], *metric[1:]) for metric in metrics_nds]  # computing fronts assumes minimisation objective -> reference (worst) point (0, 1, 1, 1) -> inverse AUC sign
             
+            ranks_nds = self.nds.do(np.array(metrics_nds), return_rank=True)[1]
+            cds = calc_crowding_distance(np.array(metrics))
+
+            for i in range(len(self.population)):  # majority class predictor would be at ranks_nds[len(self.population)] -> no impact here, as intended (iter from 0 to len(self.population) - 1)
+                self.population[i]['rank_nds'] = ranks_nds[i].item()
+                self.population[i]['cd'] = cds[i].item()
[…]

### 2025-03-02 code/classes.py
add cuda device, add model + datasets to device

@@ -84,10 +84,10 @@ class WeightClipper:
 
 
 class Dataset(torch.utils.data.Dataset):
-    def __init__(self, X: pd.DataFrame, y: pd.Series, class_pos: str, group_structure):
-        self.X = torch.tensor(X.values, dtype=torch.float)
+    def __init__(self, X: pd.DataFrame, y: pd.Series, class_pos: str, group_structure, device):
+        self.X = torch.tensor(X.values, dtype=torch.float).to(device)
 
-        self.y = torch.zeros(len(y.index))
+        self.y = torch.zeros(len(y.index)).to(device)
         self.y[y.reset_index(drop=True) == class_pos] = 1
         
         self.feature_groups = group_structure.get_included_groups_features()
@@ -501,6 +501,8 @@ class Prob:
 
 class EAGGA:
     def __init__(self, oml_dataset, class_positive, hps: dict[str, tuple | int | float], batch_size: int, patience: int, secs_per_fold: int, secs_total: int, file_path: str = None):
+        self.device = torch.device('cuda' if torch.cuda.is_available() else 'cpu')
+        
         self.data, _, self.categorical_indicator, _ = oml_dataset.get_data()
         self.class_column = oml_dataset.default_target_attribute
         self.class_positive = class_positive
@@ -642,13 +644,15 @@ class EAGGA:
                 X=data_train.loc[:, data_train.columns != self.class_column],
                 y=data_train.loc[:, self.class_column],
                 class_pos=self.class_positive,
-                group_structure=group_structure
+                group_structure=group_structure,
+                device=self.device
             )
             dataset_stop_early = Dataset(
                 X=data_stop_early.loc[:, data_stop_early.columns != self.class_column],
                 y=data_stop_early.loc[:, self.class_column],
                 class_pos=self.class_positive,
-                group_structure=group_structure
+                group_structure=group_structure,
+                device=self.device
             )
 
             data_val = self.data_train_val.loc[indices_val, :]
@@ -656,7 +660,8 @@ class EAGGA:
                 X=data_val.loc[:, data_val.columns != self.class_column],
                 y=data_val.loc[:, self.class_column],
                 class_pos=self.class_positive,
-                group_structure=group_structure
+                group_structure=group_structure,
+                device=self.device
             )
 
             model = NeuralNetwork(
@@ -664,7 +669,7 @@ class EAGGA:
                 output_size=1,  # we only use binary datasets
[…]

### 2025-03-02 code/classes.py
gracefully handle generation of featureless group structures by discarding them

@@ -841,9 +841,11 @@ class EAGGA:
             for child in [child_1, child_2]:  # mutate
                 if Prob.should_do(Prob.p_gga_mutate_overall):
                     child['group_structure'] = GroupStructure.gga_mutate(child['group_structure'])
-
-            offspring.append(child_1)
-            offspring.append(child_2)
+            
+            # add child to offspring if not featureless
+            for child in [child_1, child_2]:
+                if child['group_structure'].get_number_of_included_groups() > 0:
+                    offspring.append(child)
 
         return offspring[:self.hps['lambda']]  # in case offsping is longer than lambda (while-loop always adds 2 individuals), discard offspring after lambda
     

### 2025-03-02 code/classes.py
fixed conceptual mistake in early stopping, which would have required saving 'patience' model parameters, now simply return epoch of early stopping == True

@@ -722,27 +722,24 @@ class EAGGA:
                     if monotonicity_constraint == 1:
                         model.networks[i].apply(self.monotonicity_clipper)  # applies weight clipper recursively to network + its children, cf. https://pytorch.org/docs/stable/generated/torch.nn.Module.html#torch.nn.Module.apply
             
-            should_stop_early, stop_epoch = self.stop_early(loss_history_early_stopping, loss_fn, model, dataset_stop_early)
+            should_stop_early = self.stop_early(loss_history_early_stopping, loss_fn, model, dataset_stop_early)
             if should_stop_early:
-                return model, stop_epoch
+                return model, epoch
             epoch += 1
         
         return model, epoch
     
 
-    def stop_early(self, loss_history: list, loss_fn, model: NeuralNetwork, dataset_stop_early: Dataset) -> tuple:
+    def stop_early(self, loss_history: list, loss_fn, model: NeuralNetwork, dataset_stop_early: Dataset) -> bool:
         # stopping criterion: mean of 'patience' previous losses over [t-patience, t] < current loss at t+1
         # if True -> go back to min loss within [t-patience, t]
         loss = self.eval(loss_fn, model, dataset_stop_early, only_loss=True)['loss']
         loss_history.append(loss)
 
         if len(loss_history) > self.patience and np.mean(loss_history[-self.patience-1:-1]) < loss:
-            mask = np.ones_like(loss_history)
-            mask[:-self.patience-1] = np.inf
-            stop_epoch = np.argmin(mask * loss_history) + 1  # +1 to make it 1-based (as opposed to 0-based from np.argmin indexing)
-            print(f'Stop early: {np.mean(loss_history[-self.patience-1:-1])} < {loss}, epoch stop: {stop_epoch}')
-            return True, stop_epoch.item()
-        return False, -1
+            print(f'Stop early: {np.mean(loss_history[-self.patience-1:-1])} < {loss}')
+            return True
+        return False
     
 
     # cf. https://pytorch.org/tutorials/beginner/introyt/trainingyt.html

### 2025-03-02 code/classes.py
resolve f-string errors in sagemaker notebook

@@ -625,7 +625,7 @@ class EAGGA:
         total_layers = individual['total_layers']
         nodes_per_hidden_layer = individual['nodes_per_hidden_layer']
         group_structure = individual['group_structure']
-        print(f'Running {self.hps['cv_k']}-fold CV for individual: {total_layers} total layers, {nodes_per_hidden_layer} nodes per hidden layer, gs: {group_structure}')
+        print(f'Running {self.hps["cv_k"]}-fold CV for individual: {total_layers} total layers, {nodes_per_hidden_layer} nodes per hidden layer, gs: {group_structure}')
 
         metrics = {
             'performance': list(),
@@ -678,7 +678,7 @@ class EAGGA:
 
             metrics['performance'].append(self.eval(loss_fn, model, dataset_val))
             metrics['epochs'].append(stop_epoch)
-            print(f'Fold {fold + 1}/{self.cv.get_n_splits()} | trained for {stop_epoch} epochs | metrics: {metrics['performance'][-1]}')  # TODO: remove
+            print(f'Fold {fold + 1}/{self.cv.get_n_splits()} | trained for {stop_epoch} epochs | metrics: {metrics["performance"][-1]}')  # TODO: remove
 
         return {
             'performance': {

### 2025-03-02 code/classes.py
resolve np.pow error caused by older numpy version in sagemaker notebook

@@ -482,7 +482,7 @@ class Prob:
 
         draws_unif = np.random.uniform(low=0, high=1, size=samples)
         draws_trunc_geom = np.ceil(  # round with np.ceil, as support of trunc geom in (a, b], cf. https://en.wikipedia.org/wiki/Truncated_distribution
-            np.log(np.pow(1 - p, a) - draws_unif * (np.pow(1 - p, a) - np.pow(1 - p, b))) / np.log(1 - p),
+            np.log(np.power(1 - p, a) - draws_unif * (np.power(1 - p, a) - np.power(1 - p, b))) / np.log(1 - p),
         ).astype(int)
         
         return draws_trunc_geom

### 2025-03-02 code/classes.py
roc_auc_score requires tensors to be on cpu again, move them there

@@ -501,7 +501,8 @@ class Prob:
 
 class EAGGA:
     def __init__(self, oml_dataset, class_positive, hps: dict[str, tuple | int | float], batch_size: int, patience: int, secs_per_fold: int, secs_total: int, file_path: str = None):
-        self.device = torch.device('cuda' if torch.cuda.is_available() else 'cpu')
+        self.device_cpu = torch.device('cpu')
+        self.device_cuda = torch.device('cuda' if torch.cuda.is_available() else 'cpu')
         
         self.data, _, self.categorical_indicator, _ = oml_dataset.get_data()
         self.class_column = oml_dataset.default_target_attribute
@@ -645,14 +646,14 @@ class EAGGA:
                 y=data_train.loc[:, self.class_column],
                 class_pos=self.class_positive,
                 group_structure=group_structure,
-                device=self.device
+                device=self.device_cuda
             )
             dataset_stop_early = Dataset(
                 X=data_stop_early.loc[:, data_stop_early.columns != self.class_column],
                 y=data_stop_early.loc[:, self.class_column],
                 class_pos=self.class_positive,
                 group_structure=group_structure,
-                device=self.device
+                device=self.device_cuda
             )
 
             data_val = self.data_train_val.loc[indices_val, :]
@@ -661,7 +662,7 @@ class EAGGA:
                 y=data_val.loc[:, self.class_column],
                 class_pos=self.class_positive,
                 group_structure=group_structure,
-                device=self.device
+                device=self.device_cuda
             )
 
             model = NeuralNetwork(
@@ -669,7 +670,7 @@ class EAGGA:
                 output_size=1,  # we only use binary datasets
                 total_layers=total_layers,
                 nodes_per_hidden_layer=nodes_per_hidden_layer
-            ).to(self.device)
+            ).to(self.device_cuda)
 
             optimizer = torch.optim.AdamW(model.parameters())
             loss_fn = nn.BCEWithLogitsLoss()
@@ -770,8 +771,8 @@ class EAGGA:
             return res
 
         # compute AUC
-        predictions = torch.cat(batch_predictions)
-        targets = torch.cat(batch_targets)
+        predictions = torch.cat(batch_predictions).to(self.device_cpu)
+        targets = torch.cat(batch_targets).to(self.device_cpu)
         res['auc'] = roc_auc_score(targets, predictions)
         
         # compute interpretability metrics

### 2025-03-02 code/classes.py
add hypervolume computation + pareto front return to run_eagga

@@ -14,6 +14,7 @@ from sklearn.metrics import roc_auc_score
 
 from scipy.stats import spearmanr
 
+from pymoo.indicators.hv import HV
 from pymoo.operators.survival.rank_and_crowding.metrics import calc_crowding_distance
 from pymoo.util.nds.non_dominated_sorting import NonDominatedSorting
 
@@ -490,7 +491,6 @@ class Prob:
 
     @staticmethod
     def should_do(p: float):
-        return 1  # TODO: remove
         return np.random.uniform() <= p
     
 
@@ -530,6 +530,7 @@ class EAGGA:
         majority_class_fraction = self.data_train_val.loc[self.data_train_val.loc[:, self.class_column] == majority_class, :].shape[0] / self.data_train_val.shape[0]
         self.performance_majority_predictor = (majority_class_fraction, 0, 0, 0)
         self.nds = NonDominatedSorting()
+        self.hv = HV(ref_point=(1, 1, 1, 1), nds=False)  # set ref-pt to (1, 1, 1, 1), as pymoo always minimises, i.e. ref-pt (0, 1, 1, 1) would consider 0 to be largest value in first dim + result in hypervolume = 0, instead transform first dim of points so that auc becomes minimisation problem (via 1 - auc)
 
         # inner split, k-fold cross validation
         self.cv = StratifiedKFold(
@@ -575,6 +576,7 @@ class EAGGA:
         } for i in range(self.hps['mu'])]
 
 
+    # returns Pareto front
     def run_eagga(self):
         time_start = datetime.now()
         print(f'Start EAGGA at {time_start.isoformat()}')
@@ -598,8 +600,9 @@ class EAGGA:
             ) for individual in self.population]
             metrics_nds = metrics + [self.performance_majority_predictor]  # majority class predictor only used for non dominated sorting, won't be assigned to an individual in loop over self.population below where we add front ranks + cds to the population
             metrics_nds = [(1 - metric[0], *metric[1:]) for metric in metrics_nds]  # computing fronts assumes minimisation objective -> reference (worst) point (0, 1, 1, 1) -> inverse AUC sign
-            
-            ranks_nds = self.nds.do(np.array(metrics_nds), return_rank=True)[1]
+            metrics_nds_np = np.array(metrics_nds)
+
+            fronts, ranks_nds = self.nds.do(metrics_nds_np, return_rank=True)
             cds = calc_crowding_distance(np.array(metrics))
 
[…]

### 2025-03-02 code/classes.py
remove stop early print

@@ -745,7 +745,6 @@ class EAGGA:
         loss_history.append(loss)
 
         if len(loss_history) > self.patience and np.mean(loss_history[-self.patience-1:-1]) < loss:
-            print(f'Stop early: {np.mean(loss_history[-self.patience-1:-1])} < {loss}')
             return True
         return False
     

### 2025-03-02 code/classes.py
update stop early logic

@@ -611,8 +611,8 @@ class EAGGA:
             self.population = sorted(self.population, key=lambda individual: (individual['rank_nds'], -individual['cd']))[:self.hps['mu']]  # ascending order of front ranks (lower is better), descending order of crowding distance for tied ranks (larger is more important for front), choose best mu individuals
             
             pareto_front_idx = fronts[0]
-            self.pareto_front = metrics_nds_np[pareto_front_idx]
-            print(f'Dominated Hypervolume: {self.hv(self.pareto_front)} for Pareto front {self.pareto_front}')
+            self.pareto_front = np.subtract(metrics_nds_np[pareto_front_idx], (1, 0, 0, 0)) * (-1, 1, 1, 1)  # ensure that Pareto front format is (AUC, NF, NI, NNM) instead of (-AUC, NF, NI, NNM), which we have in metrics_nds_np
+            print(f'Dominated Hypervolume: {self.hv(self.pareto_front)} for Pareto front {np.subtract(self.pareto_front, (1, 0, 0, 0))}')
 
             if datetime.now() >= time_start + timedelta(seconds=self.secs_total):
                 self.offspring = list()  # re-set offspring so in case of json export the same individuals won't be saved as part of offspring (without metrics) and population (with metrics)
@@ -707,7 +707,8 @@ class EAGGA:
         loader_train = torch.utils.data.DataLoader(dataset_train, batch_size=self.batch_size, shuffle=True)
 
         epoch = 0
-        loss_history_early_stopping = list()
+        loss_history = list()
+        current_best = (0, model)  # (epoch, model at epoch)
         time_start = datetime.now()
         while(datetime.now() < time_start + timedelta(seconds=self.secs_per_fold)):
             model.train()  # training mode, put here as we switch to eval mode at the end of each epoch in def stop_early
@@ -730,23 +731,27 @@ class EAGGA:
                     if monotonicity_constraint == 1:
                         model.networks[i].apply(self.monotonicity_clipper)  # applies weight clipper recursively to network + its children, cf. https://pytorch.org/docs/stable/generated/torch.nn.Module.html#torch.nn.Module.apply
             
-            should_stop_early = self.stop_early(loss_history_early_stopping, loss_fn, model, dataset_stop_early)
+            should_stop_early, current_best = self.stop_early(loss_history, current_best, loss_fn, model, dataset_stop_early)
             if should_stop_early:
-                return model, epoch, True
[…]

### 2025-03-02 code/classes.py
add 1 to epoch print output (python 0 based, human-readable output should be 1 based)

@@ -686,7 +686,7 @@ class EAGGA:
 
             metrics['performance'].append(self.eval(loss_fn, model, dataset_val))
             metrics['epochs'].append(stop_epoch)
-            print(f'Fold {fold + 1}/{self.cv.get_n_splits()} | trained for {stop_epoch} epochs | stopped early: {stopped_early} | metrics: {metrics["performance"][-1]}')
+            print(f'Fold {fold + 1}/{self.cv.get_n_splits()} | trained for {stop_epoch + 1} epochs | stopped early: {stopped_early} | metrics: {metrics["performance"][-1]}')
 
         return {
             'performance': {

### 2025-03-02 code/classes.py
set mu,lambda,secs to 'prod' values, also ex- and import pareto fronts for each generation

@@ -895,12 +895,13 @@ class EAGGA:
 
         file_content = {
             'population': population,
-            'offspring': offspring
+            'offspring': offspring,
+            'pareto': self.pareto_front
         }
         with open(EAGGA.create_file_path(os.path.join(self.file_path, f'gen-{self.gen}.json')), 'w') as f:
             json.dump(file_content, f)
 
-        print('Saved population + offspring to file')
+        print('Saved population, offspring, pareto front to file')
     
 
     def load_population(self, gen):
@@ -914,8 +915,9 @@ class EAGGA:
             individual['group_structure'] = GroupStructure.from_dict(individual['group_structure'])
         for individual in self.offspring:
             individual['group_structure'] = GroupStructure.from_dict(individual['group_structure'])
+        self.pareto_front = file_content['pareto']
 
-        print('Loaded population + offspring from file, discarded previous population + offspring')
+        print('Loaded population, offspring, pareto front from file, discarded previous population, offspring, pareto front')
 
 
     @staticmethod

### 2025-03-03 code/classes.py
round all json outputs to max 5 decimals, also save all stop early losses to json

@@ -612,7 +612,7 @@ class EAGGA:
             
             pareto_front_idx = fronts[0]
             self.pareto_front = np.subtract(metrics_nds_np[pareto_front_idx], (1, 0, 0, 0)) * (-1, 1, 1, 1)  # ensure that Pareto front format is (AUC, NF, NI, NNM) instead of (-AUC, NF, NI, NNM), which we have in metrics_nds_np
-            print(f'Dominated Hypervolume: {self.hv(self.pareto_front)} for Pareto front {np.subtract(self.pareto_front, (1, 0, 0, 0))}')
+            print(f'Dominated Hypervolume: {self.hv(self.pareto_front)} for Pareto front {self.pareto_front}')
 
             if datetime.now() >= time_start + timedelta(seconds=self.secs_total):
                 self.offspring = list()  # re-set offspring so in case of json export the same individuals won't be saved as part of offspring (without metrics) and population (with metrics)
@@ -682,11 +682,12 @@ class EAGGA:
             optimizer = torch.optim.AdamW(model.parameters())
             loss_fn = nn.BCEWithLogitsLoss()
 
-            model, stop_epoch, stopped_early = self.train(optimizer, loss_fn, model, dataset_train, dataset_stop_early)
+            model, stop_epoch, stop_secs, stopped_early, losses_stop_early = self.train(optimizer, loss_fn, model, dataset_train, dataset_stop_early)
 
             metrics['performance'].append(self.eval(loss_fn, model, dataset_val))
+            metrics['performance'][-1]['losses_stop_early'] = losses_stop_early
             metrics['epochs'].append(stop_epoch)
-            print(f'Fold {fold + 1}/{self.cv.get_n_splits()} | trained for {stop_epoch + 1} epochs | stopped early: {stopped_early} | metrics: {metrics["performance"][-1]}')
+            print(f'Fold {fold + 1}/{self.cv.get_n_splits()} | trained for {stop_epoch + 1} epochs / {round(stop_secs.total_seconds(), 3)} seconds | stopped early: {stopped_early} | metrics: {metrics["performance"][-1]}')
 
         return {
             'performance': {
@@ -695,7 +696,7 @@ class EAGGA:
                 'ni': metrics['performance'][0]['ni'],
                 'nnm': metrics['performance'][0]['nnm'],
                 'auc': [performance['auc'] for performance in metrics['performance']],
-                'loss' : [performance['loss'] for performance in metrics['performance']]
[…]

### 2025-03-03 code/classes.py
enumerate self.offspring iterator to keep track of # of individual in logging

@@ -584,7 +584,8 @@ class EAGGA:
         while(datetime.now() < time_start + timedelta(seconds=self.secs_total)):
             print(f'Generation {self.gen+1}, evaluate {len(self.offspring)} individuals')
 
-            for individual in self.offspring:
+            for i, individual in enumerate(self.offspring):
+                print(f'Running {self.hps["cv_k"]}-fold CV for individual {i+1}/{len(self.offspring)}: {individual["total_layers"]} total layers, {individual["nodes_per_hidden_layer"]} nodes per hidden layer, gs: {individual["group_structure"]}')
                 individual['metrics'] = self.run_cv(individual)
                 self.population.append(individual)
 
@@ -633,7 +634,6 @@ class EAGGA:
         total_layers = individual['total_layers']
         nodes_per_hidden_layer = individual['nodes_per_hidden_layer']
         group_structure = individual['group_structure']
-        print(f'Running {self.hps["cv_k"]}-fold CV for individual: {total_layers} total layers, {nodes_per_hidden_layer} nodes per hidden layer, gs: {group_structure}')
 
         metrics = {
             'performance': list(),

### 2025-03-03 code/classes.py
remove loss history from logging output

@@ -685,9 +685,10 @@ class EAGGA:
             model, stop_epoch, stop_secs, stopped_early, losses_stop_early = self.train(optimizer, loss_fn, model, dataset_train, dataset_stop_early)
 
             metrics['performance'].append(self.eval(loss_fn, model, dataset_val))
+            # print prior to adding losses_stop_early to avoid long + rather uninformative output of loss history, only intended to be used for plotting
+            print(f'Fold {fold + 1}/{self.cv.get_n_splits()} | trained for {stop_epoch + 1} epochs / {round(stop_secs.total_seconds(), 3)} seconds | stopped early: {stopped_early} | metrics: {metrics["performance"][-1]}')
             metrics['performance'][-1]['losses_stop_early'] = losses_stop_early
             metrics['epochs'].append(stop_epoch)
-            print(f'Fold {fold + 1}/{self.cv.get_n_splits()} | trained for {stop_epoch + 1} epochs / {round(stop_secs.total_seconds(), 3)} seconds | stopped early: {stopped_early} | metrics: {metrics["performance"][-1]}')
 
         return {
             'performance': {

### 2025-03-03 code/classes.py
add tunable dropout to NN, add min_epochs before early_stopping will be considered (loop can still select an epoch prior to min_epochs but trains at least min_epochs before looking back)

@@ -27,7 +27,7 @@ import json
 
 
 class NeuralNetwork(nn.Module):
-    def __init__(self, group_structure, output_size, total_layers, nodes_per_hidden_layer):
+    def __init__(self, group_structure, output_size, total_layers, nodes_per_hidden_layer, p_dropout):
         super().__init__()
 
         self.group_structure = group_structure
@@ -42,11 +42,13 @@ class NeuralNetwork(nn.Module):
             # input layer
             modules.append(nn.Linear(len(features), nodes_per_hidden_layer))
             modules.append(nn.ReLU())
+            modules.append(nn.Dropout(p_dropout))
 
             # hidden layers
             for _ in range(1, total_layers-1):
                 modules.append(nn.Linear(nodes_per_hidden_layer, nodes_per_hidden_layer))
                 modules.append(nn.ReLU())
+                modules.append(nn.Dropout(p_dropout))
             
             network_for_group = nn.Sequential(*modules)
             self.networks.append(network_for_group)
@@ -462,6 +464,8 @@ class GroupStructure:
 
 class Prob:
     p_sample_hps = 0.5
+    gamma_shape = 2
+    gamma_scale = 0.2
     p_sample_features_selected = 0.5  # original paper uses relative # of features used across 10 decision trees, no straightforward to retrieve this from sklearn, for our NN implementation just use 0.5
     p_sample_interactions = 0.5  # original paper uses relative # of pairwise interactions used across 10 decision trees, no straightforward to retrieve this from sklearn, for our NN implementation just use 0.5
 
@@ -490,6 +494,26 @@ class Prob:
     
 
     @staticmethod
+    def r_trunc_gamma(shape: float, scale: float, samples: int, val_max: int = 1):
+        draws_trunc_gamma = np.random.gamma(shape=shape, scale=scale, size=samples)
+
+        idx = np.argwhere(draws_trunc_gamma > val_max)
+        while len(idx) > 0:
+            draws_trunc_gamma[idx] = np.random.gamma(shape=shape, scale=scale, size=samples)[idx]
+            idx = np.argwhere(draws_trunc_gamma > val_max)
+
+        return np.round(draws_trunc_gamma, 2)
+    
+
+    @staticmethod
+    def ea_mutate_gaussian(val, lower, upper):
+        val = (val - lower) / (upper - lower)
+        val = np.random.normal(loc=val, scale=0.1, size=1).item()
+        val = (val * (upper - lower)) + lower
[…]

### 2025-03-03 code/classes.py
do p dropout precision only up to 1 decimal, ensure that ea_mutate_Gaussian returns correct precision, adjust Gamma distribution parameters to have more weight on lower values and less on those > 0.5

@@ -464,8 +464,10 @@ class GroupStructure:
 
 class Prob:
     p_sample_hps = 0.5
+    
     gamma_shape = 2
-    gamma_scale = 0.2
+    gamma_scale = 0.15
+    
     p_sample_features_selected = 0.5  # original paper uses relative # of features used across 10 decision trees, no straightforward to retrieve this from sklearn, for our NN implementation just use 0.5
     p_sample_interactions = 0.5  # original paper uses relative # of pairwise interactions used across 10 decision trees, no straightforward to retrieve this from sklearn, for our NN implementation just use 0.5
 
@@ -494,7 +496,7 @@ class Prob:
     
 
     @staticmethod
-    def r_trunc_gamma(shape: float, scale: float, samples: int, val_max: int = 1):
+    def r_trunc_gamma(shape: float, scale: float, samples: int, val_max: int = 1, decimals: int = 1):
         draws_trunc_gamma = np.random.gamma(shape=shape, scale=scale, size=samples)
 
         idx = np.argwhere(draws_trunc_gamma > val_max)
@@ -502,15 +504,15 @@ class Prob:
             draws_trunc_gamma[idx] = np.random.gamma(shape=shape, scale=scale, size=samples)[idx]
             idx = np.argwhere(draws_trunc_gamma > val_max)
 
-        return np.round(draws_trunc_gamma, 2)
+        return np.round(draws_trunc_gamma, decimals)
     
 
     @staticmethod
-    def ea_mutate_gaussian(val, lower, upper):
+    def ea_mutate_gaussian(val, lower, upper, decimals: int = 1):
         val = (val - lower) / (upper - lower)
         val = np.random.normal(loc=val, scale=0.1, size=1).item()
         val = (val * (upper - lower)) + lower
-        return min(max(val, lower), upper)
+        return round(min(max(val, lower), upper), decimals)
             
 
     @staticmethod
@@ -581,7 +583,7 @@ class EAGGA:
         print('Starting init population')
         population_layers = Prob.r_trunc_geom(Prob.p_sample_hps, self.hps['mu'], self.hps['total_layers'][0], self.hps['total_layers'][1])
         population_nodes = Prob.r_trunc_geom(Prob.p_sample_hps, self.hps['mu'], self.hps['nodes_per_hidden_layer'][0], self.hps['nodes_per_hidden_layer'][1])
[…]

### 2025-03-03 code/classes.py
fix HV logging, forgot to inverse AUC when computing HV

@@ -642,7 +642,7 @@ class EAGGA:
             
             pareto_front_idx = fronts[0]
             self.pareto_front = np.subtract(metrics_nds_np[pareto_front_idx], (1, 0, 0, 0)) * (-1, 1, 1, 1)  # ensure that Pareto front format is (AUC, NF, NI, NNM) instead of (-AUC, NF, NI, NNM), which we have in metrics_nds_np
-            print(f'Dominated Hypervolume: {self.hv(self.pareto_front)} for Pareto front {self.pareto_front}')
+            print(f'Dominated Hypervolume: {self.hv(metrics_nds_np[pareto_front_idx])} for Pareto front {self.pareto_front}')
 
             if datetime.now() >= time_start + timedelta(seconds=self.secs_total):
                 self.offspring = list()  # re-set offspring so in case of json export the same individuals won't be saved as part of offspring (without metrics) and population (with metrics)

### 2025-03-03 paper/main.tex
add rough structure of the paper
add inital bullet points to paper

@@
 \begin{abstract}%   <- trailing '%' for backward compatibility of .sty file
-\blindtext
+%\blindtext
 \end{abstract}
@@
 \begin{keywords}
-  keyword one, keyword two, keyword three
+  tabular data, multi-objective optimization, deep learning
 \end{keywords}
@@
 
-\blindmathpaper
+%\blindmathpaper
+
+\section{Background and Related Works}
+
+\subsection{Tabular Data}
+- most common type of data
+- bad performance of deep learning models on tabular data
+
+% feels better in front of AutoML, as it feels more qualitative, less quantitative
+\subsection{Explainable Artificial Intelligence and Interpretability}
+\subsubsection{Why is interpretability desirable?}
+\subsubsection{Practical implementations}
+(1) post-hoc methods
+  - give examples
+  - why are they not perfectly suited
+(2) model-based
+
+\subsection{Automated Machine Learning (AutoML)}
+\subsubsection{Hyperparameter Optimization (HPO)}
+- grid + random search
+- Bayesian optimisation
+  * short intro
+  * why is it unsuitable for our task? (makes no sense to have it here, unless we introduce our problem (feature selection is high-dim binary space) in the introduction)
+- evolutionary algorithm
+  * CMA-ES
+  * NSGA-II
+    - non-dominated sorting
+    - crowding distance
+\subsubsection{Neural Architecture Search (NAS)}
+
+\subsection{Multi-Objective Optimization}
+- intro
+
+%likely shouldn't be an entire section
+\section{EAGGA}
+- NSGA-II inspired evolutionary / genetic algorithm
+- introduces group structure for more efficient optimization over the entire search space
+
+\section{Extension to neural networks}
+- why extend it to NNs?
+  * NNs notoriously uninterpretable due to complex transformation of the feature space
+  * due to (lin alg) non-linear activation functions, interactions can be modelled
+  * no monotonicity guarantees
+- our approach
+  * feature sparsity: straightforward, only train model on selected features, goes somewhat against idea of deep learning,
+    where the model is supposed to decide on its own, which features to put importance on -> WHY?
+  * feature interaction
+    - create non-fully connected network by introducing network-groups based on the equivalence relations introduced in the EAGGA paper
+    - show formula why the interacting features need to be grouped together when using ReLU (cf. photos)
+  * monotonicity constraints
+    - for groups with contraint on monotonicity: clip weights to [0, infty), bias clipping not necessary
+    - show formula of how monotonicity is enforced with this
+  * changes (-) + similarities (+) to EAGGA paper (~ in-between)
+    + 2/3 holdout outer split for 'CV'- vs test-set
+    - train + eval each individual via 5-fold CV with 20\% of train set again reserved to determine early stopping -> best to visualise,
+      * early stopping necessary (as opposed to XGB) as NN can be trained forever, whereas tree implementations used by EAGGA paper have
+        max depth, etc.
+      * early stopping using min-epochs 200, patience 100, i.e. train for $\ge$ 200 epochs, always safe model with lowest loss, whenever
+        current loss > mean(losses of last 100 (patience) epochs), stop early + return model with lowest loss
+    ~ if no early stopping, train each fold (in 5-fold CV) for max 120 secs, do entire EAGGA loop per dataset for max 8 hrs
+    ~ training on Sagemaker Notebook instance
+      * initial experiments on ml.g4dn.xlarge instance (2 vCPUs, 16GiB RAM, 1 NVIDIA T4, cf. https://aws.amazon.com/de/ec2/instance-types/g4/)
+      using cuda not much faster than on ml.t3.medium (2 Intel Xeon 8000 vCPUs, 4GiB RAM, cf. https://aws.amazon.com/de/ec2/instance-types/t3/)
+      * thus decided for more economic + ressourcenschonend t3.medium
+    - evaluation
+      * AUC + NF: straightforward
+      * NI: for each group with $\ge$ 2 features: sum from 1 to (\# features - 1), i.e ${\# features \choose 2}$,
+        then sum over all groups like this, then divide by p(p-1)/2, i.e. ${p \choose 2}$
+      * NNM: \# unconstrained features / \# all features
+    ~ feature + interaction detector: examination of decision trees not super straightforward in sklearn, hence simply use p=0.5
+    ? interaction detector (FAST)
+      * train logistic model on 80\% of data (only holdout train portion or all ??), evaluate on 20
+      * use mean accuracy (logistic metric) instead of RSS (LM metric)
+    + monotonicity detector as in paper, also use dec tree HPs from mlr3
+    ~ encoding of group structures similar, each feature will get its individual monotonicity sign from the detector, this is then fed to the dataset,
+      which then outputs the data according multiplied with the sign
+
+\section{Experimental Results and Discussion}
+
+\section{Conclusion and Future Outlook}
 

### 2025-03-04 code/classes.py
add dominated hv to json output

@@ -556,7 +556,7 @@ class EAGGA:
         majority_class_fraction = self.data_train_val.loc[self.data_train_val.loc[:, self.class_column] == majority_class, :].shape[0] / self.data_train_val.shape[0]
         self.performance_majority_predictor = (majority_class_fraction, 0, 0, 0)
         self.nds = NonDominatedSorting()
-        self.hv = HV(ref_point=(1, 1, 1, 1), nds=False)  # set ref-pt to (1, 1, 1, 1), as pymoo always minimises, i.e. ref-pt (0, 1, 1, 1) would consider 0 to be largest value in first dim + result in hypervolume = 0, instead transform first dim of points so that auc becomes minimisation problem (via 1 - auc)
+        self.hv_obj = HV(ref_point=(1, 1, 1, 1), nds=False)  # set ref-pt to (1, 1, 1, 1), as pymoo always minimises, i.e. ref-pt (0, 1, 1, 1) would consider 0 to be largest value in first dim + result in hypervolume = 0, instead transform first dim of points so that auc becomes minimisation problem (via 1 - auc)
 
         # inner split, k-fold cross validation
         self.cv = StratifiedKFold(
@@ -642,7 +642,8 @@ class EAGGA:
             
             pareto_front_idx = fronts[0]
             self.pareto_front = np.subtract(metrics_nds_np[pareto_front_idx], (1, 0, 0, 0)) * (-1, 1, 1, 1)  # ensure that Pareto front format is (AUC, NF, NI, NNM) instead of (-AUC, NF, NI, NNM), which we have in metrics_nds_np
-            print(f'Dominated Hypervolume: {self.hv(metrics_nds_np[pareto_front_idx])} for Pareto front {self.pareto_front}')
+            self.dhv = self.hv_obj(metrics_nds_np[pareto_front_idx])
+            print(f'Dominated Hypervolume: {self.dhv} for Pareto front {self.pareto_front}')
 
             if datetime.now() >= time_start + timedelta(seconds=self.secs_total):
                 self.offspring = list()  # re-set offspring so in case of json export the same individuals won't be saved as part of offspring (without metrics) and population (with metrics)
@@ -941,11 +942,12 @@ class EAGGA:
             individual['cd'] = round(individual['cd'], 5)
         for individual in offspring:
             individual['group_structure'] = individual['group_structure'].to_dict()
-
+        
         file_content = {
             'population': population,
             'offspring': offspring,
[…]

### 2025-03-04 code/classes.py
add logging

@@ -24,6 +24,14 @@ from datetime import datetime, timedelta
 import os
 from pathlib import Path
 import json
+import logging
+
+
+logging.basicConfig(filename=os.path.join('export', 'log.txt'),
+    filemode='a',
+    format='%(asctime)s,%(msecs)03d %(name)s %(levelname)s %(message)s',
+    datefmt='%Y-%m-%d %H:%M:%S',
+    level=logging.INFO)
 
 
 class NeuralNetwork(nn.Module):
@@ -535,7 +543,9 @@ class EAGGA:
         self.class_positive = class_positive
 
         if self.data.isnull().values.any():  # from manual checks: dataset 'jm1' has NaN values in 5 rows -> simply remove
-            print(f'Drop {self.data.loc[self.data.isnull().any(axis=1), :].shape[0]} NaN values of dataset {oml_dataset.name}')
+            msg = f'Drop {self.data.loc[self.data.isnull().any(axis=1), :].shape[0]} NaN values of dataset {oml_dataset.name}'
+            logging.info(msg)
+            print(msg)
             self.data = self.data.dropna(axis=0, how='any')
             self.data = self.data.reset_index(drop=True)
         
@@ -580,7 +590,9 @@ class EAGGA:
 
 
     def init_population(self):
-        print('Starting init population')
+        msg = 'Starting init population'
+        logging.info(msg)
+        print(msg)
         population_layers = Prob.r_trunc_geom(Prob.p_sample_hps, self.hps['mu'], self.hps['total_layers'][0], self.hps['total_layers'][1])
         population_nodes = Prob.r_trunc_geom(Prob.p_sample_hps, self.hps['mu'], self.hps['nodes_per_hidden_layer'][0], self.hps['nodes_per_hidden_layer'][1])
         population_p_dropout = Prob.r_trunc_gamma(Prob.gamma_shape, Prob.gamma_scale, self.hps['mu'], 1, 1)
@@ -590,7 +602,9 @@ class EAGGA:
         population_features_excluded = [list(set(all_features) - set(features_included)) for features_included in population_features_included]
         population_interactions = [GroupStructure.detector_interactions(self.data_train_val, features_included, self.class_column) for features_included in population_features_included]
         population_monotonicity_constraints = [GroupStructure.detector_monotonicity(self.data_train_val, groups_without_monotonicity, self.class_column) for groups_without_monotonicity in population_interactions]
-        print('Finished init population')
+        msg = 'Finished init population'
+        logging.info(msg)
[…]

### 2025-03-04 code/classes.py
add dataset name to log

@@ -535,6 +535,10 @@ class Prob:
 
 class EAGGA:
     def __init__(self, oml_dataset, class_positive, hps: dict[str, tuple | int | float], batch_size: int, min_epochs: int, patience: int, secs_per_fold: int, secs_total: int, file_path: str = None):
+        msg = f'Dataset {oml_dataset.name}'
+        logging.info(msg)
+        print(msg)
+        
         self.device_cpu = torch.device('cpu')
         self.device_cuda = torch.device('cuda' if torch.cuda.is_available() else 'cpu')
         

### 2025-03-05 paper/main.tex
restructure sections a little, add some more bullets

@@
 \section{Introduction}
+Tabular Data
+- most common type of data
+- bad performance of deep learning models on tabular data, recent research suggests that heavy regularisation is necessary
+
+regularisation
+- LM: ridge (L2) + lasso (L1)
+  * ridge: overcome multi-collinearity by pulling coeffs close to 0
+  * lasso: variable selection by setting coeffs = 0
+- NN:
+  * L2 regularisation implicitly included in SGD when using weight decay (pull params towards 0) -> does this combat multi-collinearity as well or serve another purpose?
+  * L1
+    + key claim of DL: feature selection obsolete, network finds important ones itself
+    + still, feature usually always has some impact (back this up), no explicit feature selection
+    + explicit feature selection simply by not inluding a feature
+  * another classical NN regularisation technique: dropout against co-adaptation
+  * early stopping
 
@@
 \section{Background and Related Works}
-
-\subsection{Tabular Data}
-- most common type of data
-- bad performance of deep learning models on tabular data
 
@@
 - Bayesian optimisation
-  * short intro
-  * why is it unsuitable for our task? (makes no sense to have it here, unless we introduce our problem (feature selection is high-dim binary space) in the introduction)
-- evolutionary algorithm
+-> both unsuitable for our task
+  + would be ok for weight decay (L2) + dropout optim
+  + but: approaches for feature selection (L1) not feasible
+    (a) 0-1 penalty on included features added to loss, problems:
+      - how to choose regularisation parameter?
+      - not all features equally important -> weighted 0-1 loss? how to choose weights?
+    (b) directly optimise over binary hpyercube -> how to optim over non-ordinal discrete space?
+-> evolutionary algorithms can be a solution for these types of problems
   * CMA-ES
@@
 \subsubsection{Neural Architecture Search (NAS)}
+not sure if super relevant as I don't use a specific NAS technology (e.g. one-shot, etc.) but simply optimise over \# layers + \# nodes as HPs
 
 \subsection{Multi-Objective Optimization}
-- intro
+- problem: all aforementioned approaches are a-priori approaches -> we choose best HPs (weight decay, features used, dropout) w.r.t. performance
+- intro MOO
 

### 2025-03-06 code/classes.py
change featureless AUC to 0.5

@@ -566,9 +566,7 @@ class EAGGA:
         self.data_train_val = self.data_train_val.reset_index(drop=True)
         self.data_test = self.data_test.reset_index(drop=True)
 
-        majority_class = self.data_train_val.loc[:, self.class_column].mode().item()
-        majority_class_fraction = self.data_train_val.loc[self.data_train_val.loc[:, self.class_column] == majority_class, :].shape[0] / self.data_train_val.shape[0]
-        self.performance_majority_predictor = (majority_class_fraction, 0, 0, 0)
+        self.performance_majority_predictor = (0.5, 0, 0, 0)
         self.nds = NonDominatedSorting()
         self.hv_obj = HV(ref_point=(1, 1, 1, 1), nds=False)  # set ref-pt to (1, 1, 1, 1), as pymoo always minimises, i.e. ref-pt (0, 1, 1, 1) would consider 0 to be largest value in first dim + result in hypervolume = 0, instead transform first dim of points so that auc becomes minimisation problem (via 1 - auc)
 

### 2025-03-07 paper/main.tex
more bullets for paper

@@
   * another classical NN regularisation technique: dropout against co-adaptation
-  * early stopping
+  * early stopping against overfitting when using iterative method
 
@@
 - our approach
+  (general algorithm)
+  * EAGGA algorithm implemented just as described in the paper
+  * group structure init
+    - feature detector
+      * as described in paper
+      * but instead of fitting 10 trees + taking rel. \# of features used as p for trunc geom, we simply use 0.5 (sklearn dectree analysis a bit tricky)
+      * also in preliminary experiments we found the sampled \# of features to be used occassionally to be > \# of non-0 values in normalised information gain filter
+        (e.g. if former is = total \# of features and one feature is indep. from target) in these edge cases we only use the features with non-0 filter values
+    - interaction detector
+      * as described in paper
+      * same as feat detect, use p=0.5 to sample \# of interactions used instead of 10 dec trees
+      * also for FAST algorithm don't use RSS (lin reg metric) but mean accuracy, as we fit log reg
+    - monotonicity detector
+      * as described in paper
+      * use default HPs (max depth 30, minsplie = 20) of mlr3 classification tree implementation, as this is the library the paper uses
+    -> both interaction + monotonicity detectors fit their models on 80\% of train split from holdout + eval on remaining 20\%
+  (implementation details)
+  * group structures + datasets
+  * neural network
+    - "instead of XGB we use NN, ..., this requires special architecture, etc. pp."
+  * holdout, cv, early stopping
+
+  (old notes, use in new)
   * feature sparsity: straightforward, only train model on selected features, goes somewhat against idea of deep learning,

### 2025-03-08 paper/main.tex
some more bullets on my approach, restructuring of old bullets

--- 2025-03-05 paper/main.tex
@@
     - crowding distance
+    -> give intuition why non-dom-sort (asc) + crowding dist (desc) makes sense for selecting individuals
 \subsubsection{Neural Architecture Search (NAS)}
@@
 - our approach
-  * feature sparsity: straightforward, only train model on selected features, goes somewhat against idea of deep learning,
-    where the model is supposed to decide on its own, which features to put importance on -> WHY?
-  * feature interaction
-    - create non-fully connected network by introducing network-groups based on the equivalence relations introduced in the EAGGA paper
-    - show formula why the interacting features need to be grouped together when using ReLU (cf. photos)
-  * monotonicity constraints
-    - for groups with contraint on monotonicity: clip weights to [0, infty), bias clipping not necessary
-    - show formula of how monotonicity is enforced with this
-  * changes (-) + similarities (+) to EAGGA paper (~ in-between)
-    + 2/3 holdout outer split for 'CV'- vs test-set
-    - train + eval each individual via 5-fold CV with 20\% of train set again reserved to determine early stopping -> best to visualise,
-      * early stopping necessary (as opposed to XGB) as NN can be trained forever, whereas tree implementations used by EAGGA paper have
-        max depth, etc.
-      * early stopping using min-epochs 200, patience 100, i.e. train for $\ge$ 200 epochs, always safe model with lowest loss, whenever
-        current loss > mean(losses of last 100 (patience) epochs), stop early + return model with lowest loss
-    ~ if no early stopping, train each fold (in 5-fold CV) for max 120 secs, do entire EAGGA loop per dataset for max 8 hrs
-    ~ training on Sagemaker Notebook instance
-      * initial experiments on ml.g4dn.xlarge instance (2 vCPUs, 16GiB RAM, 1 NVIDIA T4, cf. https://aws.amazon.com/de/ec2/instance-types/g4/)
+  (general algorithm)
+  * EAGGA algorithm implemented just as described in the paper
+  * HP init
+    - NN total layers $\in\{3, ..., 10\}$, i.e. hidden layers $\in\{1, ..., 8\}$, init from trunc geom with p=0.5
+    - NN nodes per hidden layer (for each group) $\in\{3, ..., 20\}$, init from trunc geom with p=0.5
+    - NN dropout \% $\in[0, 1]$, init from trunc gamma with shape=2, scale=0.15
+  * group structure init
+    - feature detector
+      * as described in paper
+      * but instead of fitting 10 trees + taking rel. \# of features used as p for trunc geom, we simply use 0.5 (sklearn dectree examination not straightforward)
+      * also in preliminary experiments we found the sampled \# of features to be used occassionally to be > \# of non-0 values in normalised information gain filter
+        (e.g. if former is = total \# of features and one feature is indep. from target) in these edge cases we only use the features with non-0 filter values
+    - interaction detector
+      * as described in paper
+      * same as for feat detect: use p=0.5 to sample \# of interactions used instead of 10 dec trees
+      * also for FAST algorithm don't use RSS (lin reg metric) but mean accuracy, as we fit log reg
+    - monotonicity detector
+      * as described in paper
+      * use default HPs (max depth 30, minsplie = 20) of mlr3 classification tree implementation, as this is the library the paper uses
+    -> both interaction + monotonicity detectors fit their models on 80\% of train split from holdout + eval on remaining 20\%
+  (implementation details)
+  * group structures + datasets
+    - group structures implemented as described in the paper with additional list encoding sign of monotonicity of the individual features as detected by the
+      monotonicity detector
+    - included features are passed to dataset, dataset outputs included features, multiplied by the features' individual monotonicity (-1 / 1)
+    - entire group structure is passed to NN, where architecture is built accordingly
+  * neural network
+    - "instead of XGB as in the original paper, we apply the method to NNs to examine whether this type of regularisation can achieve interpretability on NNs while
+      outperforming EAGGA on XGB (original paper), ..., this requires special architecture, etc. pp."
+    - NN
+      * comprises of "sub-NNs", one for each equivalence relation -> basically non-fully connected NN
+      * hidden layers use ReLU activation and dropout afterwards, then a shared output layer with concatenated sub-NNs' outputs as input and sigmoid activation
+        (implicit in the loss function for better numerical stability, NN itself outputs "logits", which refers to pre-activation output in pytorch)
+    - feature sparsity thus achieved by only training on included feature groups -> goes somewhat against of deep learning where model is supposed to decide itself,
+      which feature to "use" / put importance on -> ELABORATE
+    - feature interaction achieved by grouping, max-operation in ReLU induces interaction effect (different kind of interaction than e.g. in LM with multiplication)
+      -> equation why the interacting features need to be grouped together when using ReLU, cf. photos
+    - monotonicty constraint achieved by clipping weights to [0, infty) for restricted equivalence relations (monotonic decrease achieved via dataset object multiplying
+      features with their individual signs), bias clipping not necessary (constant additive term)
+      -> equation how monotonicity is enforced with this
+  * evaluation, holdout, cv, early stopping
+    - evaluation via dominated hypervolume along AUC-ROC, NF, NI, NNM as defined in original paper
+      * NF simply rel. \# of included features
+      * NI = sum of all possible pairwise interactions in each group over all possible pairwise interactions among all features = $\frac{\sum_g^G {p_g \choose 2}}{{p \choose 2}}$
+      * NNM = rel. \# unconstrained features
+    - outer holdout split: 2/3 train, 1/3 test (as in paper)
+      * run EAGGA on holdout train split, i.e. train + select best $\mu=100$ individuals based on non-dom-sorting (ascending) + crowding distance (descending) as tie breaker
+    - inner CV split on outer train portion: 5-fold (as in paper)
+      * in each fold fit model on 80\% of CV-train portion, use remaining 20\% of CV-train for early stopping
+    - early stopping criterion
+      * always train for min 200 epochs, keep track of model with lowest loss
+      * after that, use patience of 100: if current model's loss is > mean of last 100 epochs' losses
+      * if no early stopping, train each model for max 10 minutes
+      * then stop training, return model with lowest loss
+    -> add graphic of entire dataset split
+    - then evaluate on last fold from CV + keep best $\mu$ individuals + generate $\lambda=10$ offspring for new generation + repeat
+  * hardware: training on Sagemaker Notebook instance
+    - initial experiments on ml.g4dn.xlarge instance (2 vCPUs, 16GiB RAM, 1 NVIDIA T4, cf. https://aws.amazon.com/de/ec2/instance-types/g4/)
     using cuda not much faster than on ml.t3.medium (2 Intel Xeon 8000 vCPUs, 4GiB RAM, cf. https://aws.amazon.com/de/ec2/instance-types/t3/)
-      * thus decided for more economic + ressourcenschonend t3.medium
-    - evaluation
-      * AUC + NF: straightforward
-      * NI: for each group with $\ge$ 2 features: sum from 1 to (\# features - 1), i.e ${\# features \choose 2}$,
-        then sum over all groups like this, then divide by p(p-1)/2, i.e. ${p \choose 2}$
-      * NNM: \# unconstrained features / \# all features
-    ~ feature + interaction detector: examination of decision trees not super straightforward in sklearn, hence simply use p=0.5
-    ? interaction detector (FAST)
-      * train logistic model on 80\% of data (only holdout train portion or all ??), evaluate on 20
-      * use mean accuracy (logistic metric) instead of RSS (LM metric)
-    + monotonicity detector as in paper, also use dec tree HPs from mlr3
-    ~ encoding of group structures similar, each feature will get its individual monotonicity sign from the detector, this is then fed to the dataset,
-      which then outputs the data according multiplied with the sign
+    - thus decided for more economic + ressourcen-schonend t3.medium
 

### 2025-03-08 paper/main.tex
add software references
rename some headlines, add known software bugs (+ their circumvention)

@@
 
-\ShortHeadings{Extending EAGGA to Neural Networks}{Simon Stürzebecher}
+\ShortHeadings{Interpretable Neural Networks using EAGGA}{Simon Stürzebecher}
 \firstpageno{1}
@@
 
-\title{Extending EAGGA to Neural Networks}
+\title{Interpretable Neural Networks using EAGGA}
 
@@
 % feels better in front of AutoML, as it feels more qualitative, less quantitative
-\subsection{Explainable Artificial Intelligence and Interpretability}
-\subsubsection{Why is interpretability desirable?}
-\subsubsection{Practical implementations}
+\subsection{Interpretability}
+- Why Interpretability is desirable?
+- Practical implementations
   (1) post-hoc methods
@@
 
-\subsection{Automated Machine Learning (AutoML)}
-\subsubsection{Hyperparameter Optimization (HPO)}
+\subsection{Hyperparameter Optimization (HPO)}
 - grid + random search
@@
     -> give intuition why non-dom-sort (asc) + crowding dist (desc) makes sense for selecting individuals
-\subsubsection{Neural Architecture Search (NAS)}
+\subsection{Neural Architecture Search (NAS)}
 not sure if super relevant as I don't use a specific NAS technology (e.g. one-shot, etc.) but simply optimise over \# layers + \# nodes as HPs
@@
 - intro MOO
+\subsubsection{Scalarisation}
+\subsubsection{Pareto-optimality}
 
@@
     - early stopping criterion
-      * always train for min 200 epochs, keep track of model with lowest loss
+      * for each fold, always train for min 200 epochs, keep track of model with lowest loss
       * after that, use patience of 100: if current model's loss is > mean of last 100 epochs' losses
-      * if no early stopping, train each model for max 10 minutes
+      * if no early stopping, train each fold for max 10 minutes
       * then stop training, return model with lowest loss
@@
     - thus decided for more economic + ressourcen-schonend t3.medium
+  * known bugs
+    - in rare cases (anecdotally every ~2 datasets), gga\_mutate seems to be generating -1 as monotonicity attribute, despite np.random.randint(low=0, high=2, size=1)
+      * loop crashes at group\_structure creation, for these cases subtract previous runtime from 8hrs + load from last generation in output via load\_population
+      * so far only happened for madeline in after gen-8.json after 5hrs 45mins (at start of gen-9, which wasn't exported yet) -> loaded this and ran for another 2hrs 15mins
 
@@
 \appendix
+\section{Software used}
+for implementation we used openml \cite{OpenML}, \cite{OpenMLPython}, numpy \cite{numpy}, pandas \cite{pandas1}, \cite{pandas2}, pytorch \textbf{TODO},
+scikit-learn \cite{scikit-learn}, scipy \cite{SciPy}, pymoo \cite{pymoo}, and tqdm \cite{tqdm}
+
 \section{}

### 2025-03-09 paper/main.tex
little bit of restructuring

@@
   * early stopping against overfitting when using iterative method
+-> all of those require specifying Hyperparameters in advance
 
@@
 - Bayesian optimisation
--> both unsuitable for our task
-  + would be ok for weight decay (L2) + dropout optim
-  + but: approaches for feature selection (L1) not feasible
-    (a) 0-1 penalty on included features added to loss, problems:
-      - how to choose regularisation parameter?
-      - not all features equally important -> weighted 0-1 loss? how to choose weights?
-    (b) directly optimise over binary hpyercube -> how to optim over non-ordinal discrete space?
--> evolutionary algorithms can be a solution for these types of problems
+- evolutionary algortihms
   * CMA-ES
-  * NSGA-II
-    - non-dominated sorting
-    - crowding distance
-    -> give intuition why non-dom-sort (asc) + crowding dist (desc) makes sense for selecting individuals
+  * differential evolution
+
 \subsection{Neural Architecture Search (NAS)}
-not sure if super relevant as I don't use a specific NAS technology (e.g. one-shot, etc.) but simply optimise over \# layers + \# nodes as HPs
+give background on NAS (e.g. one-shot, etc.), but for this project wanted to have a "minimal viable product" to test whether NNs can yield competitive performance
+on tabular ML + be interpretable using EAGGA at all before optimising the inner networks to the max
 
@@
 - intro MOO
-\subsubsection{Scalarisation}
 \subsubsection{Pareto-optimality}
+\subsubsection{A-priori}
+\subsubsection{A-posteriori}
+- NSGA-II (MOEA)
+  * non-dominated sorting
+  * crowding distance
+  -> give intuition why non-dom-sort (asc) + crowding dist (desc) makes sense for selecting individuals
 
@@
 \section{EAGGA}
+- for our problem (interpretability) we cannot use regular EA algorithms as we are also interested in reducing pairwise interactions + non-monotone feature effects
+- this creates high-dim search space as shown in EAGGA paper -> they provide a framework to efficiently traverse through a space that is reduced to only sensible HP
+  configuration (e.g. interaction of A,B with A monot incr and B monot decr -> couldn't be guaranteed -> don't look at it in first place)
 - NSGA-II inspired evolutionary / genetic algorithm
@@
 \section{Conclusion and Future Outlook}
-
-Here is a citation \cite{EAGGA}.
+- MO BO on group structure space possible?
+- here is a citation \cite{EAGGA}.
 
@@
 \section{Software used}
-for implementation we used openml \cite{OpenML}, \cite{OpenMLPython}, numpy \cite{numpy}, pandas \cite{pandas1}, \cite{pandas2}, pytorch \textbf{TODO},
+for implementation we used openml \cite{OpenML}, \cite{OpenMLPython}, numpy \cite{numpy}, pandas \cite{pandas1}, \cite{pandas2}, pytorch \cite{PyTorch},
 scikit-learn \cite{scikit-learn}, scipy \cite{SciPy}, pymoo \cite{pymoo}, and tqdm \cite{tqdm}

### 2025-03-09 paper/main.tex
started writing on interpretability

@@
 \subsection{Interpretability}
-- Why Interpretability is desirable?
-- Practical implementations
-  (1) post-hoc methods
-    - give examples
-    - why are they not perfectly suited
-  (2) model-based
+As there is no clear definition, we will consider ``interpretability'' as ``the ability to provide \textit{explanations} in \textit{understandable terms} to a human'',
+where \textit{explanations} are logical decision rules and \textit{understandable terms} relate to commonly used terms in the domain of the problem,
+as suggested by \citet[chap. 1]{survey_NN_interpretability}.
+We use the term ``explanability'' in an exchangeable manner with ``interpretability'', as is commonly done.
+
+(Why Interpretability is desirable?)
+- in many ways desirable, as per \citet[pp. 3-4]{Zach2019InterpretabilityOD}, e.g.
+  * \textbf{gaining trust}: most notable use case, e.g. medical diagnosis: predictions need to be validated by doctors \citep[p. 1]{review_NN_interpretability}
+  * \textbf{model debugging}: model is optimised w.r.t. loss and judged on accuracy, having an interpretable model + model induction process can help find ways to improve accuracy
+  * \textbf{scientific understanding}: when only models can make sense of increasingly complex data, interpretability enables extraction of learnt knowledge encoded in the model
+  * \textbf{subconcious bias}, e.g. loan approval: must ensure that decision is not discriminatory
+  * \textbf{regulatory}, such as EU's "Right to Explanation" warranted by the GDPR \citep[p. 1]{review_NN_interpretability}
+
+Commonly, methods for model interpretation are divided into intrinsic methods, where the search space only comprises models with a structure simple enough to be
+considered ``explainable'' (such as tree-based or simple linear models), and post-hoc methods, where interpretation methods are applied after model training.
+Amongst post-hoc techniques, we can further divide the space into model-specific (such as analysing GLM coefficients) and model-agnostic
+(e.g. partial depence plots, ALE) methods. \citet[chap. 3.2]{molnar2022}
+
+(Taxonomy)
+As standard neural networks (NNs) are inherently non-interpretable models, the taxonomy introduced by \citet[chap. 2]{survey_NN_interpretability} proves particulary useful
+to categorise NN explanability methods along 3 dimensions:
+- \textbf{passive vs active approaches}
+- \textbf{type of explanations}
+- \textbf{local vs global interpretability}
 
@@
 \section{Software used}
-for implementation we used openml \cite{OpenML}, \cite{OpenMLPython}, numpy \cite{numpy}, pandas \cite{pandas1}, \cite{pandas2}, pytorch \cite{PyTorch},
-scikit-learn \cite{scikit-learn}, scipy \cite{SciPy}, pymoo \cite{pymoo}, and tqdm \cite{tqdm}
+for implementation we used
+openml \cite{OpenML}, \cite{OpenMLPython},
+numpy \cite{numpy},
+pandas \cite{pandas1}, \cite{pandas2},
+pytorch \cite{PyTorch},
+scikit-learn \cite{scikit-learn},
+scipy \cite{SciPy},
+pymoo \cite{pymoo}, and
+tqdm \cite{tqdm}
 

### 2025-03-10 code/keywords.py
export plots as 300 dpi pngs

@@ -1,0 +1,3 @@
+oml_task_ids = [37, 43, 3903, 3904, 3913, 3918, 10093, 9946, 146819, 359955, 189922, 359962, 190392, 167120, 190137, 190410, 168350, 359975, 359972, 146820]
+# define positive classes
+positive_classes = ['tested_positive', '1', True, True, 'yes', True, '2', '2', '1', '2', '1', True, '1', '1', '2', '1', '2', 'Anomaly', '1', '2']

### 2025-03-10 paper/main.tex
more bullets on interpretability

--- 2025-03-09 paper/main.tex
@@
 \subsection{Interpretability}
-- Why Interpretability is desirable?
-- Practical implementations
-  (1) post-hoc methods
-    - give examples
-    - why are they not perfectly suited
-  (2) model-based
+As there is no clear definition for interpretability, we will consider it as ``the ability to provide \textit{explanations} in \textit{understandable terms} to a human'',
+where \textit{explanations} are logical decision rules and \textit{understandable terms} relate to commonly used terms in the domain of the problem,
+as suggested by \citet[chap. 1]{survey_NN_interpretability}.
+Further, we use the term ``explanability'' in an exchangeable manner with ``interpretability'', as is commonly done.
+
+(Why Interpretability is desirable?)
+- in many ways desirable, as per \citet[pp. 3-4]{Zach2019InterpretabilityOD}, e.g.
+  * \textbf{gaining trust}: most notable use case, e.g. medical diagnosis: predictions need to be validated by doctors \citep[p. 1]{review_NN_interpretability}
+  * \textbf{model debugging}
+    - model is optimised w.r.t. loss and judged on accuracy
+    - might have unexpected performance drops in certain situations, makes model unreliable \citep[1B]{survey_NN_interpretability}
+    - having an interpretable model + model induction process can help find ways to improve accuracy + reliability
+  * \textbf{scientific understanding}: when only models can make sense of increasingly complex data anymore, interpretability enables extraction of learnt knowledge encoded
+    in the model to make it accessible + reliable to humans (also mentioned here \citep[p. 1]{review_NN_interpretability})
+  * \textbf{subconcious bias}, e.g. loan approval: must ensure that decision is not discriminatory
+  * \textbf{regulatory}
+    - such as EU's "Right to Explanation" warranted by the GDPR \citep[p. 1]{review_NN_interpretability} or
+    - drug approval processes in situations where a machine learning model discovered a new drug, process needs to be transparent for the regulator to approve \citep[1B]{survey_NN_interpretability}
+
+Commonly, methods for model interpretation are divided into intrinsic methods, where the search space only comprises models with a structure simple enough to be
+considered ``explainable'' (such as tree-based or simple linear models), and post-hoc methods, where interpretation methods are applied after model training.
+Amongst post-hoc techniques, we can further divide the space into model-specific (such as analysing GLM coefficients) and model-agnostic
+(e.g. partial depence plots, ALE) methods. \citet[chap. 3.2]{molnar2022}
+
+(Taxonomy)
+\citet[chap. 2]{survey_NN_interpretability} extend this distinction to a 3-dimensional taxonomy, allowing for better categorisation of neural networks (NN), a model class
+that in its regular feedforwad form is inherently non-interpretable.
+% alternative
+%As regular feedforward neural networks (NNs) are inherently non-interpretable models, the taxonomy introduced by \citet[chap. 2]{survey_NN_interpretability} proves
+%particulary useful to categorise NN explanability methods along 3 dimensions:
+- \textbf{Passive vs Active Approaches}
+  * passive: post-hoc
+  * active: changing architecture or training process to make model interpretable
+- \textbf{Type of Explanations}
+  * examples: providing examples of what leads to desired output
+  * attribution: attributing effect on output for a specific feature
+  * hidden semantics: examine what type of inputs specific neurons / layers pick up on
+  * rules: logical rules, e.g. if-then, trees
+- \textbf{Local vs Global Interpretability}
+  * local: explanation based on individual samples
+  * semi-local: explanation based on set of samples, e.g. grouped together by some criterion
+  * global: explaining network as a whole
 

### 2025-03-10 paper/main.tex
more bullet on interpretability, mostly evaluation

@@
   * \textbf{regulatory}
-    - such as EU's "Right to Explanation" warranted by the GDPR \citep[p. 1]{review_NN_interpretability} or
+    - such as EU's "Right to Explanation" warranted by the GDPR \citep[p. 1]{review_NN_interpretability}, \citep[1B]{survey_NN_interpretability} or
     - drug approval processes in situations where a machine learning model discovered a new drug, process needs to be transparent for the regulator to approve \citep[1B]{survey_NN_interpretability}
@@
   * global: explaining network as a whole
+
+(Evaluation)
+- objective evaluation difficult, as no clear definition of interpretability
+- \citet[3]{DoshiVelez2017TowardsAR} propose a taxonomy to categorise possible evaluation methods
+  * \textbf{application-grounded evaluation}
+    - most rigorous method, also most expensive + time consuming
+    - idea: evaluate ``interpretability model'' directly w.r.t. the task -> human experts evaluate the outcome
+    - e.g.: model performing medical diagnosis on patients would be evaluated by based on doctors doing the same
+  * \textbf{human-grounded metrics}
+    - similar to application-grounded, but tries to simplify task (while preserving its essence) so that laypersons can do it
+    - cheaper due to larger, less qualified subject pool
+    - especially suitable if only general concepts of the tasks need to be validated
+    - examples for evaluation set-up
+      * binary forced choice: human evaluator chooses which of two explanations he finds better
+      * forward simulation: evaluator must correctly simulate model output when presented model input + explanation % perhaps binary forced choice is already enough to include, this (and the third one is a bit awkward to explain why this is good for interpretability imo)
+  * \textbf{functionally-grounded evaluation}
+    - assess explanation quality via some formally defined proxy for interpretability
+    - no human time nor cost required beyond initial formulation of proxy
+    - e.g. if we already have an interpretable model class (e.g. identified via human-grounded evaluation), can then rank different models of that class based on proxy
+    - main challenge: what proxies
 

### 2025-03-12 paper/main.tex
bullets on model-free HPO

@@
 \usepackage{jmlr2e}
+
+\usepackage{mathtools}
+\usepackage{bm}
 
@@
 \citet[chap. 2]{survey_NN_interpretability} extend this distinction to a 3-dimensional taxonomy, allowing for better categorisation of neural networks (NN), a model class
-that in its regular feedforwad form is inherently non-interpretable.
+that in its fully-connected feedforwad form is inherently non-interpretable.
 % alternative
-%As regular feedforward neural networks (NNs) are inherently non-interpretable models, the taxonomy introduced by \citet[chap. 2]{survey_NN_interpretability} proves
+%As fully-connected feedforward neural networks (NNs) are inherently non-interpretable models, the taxonomy introduced by \citet[chap. 2]{survey_NN_interpretability} proves
 %particulary useful to categorise NN explanability methods along 3 dimensions:
@@
 \subsection{Hyperparameter Optimization (HPO)}
-- grid + random search
-- Bayesian optimisation
-- evolutionary algortihms
+- give intro to HPO = black-box optim problem -> we can apply any black box algo
+
+[model free]
+(grid + random search)
+- most basic strategy: grid search, for all HPs define range of interest, then evaluate cartesian product of those
+- random search
+  * randomly sample value for each HP until it runs out of budget
+  * explorative: for budget $B$, each HP will (almost guaranteed) be queried with $B$ different values, whereas grid search only $B^{1/N}$ for $N$ HPs
+  * thus better than grid in case of some HPs being non-informative (almost always the case), as it doesn't waste time specifically exploring these (as does grid)
+  * usually useful as baseline, as no assumptions about model, easy to parallelise, high exploration, and in expectation will ``achieve performance arbitrarily close to the optimum''
+
+(evolutionary algorithms)
+- another model-free class of optimisation algorithms, inspired by naturally occuring evolution
+- based on a population, which in each generation (iteration) generates $\lambda$ offspring via 3 operators (all sub-bullets from \citet[pp. 10-]{genetic_algos})
+  * reproduction: pick an individual to reproduce with p proportional to its fitness
+  * crossover
+    - select 2 ``parents'' from pool of reproducing individuals
+    - for a position $k$ in their HP configs, with certain p, swap their values after $k$, yielding 2 children
+    %use example from book page 12
+  * mutation: with certain p, change values of an individual, e.g. by adding Gaussian noise (Gaussian mutation) to real values, or flipping bit for binary values
+- selection at end of each genertion: keep $\mu$ best individuals (based on some fitness function), either from only offspring (``$(\mu,\lambda)$-selection'')
+  or (more commonly) from population + offspring (``$(\mu+\lambda)$-selection'', guaranteed to keep best individual)
+- pro: conceptually simple + can handle even complex parameter spaces (continuous, discrete, hierarchical, etc.) given appropriate implementation of operators
+all above starting at HPO from \citep[chap. 1.3]{feurer_hyperparameter_2019}
+- prominent examples
   * CMA-ES
-  * differential evolution
+    - ``Covariance Matrix Adapation Evolution Strategy''
+    - offspring generation exclusively via multivariate normal \citep[p. 8]{hansen2023cmaevolutionstrategytutorial}
+      * mean = weighted average of previous generation
+      * covariance = weighted covariance of previous generation, with weighting as for mean
+    - weighing scheme done to sample in a way as to reproduce previously successful (i.e. selected) steps \citep[p. 11]{hansen2023cmaevolutionstrategytutorial}
+  * differential evolution \citep[-]{differential_evolution}
+    - init population randomly so that entire param space is covered
+      * $D$-dim vectors $\boldsymbol{x}_{i,G}$ with $i=1,2,...,n$ for gen $G$
+    - mutation
+      * for each $\boldsymbol{x}_{i,G}$, generate \textbf{mutant vector} $\boldsymbol{v}_{i,G+1}=\boldsymbol{x}_{r_1,G}+F\cdot(\boldsymbol{x}_{r_2,G}-\boldsymbol{x}_{r_3,G})$
+      * $r_1,r_2,r_3\in\{1,2,...,n\}$ mutually different random idx + different from $i$
+      * constant $F\in[0,2]$
+    - crossover
+      * \textbf{trial vector} $\boldsymbol{u}_{i,G+1}=(u_{i,G+1}^{(1)},u_{i,G+1}^{(2)},...,u_{i,G+1}^{(D)})$
+        - chose random index $R\in\{1,2,...,D\}$
+        - sample $p\sim U[0,1]$
+        - define crossover constant $\text{CR}\in[0,1]$
+        - $u_{i,G+1} = \begin{dcases}
+          v_{i,G+1}^{(j)} & \text{if } p\le\text{CR}\text{ or } j=R \\
+          x_{i,G}^{(j)} & \text{else}
+         \end{dcases}$
+    - selection: greedy, compare fitness of $\boldsymbol{u}_{i,G+1}$ with $\boldsymbol{x}_{i,G}$, pick better
+
+[model based]
+(Bayesian optimisation)
 
@@
   * known bugs
-    - in rare cases (anecdotally every ~2 datasets), gga\_mutate seems to be generating -1 as monotonicity attribute, despite np.random.randint(low=0, high=2, size=1)
+    - in rare cases (anecdotally once every ~5-10 datasets), gga\_mutate seems to be generating -1 as monotonicity attribute, despite np.random.randint(low=0, high=2, size=1)
       * loop crashes at group\_structure creation, for these cases subtract previous runtime from 8hrs + load from last generation in output via load\_population

### 2025-03-12 paper/main.tex
bullets for model-based HPO

@@
 (grid + random search)
-- most basic strategy: grid search, for all HPs define range of interest, then evaluate cartesian product of those
+- most basic strategy: grid search, for all HPs define range of interest, then evaluate cartesian product of those -> scales poorly in \# of HP dimensions + \# of query points per HP range
 - random search
   * randomly sample value for each HP until it runs out of budget
-  * explorative: for budget $B$, each HP will (almost guaranteed) be queried with $B$ different values, whereas grid search only $B^{1/N}$ for $N$ HPs
+  * explorative: for budget $B$, each HP will (likely) be queried with $B$ different values, whereas grid search only $B^{1/N}$ for $N$ HPs
   * thus better than grid in case of some HPs being non-informative (almost always the case), as it doesn't waste time specifically exploring these (as does grid)
@@
 - another model-free class of optimisation algorithms, inspired by naturally occuring evolution
-- based on a population, which in each generation (iteration) generates $\lambda$ offspring via 3 operators (all sub-bullets from \citet[pp. 10-]{genetic_algos})
+- based on a population, which iteratively (iteration = generation) generates $\lambda$ offspring via 3 operators (all sub-bullets from \citet[pp. 10-]{genetic_algos})
   * reproduction: pick an individual to reproduce with p proportional to its fitness
@@
 [model based]
+- contrast to model free, fit a surrogate model on blackbox problem + optimise this
 (Bayesian optimisation)
+- iterative algo comprising of
+  * surrogate model for black box problem, needs to be able to model mean + variance
+  * acquisition function, to decide which point to query next
+- in each iteration
+  * fit surrogate model on all data points (posterior distribution)
+  * get highest utility point from acquisition function for newly fitted surrogate model -> add to data points
+  -> hence Bayesian, get new query point given already previously fitted points
+  * acquisition function trades off exploration + exploitation of surrogate
+  * BO up to here from \citep[chap. 1.3.2]{feurer_hyperparameter_2019} + \citep[pp. 2-3]{frazier2018tutorialbayesianoptimization}
+- no optimising of model directly, instead iteratively optimise acquisition function
+- common choices for surrogate models
+  * GPs
+    - pros
+      * fully specified by mean + covariance
+        - covariance function (aka kernel) solely determines quality of GP
+        - kernel as function of two points from search space, yields their covariance
+        - usual property of kernels: the closer points are in search space, the more strongly their correlation \citep[p. 5]{frazier2018tutorialbayesianoptimization}
+      * well-calibrated uncertainty estimates
+      * closed-form computability
+    - con: neither scales well in \# of data points nor in \# of HP dimensions BUT workaround: sparse GPs, approx full GP with small subset of original dataset
+  * random forests
+    - can handle complex search spaces (high-dim, categorical, hierarchical), unlike GPs
+    - computational complexity scales far better than GPs
+      * GPs $O(n^3)$ fitting, $O(n^2)$ predicting
+      * RFs $O(n \log n)$ fitting, $O(\log n)$ predicting
+- BO stuff from \citep[chap. 1.3.2]{feurer_hyperparameter_2019}
 

### 2025-03-12 paper/main.tex
bullets on NAS

@@
 - based on a population, which iteratively (iteration = generation) generates $\lambda$ offspring via 3 operators (all sub-bullets from \citet[pp. 10-]{genetic_algos})
-  * reproduction: pick an individual to reproduce with p proportional to its fitness
-  * crossover
+  * \textbf{reproduction}: pick an individual to reproduce with p proportional to its fitness
+  * \textbf{crossover}
     - select 2 ``parents'' from pool of reproducing individuals
@@
     %use example from book page 12
-  * mutation: with certain p, change values of an individual, e.g. by adding Gaussian noise (Gaussian mutation) to real values, or flipping bit for binary values
+  * \textbf{mutation}: with certain p, change values of an individual, e.g. by adding Gaussian noise (Gaussian mutation) to real values, or flipping bit for binary values
 - selection at end of each genertion: keep $\mu$ best individuals (based on some fitness function), either from only offspring (``$(\mu,\lambda)$-selection'')
@@
 - common choices for surrogate models
-  * GPs
+  * \textbf{gaussian processes}
     - pros
@@
         - kernel as function of two points from search space, yields their covariance
+        % TODO: write down kernel equation, cf. p 10ff https://ieeexplore.ieee.org/document/7352306
         - usual property of kernels: the closer points are in search space, the more strongly their correlation \citep[p. 5]{frazier2018tutorialbayesianoptimization}
@@
     - con: neither scales well in \# of data points nor in \# of HP dimensions BUT workaround: sparse GPs, approx full GP with small subset of original dataset
-  * random forests
+  * \textbf{random forests}
     - can handle complex search spaces (high-dim, categorical, hierarchical), unlike GPs
@@
       * RFs $O(n \log n)$ fitting, $O(\log n)$ predicting
-- BO stuff from \citep[chap. 1.3.2]{feurer_hyperparameter_2019}
+- common choices for acquisition functions
+  % TODO: cf. p. 13 cf. p 10ff https://ieeexplore.ieee.org/document/7352306
+  * \textbf{expected improvement}
+  * \textbf{Thompson sampling}
+- all BO stuff if not specified differently comes from \citep[chap. 1.3.2]{feurer_hyperparameter_2019}
 
 \subsection{Neural Architecture Search (NAS)}
-give background on NAS (e.g. one-shot, etc.), but for this project wanted to have a "minimal viable product" to test whether NNs can yield competitive performance
-on tabular ML + be interpretable using EAGGA at all before optimising the inner networks to the max
+- approaches specifically for NN HPO due to flexible structure of NNs and thus very large search space (each layer can have different \# nodes, different activations, for CNN different pooling, etc.
+operations)
+- we don't employ NAS for our extension as research focusses on the NLP and image domains, % TODO: find some more references
+  whose datasets exhibit strong correlation amongst features (i.e. tokens or pixels, respectively), an effect that is much weaker for tabular data \citep[p. 1]{Borisov_2024}
+- at least want to give a little overview over most notable approaches
+  * \textbf{cell search space}
+    - modularise NN architecture into cells, NN as chain-structure of those
+    - cell = basic building block, fixed component of an NN
+      * e.g. linear layer with specific \# of neurons in feedforward
+      * or convolutional / pooling / etc layer for CNN
+    - then optimise sequential placement of the cells as HPO problem
+      * e.g. via random search or BO
+      * \citet[p. 3]{zoph2017neuralarchitecturesearchreinforcement} + \citet[pp. 2-4]{Zoph_2018_CVPR} use RNN to recursively optimise HPs of a cell given already determined
+        previous cells (+ HPs of current cell), which is trained via RL with the different HPs the RNN predicts being the actions and performance on held-out data being reward
+    - \citep[chap. 3.2]{elsken_neural_2019}
+  * \textbf{one-shot model}
+    - trains one overall ``fabric'' comprising all architectures of search space
+    - visualisation as DAG, nodes are nodes, edges in-between are operations on a node
+    - each path through DAG (from input to output node) represents one architecture of search space -> train entire DAG, then pick optimal path (architecture)
+    - pro: very efficient training, individual architectures share operations along edges they share, training one-shot model trains all subsumed models (more expensive
+      than training a single model but less expensive than trying out all configurations included in the fabric)
+    - \citep[pp. 1-2, p.8]{saxena2017convolutionalneuralfabrics}
+    % don't necessarily talk about co-adaptation problem
 

### 2025-03-12 paper/main.tex
first bullets on MOO a posteriori

@@
 \begin{keywords}
-  tabular data, multi-objective optimization, deep learning
+  tabular data, multi-objective optimization, interpretability, deep learning
 \end{keywords}
@@
 \subsection{Multi-Objective Optimization}
-- problem: all aforementioned approaches are a-priori approaches -> we choose best HPs (weight decay, features used, dropout) w.r.t. performance
 - intro MOO
@@
 \subsubsection{A-posteriori}
+- problem a priori
+  * either restrict search space to ``enforce our will'' (e.g. only use 50\% of features), optimise only prediction performance + use this
+  * or leave search space unrestricted but adjust loss function to incorporate multiple objectives + take optimum from there
+  * in either case: no knowledge of interplay between HP config + performance on all objectives
+  * a-posteriori evaluates configs just as a-priori, but keeps track of multiple ``best'' (non-dominated) solutions -> makes relationship between HP config + objectives visible
 - NSGA-II (MOEA)
@@
 \section{EAGGA}
+- combines all previously mentioned topics: interpretability, HPO, MOO
 - for our problem (interpretability) we cannot use regular EA algorithms as we are also interested in reducing pairwise interactions + non-monotone feature effects

### 2025-03-13 paper/main.tex
add HPO intro comments + formal definition of problem, start MOO bullets

@@
 \usepackage{mathtools}
+\usepackage{amsmath}  % use argmin + argmax, cf. https://tex.stackexchange.com/a/5255
+\DeclareMathOperator*{\argmax}{arg\,max}
+\DeclareMathOperator*{\argmin}{arg\,min}
 \usepackage{bm}
@@
 \subsection{Hyperparameter Optimization (HPO)}
-- give intro to HPO = black-box optim problem -> we can apply any black box algo
+- in contrast to model parameters, which are optimised during training, hyperparameters are parameters describing the ML algo + are specified before training
+- still often have great impact on final (trained) model performance -> can be optimised too
+- formal definition
+  * dataset $\mathcal{D}\subseteq\mathcal{X}\times\mathcal{Y}$ consisting of $n$ tuples $(\boldsymbol{x}^{(i)}, y^{(i)})\stackrel{i.i.d.}{\sim}\mathbb{P}_{\boldsymbol{x}y}$
+    (data-generating distribution), i.e. $\mathcal{D}=\{(\boldsymbol{x}^{(i)}, y^{(i)})\}_{i=1}^n$
+  * ML algo $\mathcal{I}(\cdot,\boldsymbol\lambda)$ with HP config $\boldsymbol\lambda\in\Lambda$ maps dataset to fitted model
+    $f:\mathcal{X}\rightarrow\mathbb{R}^g\subseteq\mathcal{Y}$ (where $g\in\mathbb{N}$ is \# of classes of target),
+    i.e. $\mathcal{I}:(\mathbb{D}\times\Lambda)\rightarrow\mathcal{H}$, with $\mathbb{D}$ space of all finite datasets and $\mathcal{H}$ hypothesis space (space of all models $f$)
+  * denote $\mathcal{I}_{\boldsymbol\lambda}$ as (untrained) model with fixed HP config $\boldsymbol\lambda$
+  * loss function $L$
+  * goal of model training: for fixed $\boldsymbol\lambda$, have $\mathcal{I}$ find model $f$ minimising expected generalisation error
+    $GE(\mathcal{I},\boldsymbol\lambda,\mathcal{D},L)=\mathbb{E}_{(\boldsymbol{x},y)\sim\mathbb{P}_{\boldsymbol{x}y}}[L(y,\mathcal{I}_{\boldsymbol\lambda}(\mathcal{D})(\boldsymbol{x}))]$
+  * goal of HP optim: find $\boldsymbol\lambda$ minimising expected generalisation error, i.e. $\argmin_{\boldsymbol\lambda\in\Lambda} GE(\mathcal{I},\boldsymbol\lambda,\mathcal{D},L)$
+- \citep[p. 3]{10.1145/3610536}
+- HPO = black-box optim problem -> we can apply any black box algo to tune
 
@@
 \subsection{Multi-Objective Optimization}
-- intro MOO
+- in most practical use cases one doesn't just want to optimise for performance but e.g. also for interpretability using some proxy metrics
+- formal definition
 \subsubsection{Pareto-optimality}
 \subsubsection{A-priori}
+- requires specifying trade-off between objectives a-priori
+- simplest approach: different versions of \textbf{scalarization}
+  * weighted sum of objective functions
+    - $\min_{x\in\mathcal{X}} w_i f_i(x)$ with $\sum_{i=1}^k w_i=1$ and $w_i>0,\forall i=1,...,k$
+    - \citep[p. 11]{10.1145/3610536}
+  * Chebyshev: provably converges to Pareto front
 \subsubsection{A-posteriori}

### 2025-03-13 paper/main.tex
add bullets Pareto optimality

@@
 - formal definition
+  * vector of objectives $c_1,c_2,...,c_m=\boldsymbol{c}:\Lambda\rightarrow\mathbb{R}^m$
+  * goal: minimise vector
+  * \citep[p. 11]{10.1145/3610536}
 \subsubsection{Pareto-optimality}
+- problem: usually conflicting objectives, i.e. minimisation of $\boldsymbol{c}$ along all dimensions not possible
+- thus aim to find trade-off solutions of non-dominated points
+- we say a point dominates another
+  * if there is no other point strictly better in at least one dimension and better or equal in the remaining ones
+  * formally: $\boldsymbol\lambda$ dominates $\boldsymbol\lambda'$ ($\boldsymbol\lambda\prec\boldsymbol\lambda'$) if and only if
+    $\forall i\in\{1,...,m\}:c_i(\boldsymbol\lambda) \le c_i(\boldsymbol\lambda') \wedge \exists j\in\{1,...,m\}:c_j(\boldsymbol\lambda) < c_j(\boldsymbol\lambda')$
+  * \citep[pp. 7f]{10.1145/3610536} + \citep[pp. 198f]{genetic_algos}
+- Pareto set: set of nondominated points $\mathcal{P}:=\{\boldsymbol\lambda\in\Lambda|\nexists\boldsymbol\lambda'\in\Lambda\text{ s.t. }\boldsymbol\lambda'\prec\boldsymbol\lambda\}$
+- Pareto front: image of nondominated points
+- goal: find set of nondominated points $\hat{\mathcal{P}}$ approximating true Pareto set $\mathcal{P}$ well
 \subsubsection{A-priori}

### 2025-03-13 paper/main.tex
bullets on a-priori MOO methods

@@
 \subsubsection{Pareto-optimality}
-- problem: usually conflicting objectives, i.e. minimisation of $\boldsymbol{c}$ along all dimensions not possible
+- problem: usually conflicting objectives, i.e. minimisation of $\boldsymbol{c}$ not possible along all dimensions
 - thus aim to find trade-off solutions of non-dominated points
@@
 \subsubsection{A-priori}
-- requires specifying trade-off between objectives a-priori
-- simplest approach: different versions of \textbf{scalarization}
+- requires specifying trade-off between objectives a-priori, will outline 2 popular approaches
+- e.g. different versions of \textbf{scalarization}
   * weighted sum of objective functions
-    - $\min_{x\in\mathcal{X}} w_i f_i(x)$ with $\sum_{i=1}^k w_i=1$ and $w_i>0,\forall i=1,...,k$
+    - $\argmin_{\boldsymbol\lambda\in\Lambda} w_i c_i(\boldsymbol\lambda)$ with $\sum_{i=1}^k w_i=1$ and $w_i>0,\forall i=1,...,k$
+    - drawbacks
+      * solution sensitive to weights
+      * different users might have different opinions on weights \citep[chap. 3.1]{NSGA}
     - \citep[p. 11]{10.1145/3610536}
-  * Chebyshev: provably converges to Pareto front
+  * $\epsilon$-constraint
+    - translate all but one objective into constraint, then optimise remaining objective under subject to the constraints
+    - $\text{w.l.o.g.} \argmin_{\boldsymbol\lambda\in\Lambda} c_1(\boldsymbol\lambda) \text{ s.t. } c_2(\boldsymbol\lambda)\le\epsilon_2,...,c_m(\boldsymbol\lambda)\le\epsilon_m$
+    - conceputally similar to weighted sum: also sensitive to constraints, must be chosen sensibly
+    - \citep[p. 12]{10.1145/3610536}
+- \textbf{lexicographic method}
+  * define priorisation of objectives
+  * greedily optimise objectives in order of priority, constraint to the solutions of the already optimised (higher-priority) objectives
+  * again very dependent on user-defined priorisation
+  * \citep[p. 13749]{lexicographic_MOO}  % TODO: perhaps find a different source more related to ML
 \subsubsection{A-posteriori}
@@
   * in either case: no knowledge of interplay between HP config + performance on all objectives
+  * in practical applications, it is oftentimes useful to make the decision of which point of the Pareto set to use a-posteriori, e.g. if only a slight decrease in
+    one objective translates to a significant improvement in another that would have been missed if the problem was optimised using a-priori methods
   * a-posteriori evaluates configs just as a-priori, but keeps track of multiple ``best'' (non-dominated) solutions -> makes relationship between HP config + objectives visible

### 2025-03-13 paper/main.tex
finish MOO bullets

--- 2025-03-13 paper/main.tex
@@
 - formal definition
+  * vector of objectives $c_1,c_2,...,c_m=\boldsymbol{c}:\Lambda\rightarrow\mathbb{R}^m$
+  * goal: minimise vector
+  * \citep[p. 11]{10.1145/3610536}
 \subsubsection{Pareto-optimality}
+- problem: usually conflicting objectives, i.e. minimisation of $\boldsymbol{c}$ not possible along all dimensions
+- thus aim to find trade-off solutions of non-dominated points
+- we say a point dominates another
+  * if there is no other point strictly better in at least one dimension and better or equal in the remaining ones
+  * formally: $\boldsymbol\lambda$ dominates $\boldsymbol\lambda'$ ($\boldsymbol\lambda\prec\boldsymbol\lambda'$) if and only if
+    $\forall i\in\{1,...,m\}:c_i(\boldsymbol\lambda) \le c_i(\boldsymbol\lambda') \wedge \exists j\in\{1,...,m\}:c_j(\boldsymbol\lambda) < c_j(\boldsymbol\lambda')$
+  * \citep[pp. 7f]{10.1145/3610536} + \citep[pp. 198f]{genetic_algos}
+- Pareto set: set of nondominated points $\mathcal{P}:=\{\boldsymbol\lambda\in\Lambda|\nexists\boldsymbol\lambda'\in\Lambda\text{ s.t. }\boldsymbol\lambda'\prec\boldsymbol\lambda\}$
+- Pareto front: image of nondominated points
+- goal: find set of nondominated points $\hat{\mathcal{P}}$ approximating true Pareto set $\mathcal{P}$ well
 \subsubsection{A-priori}
-- requires specifying trade-off between objectives a-priori
-- simplest approach: different versions of \textbf{scalarization}
+- requires specifying trade-off between objectives a-priori, will outline 2 popular approaches
+- e.g. different versions of \textbf{scalarization}
   * weighted sum of objective functions
-    - $\min_{x\in\mathcal{X}} w_i f_i(x)$ with $\sum_{i=1}^k w_i=1$ and $w_i>0,\forall i=1,...,k$
+    - $\argmin_{\boldsymbol\lambda\in\Lambda} w_i c_i(\boldsymbol\lambda)$ with $\sum_{i=1}^k w_i=1$ and $w_i>0,\forall i=1,...,k$
+    - drawbacks
+      * solution sensitive to weights
+      * different users might have different opinions on weights \citep[chap. 3.1]{NSGA}
     - \citep[p. 11]{10.1145/3610536}
-  * Chebyshev: provably converges to Pareto front
+  * $\epsilon$-constraint
+    - translate all but one objective into constraint, then optimise remaining objective under subject to the constraints
+    - $\text{w.l.o.g.} \argmin_{\boldsymbol\lambda\in\Lambda} c_1(\boldsymbol\lambda) \text{ s.t. } c_2(\boldsymbol\lambda)\le\epsilon_2,...,c_m(\boldsymbol\lambda)\le\epsilon_m$
+    - conceputally similar to weighted sum: also sensitive to constraints, must be chosen sensibly
+    - \citep[p. 12]{10.1145/3610536}
+- \textbf{lexicographic method}
+  * define priorisation of objectives
+  * greedily optimise objectives in order of priority, constraint to the solutions of the already optimised (higher-priority) objectives
+  * again very dependent on user-defined priorisation
+  * \citep[p. 13749]{lexicographic_MOO}  % TODO: perhaps find a different source more related to ML
 \subsubsection{A-posteriori}
@@
   * in either case: no knowledge of interplay between HP config + performance on all objectives
+  * in practical applications, it is oftentimes useful to make the decision of which point of the Pareto set to use a-posteriori, e.g. if only a slight decrease in
+    one objective translates to a significant improvement in another that would have been missed if the problem was optimised using a-priori methods
   * a-posteriori evaluates configs just as a-priori, but keeps track of multiple ``best'' (non-dominated) solutions -> makes relationship between HP config + objectives visible
-- NSGA-II (MOEA)
-  * non-dominated sorting
-  * crowding distance
-  -> give intuition why non-dom-sort (asc) + crowding dist (desc) makes sense for selecting individuals
+- aside from usualy baselines grid and random seach there are multi-objective BO adaptations, mainly using either of two approaches
+  (1) fit single surrogate model on scalarised objectives, e.g. ParEGO, which employs the augemented Tchebycheff function as scalarisation to ensure the Pareto front is
+      explored sufficiently \citep[pp. 15f]{10.1145/3610536} + \citep[pp. 54-56]{ParEGO}
+  (2) fit one surrogate per objective, then
+    - either one acquisition function per objective -> return set of promising next candidates
+    - or one overall acquisition function aggregating surrogates, e.g. EHI, maximises expected improvement of hypervolume \citep[p. 16]{10.1145/3610536} + \citep[pp. 8f]{EHI}
+- lastly, there is also multi-objective EA (MOEA) algorithms to explore Pareto front, one of which is NSGA-II (Nondominated Sorting Genetic Algorithm)
+  * improves upon predecessor NSGA \citep{NSGA} by making it parameterless, ensuring elitism, and reducing computational complexity of ranking of individuals \citep[p. 182]{NSGA_II}
+  * uses all the regular operators, i.e. reproduction, mutation, crossover can be used from single-objective EA (SOEA)
+  * difference to SOEA: ranking of individuals (SOEA uses scalar fitness, MOEA mutliple objectives)
+    - ranking mechanism based on 2 parts
+      * non-dominated sorting, ensures elitism
+        - iterative procedure to rank individuals in population by their fronts
+        (1) determine pareto front, assign rank 0
+        (2) remove pareto front from population
+        (3) if individuals left, repeat from 1, increment rank \#
+        - \citep[p. 201]{genetic_algos} + \citep[pp. 183f]{NSGA_II}
+      * crowding distance, ensures sufficient diversity, i.e. exploration of pareto front
+        - assigns score depending on how crowded area around individual in objective space is
+        - crowding distance of an individual = mean side length of cuboid spanned by its nearest neighbours as vertices in objective space
+        - individuals without two neighbour in any dimension are assigned infinitely high distance value
+        - \citep[p. 185]{NSGA_II}
+        -> the less the crowding distance, the more ``crowded'' an area is by other individuals
+      -> rank individuals by nds front rank (ascending) + crowding distance (descending) as tie breaker
+  * ranking used for selecting $\mu$ best individuals to keep for next generation + for reproduction: via binary tournament solution: select two random individuals,
+    pick best w.r.t. nds + cd for offspring creation (mutation / crossover)
+  * using nds for ranking / fitness makes intuitive sense, but why cd?
+    - goal of MOO: not simply optimise HV, but approx true Pareto front well
+    - only having individuals representing one area in objective space makes Pareto front estimate very ``unstable'': removing this area (i.e. solutions form that area)
+      would ``collapse'' entire front
+    - instead having multiple areas be represented makes the estimate ``stable'': removing one area wouldn't impact Pareto front estimate a lot
+    % TODO: reference?
 

### 2025-03-13 paper/main.tex
add bullets on evaluation of Pareto front estimates + reference for crowding distance intuition

@@
 - goal: find set of nondominated points $\hat{\mathcal{P}}$ approximating true Pareto set $\mathcal{P}$ well
+(evaluation)
+- if knowledge over true Pareto-front, evaluating set of individuals can be done based on distance of estimated to true Pareto front
+- if no knowledge over true Pareto-front, volume-based approaches are popular, which measure volume between Pareto front estim and some chosen
+  reference point (usually worst point in objective space)
+- \citep[pp. 8-10]{10.1145/3610536}
 \subsubsection{A-priori}
@@
     - instead having multiple areas be represented makes the estimate ``stable'': removing one area wouldn't impact Pareto front estimate a lot
-    % TODO: reference?
+    - \citep[p. 185, pp. 189-192]{genetic_algos}
 

### 2025-03-13 paper/main.tex
EAGGA bullets

@@
     (data-generating distribution), i.e. $\mathcal{D}=\{(\boldsymbol{x}^{(i)}, y^{(i)})\}_{i=1}^n$
-  * ML algo $\mathcal{I}(\cdot,\boldsymbol\lambda)$ with HP config $\boldsymbol\lambda\in\Lambda$ maps dataset to fitted model
-    $f:\mathcal{X}\rightarrow\mathbb{R}^g\subseteq\mathcal{Y}$ (where $g\in\mathbb{N}$ is \# of classes of target),
-    i.e. $\mathcal{I}:(\mathbb{D}\times\Lambda)\rightarrow\mathcal{H}$, with $\mathbb{D}$ space of all finite datasets and $\mathcal{H}$ hypothesis space (space of all models $f$)
+  * ML algo $\mathcal{I}(\cdot,\boldsymbol\lambda)$ with HP config $\boldsymbol\lambda\in\boldsymbol\Lambda$ maps dataset to model
+    $\hat{f}:\mathcal{X}\rightarrow\mathbb{R}^g\subseteq\mathcal{Y}$ (where $g\in\mathbb{N}$ is \# of classes of target),
+    i.e. $\mathcal{I}:(\mathbb{D}\times\boldsymbol\Lambda)\rightarrow\mathcal{H}$, with $\mathbb{D}$ space of all finite datasets and $\mathcal{H}$ hypothesis space (space of all models $\hat{f}$)
   * denote $\mathcal{I}_{\boldsymbol\lambda}$ as (untrained) model with fixed HP config $\boldsymbol\lambda$
   * loss function $L$
-  * goal of model training: for fixed $\boldsymbol\lambda$, have $\mathcal{I}$ find model $f$ minimising expected generalisation error
+  * goal of model training: for fixed $\boldsymbol\lambda$, have $\mathcal{I}$ find model $\hat{f}$ minimising expected generalisation error
     $GE(\mathcal{I},\boldsymbol\lambda,\mathcal{D},L)=\mathbb{E}_{(\boldsymbol{x},y)\sim\mathbb{P}_{\boldsymbol{x}y}}[L(y,\mathcal{I}_{\boldsymbol\lambda}(\mathcal{D})(\boldsymbol{x}))]$
-  * goal of HP optim: find $\boldsymbol\lambda$ minimising expected generalisation error, i.e. $\argmin_{\boldsymbol\lambda\in\Lambda} GE(\mathcal{I},\boldsymbol\lambda,\mathcal{D},L)$
+  * goal of HP optim: find $\boldsymbol\lambda$ minimising expected generalisation error, i.e. $\argmin_{\boldsymbol\lambda\in\boldsymbol\Lambda} GE(\mathcal{I},\boldsymbol\lambda,\mathcal{D},L)$
 - \citep[p. 3]{10.1145/3610536}
@@
 - formal definition
-  * vector of objectives $c_1,c_2,...,c_m=\boldsymbol{c}:\Lambda\rightarrow\mathbb{R}^m$
+  * vector of objectives $c_1,c_2,...,c_m=\boldsymbol{c}:\boldsymbol\Lambda\rightarrow\mathbb{R}^m$
   * goal: minimise vector
@@
   * \citep[pp. 7f]{10.1145/3610536} + \citep[pp. 198f]{genetic_algos}
-- Pareto set: set of nondominated points $\mathcal{P}:=\{\boldsymbol\lambda\in\Lambda|\nexists\boldsymbol\lambda'\in\Lambda\text{ s.t. }\boldsymbol\lambda'\prec\boldsymbol\lambda\}$
+- Pareto set: set of nondominated points $\mathcal{P}:=\{\boldsymbol\lambda\in\boldsymbol\Lambda|\nexists\boldsymbol\lambda'\in\boldsymbol\Lambda\text{ s.t.
+}\boldsymbol\lambda'\prec\boldsymbol\lambda\}$
 - Pareto front: image of nondominated points
@@
   * weighted sum of objective functions
-    - $\argmin_{\boldsymbol\lambda\in\Lambda} w_i c_i(\boldsymbol\lambda)$ with $\sum_{i=1}^k w_i=1$ and $w_i>0,\forall i=1,...,k$
+    - $\argmin_{\boldsymbol\lambda\in\boldsymbol\Lambda} w_i c_i(\boldsymbol\lambda)$ with $\sum_{i=1}^k w_i=1$ and $w_i>0,\forall i=1,...,k$
     - drawbacks
@@
     - translate all but one objective into constraint, then optimise remaining objective under subject to the constraints
-    - $\text{w.l.o.g.} \argmin_{\boldsymbol\lambda\in\Lambda} c_1(\boldsymbol\lambda) \text{ s.t. } c_2(\boldsymbol\lambda)\le\epsilon_2,...,c_m(\boldsymbol\lambda)\le\epsilon_m$
+    - $\text{w.l.o.g.} \argmin_{\boldsymbol\lambda\in\boldsymbol\Lambda} c_1(\boldsymbol\lambda) \text{ s.t. } c_2(\boldsymbol\lambda)\le\epsilon_2,...,c_m(\boldsymbol\lambda)\le\epsilon_m$
     - conceputally similar to weighted sum: also sensitive to constraints, must be chosen sensibly
@@
 \section{EAGGA}
-- combines all previously mentioned topics: interpretability, HPO, MOO
-- for our problem (interpretability) we cannot use regular EA algorithms as we are also interested in reducing pairwise interactions + non-monotone feature effects
-- this creates high-dim search space as shown in EAGGA paper -> they provide a framework to efficiently traverse through a space that is reduced to only sensible HP
-  configuration (e.g. interaction of A,B with A monot incr and B monot decr -> couldn't be guaranteed -> don't look at it in first place)
-- NSGA-II inspired evolutionary / genetic algorithm
-- introduces group structure for more efficient optimization over the entire search space
+- NSGA-II inspired evolutionary + genetic algorithm
+- uses AUC-ROC as performance metric and 3 interpretation metrics
+  * NF: rel. \# of features used in model
+  * NI: rel. \# of pairwise feature interaction effects
+  * NNM: rel. \# of non-monotone feature effects
+- aims to find high performing models (AUC) with low NF, low NI, and NNM -> ensures better interpretability than simply optimising for performance
+  but creates very large objective space $\check{\boldsymbol\Lambda}$ to be optimised over with
+  tuples $(\boldsymbol\lambda, \boldsymbol{s}, \boldsymbol{I}_s, \boldsymbol{m}_{\boldsymbol{I}_{\boldsymbol{s}}})=\check{\boldsymbol\lambda}\in\check{\boldsymbol\Lambda}$
+  with
+  * $\boldsymbol\lambda\in\boldsymbol\Lambda$ ``regular'' model HP config
+  * $\boldsymbol{s}\in\{0,1\}^p$ vector denoting feature usage
+  * $\boldsymbol{I}_s\in\{0,1\}^{p\times p}$ matrix denoting pairwise interactions
+  * $\boldsymbol{m}_{\boldsymbol{I}_{\boldsymbol{s}}}\in\{-1,0,1\}^p$ vector denoting monotonicity constraint of each feature (-1 decreasing, 0 none, 1 increasing)
+- thus would require to solve $\argmin_{\check{\boldsymbol\lambda}\in\check{\boldsymbol\Lambda}} (GE(\mathcal{I}_{\check{\boldsymbol\lambda}},\mathcal{D}),NF(\hat{f}_{\mathcal{D},\check{\boldsymbol\lambda}}),NI(\hat{f}_{\mathcal{D},\check{\boldsymbol\lambda}}),NNM(\hat{f}_{\mathcal{D},\check{\boldsymbol\lambda}}))$
+- \citep[pp. 540f]{EAGGA}
+- introduces equivalence relation $R$ ``allowed to interact'' on features that significantly reduces search space for more efficient optimization
+- thus: augmented search space $\tilde{\boldsymbol\Lambda}=\boldsymbol\Lambda\times\mathcal{G}$ comprising model HP space and group structure space
+  * each group structure $\boldsymbol{G}\in\mathcal{G}$ consists of $g$ groups
+    - $G_1$ set of excluded features
+    - $\forall k=2,...,g:G_k=(E_k,M_{E_k})$ tuple
+      * $E_k$ set of features allowed to interact with each other
+      * $M_{E_k}$ monotonicity constraint of entire group $k$
+- EA on model hyperparameters $\boldsymbol\Lambda$
+- GGA on group structure space $\mathcal{G}$
+- \citep[pp. 541f]{EAGGA}
+- futher introduces special initialisation of group structures to increase sample-efficiency of EAGGA algorithm compared to random init
+- \citep[pp. 542f]{EAGGA}
 
 \section{Extension to neural networks}
+- original EAGGA applied to XGBoost model proved to vastly outperform union of competitor models w.r.t. dominated hypervolume and comparable or better than ParEGO on
+  extended search space $\check{\boldsymbol\Lambda}$ \citep[pp. 543-545]{EAGGA}
 - why extend it to NNs?

### 2025-03-13 paper/main.tex
add plots

@@
 %\blindtext
+- propose a network architecture specifically suited for the EAGGA algorithm
+- achieve comparable performance as training XGBoost with EAGGA
 \end{abstract}
@@
 \section{Experimental Results and Discussion}
+\begin{figure}
+  \centering
+  \includegraphics[width=0.9\linewidth]{../code/export/plot_test_dhvs.png}  % scale figure, cf. https://tex.stackexchange.com/a/16584
+  \caption{caption}
+  \label{fig-test-dhvs}
+\end{figure}
 
@@
 tqdm \cite{tqdm}
+
+\section{}
+\begin{figure}
+  \centering
+  \includegraphics[width=0.9\linewidth]{../code/export/plot_early_stopping_losses_pareto_set.png}
+  \caption{caption}
+  \label{fig-es-losses}
+\end{figure}
+
+\section{}
+\begin{figure}
+  \centering
+  \includegraphics[width=0.9\linewidth]{../code/export/plot_val_dhvs_over_time.png}
+  \caption{caption}
+  \label{fig-val-dhvs}
+\end{figure}
 

### 2025-03-13 paper/main.tex
bullets 'experimental results and discussion

@@
 \subsubsection{A-posteriori}
+\label{sec-moo-post}
 - problem a priori
@@
       * comprises of "sub-NNs", one for each equivalence relation -> basically non-fully connected NN
-      * hidden layers use ReLU activation and dropout afterwards, then a shared output layer with concatenated sub-NNs' outputs as input and sigmoid activation
-        (implicit in the loss function for better numerical stability, NN itself outputs "logits", which refers to pre-activation output in pytorch)
+      * hidden layers use ReLU activation and dropout afterwards, then a shared output layer with concatenated sub-NNs' outputs as input and sigmoid activation with
+         binary cross entropy loss (sigmoid implicit in the loss function for better numerical stability, NN itself outputs "logits", which refers to pre-activation output in pytorch)
     - feature sparsity thus achieved by only training on included feature groups -> goes somewhat against of deep learning where model is supposed to decide itself,
@@
 \end{figure}
+- Figure \ref{fig-test-dhvs} suggests comparable performance to XGBoost EAGGA
+- overview over pareto sets can be accessed on our github repo at ~/code/export/*.csv
+  * NN HPs
+    - total layers mostly 3-4, most commonly 3, goes as high as 7 (Satellite (36), diabetes (8))
+    - nodes per hidden layer mostly 3-6, goes as high as 12 (diabetes (8))
+    - p dropout goes as high as 0.7 (climate (18), spambase (57)), but mostly in the 0.1 to 0.3 range
+  * group structures
+    - great diversity in NF across datasets
+      * phoneme (5) up to 1
+      * blood (4) up to 1
+      * banknote (4) up to 0.75
+      * diabetes (8) up to 0.5
+      * climate (18) up to 0.67
+      -> low p datasets in the benchmark tend to have higher NF
+        - possible consequence of shorter evaluation time
+        - low p datasets have much higher \# generations
+        - group structure space much more likely to be exhausted, i.e. more exploration of NN hps
+    - NI, NNM consequently (bounded by NF, can never be more than max \# for respective \# included features) rather low low
+  * dhv contributions
+    - predominately low for fitted models
+    - mostly highest for featureless learner predicting majority class
+    -> sign of good exploration of pareto front + stable estimate, refer Section \ref{sec-moo-post}
+- loss graphs Figure \ref{fig-es-losses}
+  * suggests models could have benefitted from longer training on some datasets, as some loss curves haven't converged when stopping criterion hit
+  * crit was likely triggered by short-term spike, thus could potentially be resolved by comparing average of last k loss against average of >patience< losses prior
+    to that to not be as exposed to short-term spikes in loss
+  * on other datasets, graphs suggest earlier stop would have been totally fine as the networks have long converged, but longer training likely not an issue as loss
+    graphs come from early stopping dataset portion, which is disjunct from training set
+- dhv over generations Figure \ref{fig-val-dhvs}
+  * compute on val sett, i.e. had 5 folds per individual
+    - NF, NI, NNM always the same for each fold
+    - but 5 different AUCs
+    -> hypervolume of (mean(AUC 1, AUC 2, ..., AUC 5), NF, NI, NNM)
+  * artifacts / drops along y likely due to inconsistent computation of Pareto front by third-party library
+    - preliminary experiments on dummy pareto fronts: noticed nds function returning different rankings for same front
+    - returned ranking switched back and forth between two slightly different options (only 1 or 2 indices were swapped)
+    - not sure what caused this, as made same observation using another library that was planned as alternative
+  * all but 3 datasets (2 of which only trained for 1 generation, aynway) show improvement of dominated hypervolume over generations
+  * BUT: absolute as well as improvement almost negligible -> no large dhv drop would we just evaluate the models gotten from the detectors
 
@@
   \includegraphics[width=0.9\linewidth]{../code/export/plot_early_stopping_losses_pareto_set.png}
-  \caption{caption}
+  \caption{Pareto set loss of all datasets, evaluated on early stopping set. Same colours denote losses come from folds of the same individual.
+            x-axis portrait epochs, y-axis binary cross entropy loss.}
   \label{fig-es-losses}
@@
   \includegraphics[width=0.9\linewidth]{../code/export/plot_val_dhvs_over_time.png}
-  \caption{caption}
+  \caption{Dominated hypervolume over generations, evaluated on validation set.
+            x-axis portrait generations, y-axis dominated hypervolume using mean AUC over folds.}
   \label{fig-val-dhvs}

### 2025-03-13 paper/main.tex
add bullets here and there

--- 2025-03-13 paper/main.tex
@@
       * comprises of "sub-NNs", one for each equivalence relation -> basically non-fully connected NN
-      * hidden layers use ReLU activation and dropout afterwards, then a shared output layer with concatenated sub-NNs' outputs as input and sigmoid activation
-        (implicit in the loss function for better numerical stability, NN itself outputs "logits", which refers to pre-activation output in pytorch)
+      * hidden layers use ReLU activation and dropout afterwards, then a shared output layer with concatenated sub-NNs' outputs as input and sigmoid activation with
+        (sigmoid implicit in the loss function for better numerical stability, NN itself outputs "logits", which refers to pre-activation output in pytorch)
+      * loss is binary cross entropy loss, optimizer is AdamW with default params
     - feature sparsity thus achieved by only training on included feature groups -> goes somewhat against of deep learning where model is supposed to decide itself,
@@
     -> add graphic of entire dataset split
-    - then evaluate on last fold from CV + keep best $\mu$ individuals + generate $\lambda=10$ offspring for new generation + repeat
+    - then evaluate on last remaining from CV + keep best $\mu$ individuals + generate $\lambda=10$ offspring for new generation + repeat
+    - final evaluation of individuals of pareto set
+      * for each individual, train model on training set of hold-out for max \# of epochs taken during CV-training
+      * decided on max \# epochs instead of mean after looking at loss graphs in preliminary experiments, also refer Figure \ref{fig-es-losses}
+      * those show that there is no uptick in losses on the early stopping portion (disjunct from training portion) of the set, hence max reasonable
   * hardware: training on Sagemaker Notebook instance
@@
     - thus decided for more economic + ressourcen-schonend t3.medium
+  * did not train on philippine and gina datasets (308, 970 features, respectively), as they ran out of memory when computing interaction detectors
   * known bugs
@@
 \end{figure}
+- preliminary
+  * our implementation of feature + interaction detectors likely tend to include
+    - more features + interactions for high p datasets
+    - less for low p datasets
+    - remember: original paper samples \# of features included from truncated geometric distribution, what is not mentioned is probability of this distribution
+      is determined from fitting 10 trees + looking at relative \# of features used (cf. their github repo ~/R/TunerEAGGA.R function get\_n\_selected\_rpart())
+    -> mlr3 default decision tree max depth 30
+      * i.e. datasets with p < 30 likely use all features, which translates to trunc geom prob = 1
+      * vice-versa for p >> 30 datasets relative \# features < 0.5 (our trunc geom prob)
+      * similar reasoning for pairwise interactions
+  * individuals not in Pareto sets occassionally exhibit AUC < 0.5
+    - this is not by mistake, usually AUC on binary target can simply be inverted (simply predict the opposite class)
+    - this cannot easily be inverted due to monotonicity constraints (weights cannot simply be multiplied by -1 if constraint)
+    - could happen e.g. due to AdamW and weight clipping (employed to enforce monotonicity)
+      * AdamW uses momentum
+      * possible scenario: large momentum would yield negative weights, but after epoch they're clipped to [0,infty) + training stops due to early stopping
+      * weight clipping very imperfect way of enforcing monotonicity, but currently in pytorch unfortunately only way to implement this
+- Figure \ref{fig-test-dhvs} suggests comparable performance to XGBoost EAGGA
+- overview over pareto sets can be accessed on our github repo at ~/code/export/*.csv
+  * NN HPs
+    - total layers mostly 3-4, most commonly 3, goes as high as 7 (Satellite (36), diabetes (8))
+    - nodes per hidden layer mostly 3-6, goes as high as 12 (diabetes (8))
+    - p dropout goes as high as 0.7 (climate (18), spambase (57)), but mostly in the 0.1 to 0.3 range
+  * group structures
+    - great diversity in NF across datasets
+      * phoneme (5) up to 1
+      * blood (4) up to 1
+      * banknote (4) up to 0.75
+      * diabetes (8) up to 0.5
+      * climate (18) up to 0.67
+      -> low p datasets in the benchmark tend to have higher NF
+        - possible consequence of shorter evaluation time
+        - low p datasets have much higher \# generations
+        - group structure space much more likely to be exhausted, i.e. more exploration of NN hps
+    - NI, NNM consequently (bounded by NF, can never be more than max \# for respective \# included features) rather low low
+  * dhv contributions
+    - predominately low for fitted models
+    - mostly highest for featureless learner predicting majority class
+    -> sign of good exploration of pareto front + stable estimate, refer Section \ref{sec-moo-post}
+- loss graphs Figure \ref{fig-es-losses}
+  * suggests models could have benefitted from longer training on some datasets, as some loss curves haven't converged when stopping criterion hit
+  * crit was likely triggered by short-term spike, thus could potentially be resolved by comparing average of last k loss against average of >patience< losses prior
+    to that to not be as exposed to short-term spikes in loss
+  * on other datasets, graphs suggest earlier stop would have been totally fine as the networks have long converged, but longer training likely not an issue as loss
+    graphs come from early stopping dataset portion, which is disjunct from training set
+- dhv over generations Figure \ref{fig-val-dhvs}
+  * compute on val sett, i.e. had 5 folds per individual
+    - NF, NI, NNM always the same for each fold
+    - but 5 different AUCs
+    -> hypervolume of (mean(AUC 1, AUC 2, ..., AUC 5), NF, NI, NNM)
+  * artifacts / drops along y likely due to inconsistent computation of Pareto front by third-party library
+    - preliminary experiments on dummy pareto fronts with 10x4 metrics: noticed nds function returning different rankings for same front
+    - returned ranking switched back and forth between two only slightly different options (only 1 or 2 indices were swapped)
+    - not sure what caused this, as made same observation using another library that was planned as alternative
+    - thus unfortunately not fixable for me
+  * all but 3 datasets (2 of which only trained for 1 generation, aynway) show improvement of dominated hypervolume over generations
+  * BUT: absolute as well as relative improvement almost negligible -> no large final dhv decrease would we just have evaluated the models gotten from the detectors
+    - this was also the reason we didn't run EAGGA on philippine and gina simply without the detectors (as that's where they crashed) -> evidence suggested that detectors are vital to initial performance
+    - original paper supports this assumption \citep[Fig. 4, p. 545]{EAGGA}
 
@@
 - MO BO on group structure space possible?
-- here is a citation \cite{EAGGA}.
 
@@
 
-\section{}
+\section{Plots}
 \begin{figure}
@@
   \includegraphics[width=0.9\linewidth]{../code/export/plot_early_stopping_losses_pareto_set.png}
-  \caption{caption}
+  \caption{Pareto set loss of all datasets, evaluated on early stopping set. Same colours denote losses coming from folds of the same individual.
+            x-axis portrait epochs, y-axis binary cross entropy loss.}
   \label{fig-es-losses}
@@
 
-\section{}
 \begin{figure}

### 2025-03-13 paper/main.tex
bullets conclusion

@@
 - most common type of data
-- bad performance of deep learning models on tabular data, recent research suggests that heavy regularisation is necessary
+- still difficult for neural networks  % TODO: reference, or better only in introduction
+- recent research suggests that heavy regularisation is necessary  % TODO: reference, or better only in introduction
 
@@
 - Figure \ref{fig-test-dhvs} suggests comparable performance to XGBoost EAGGA
+  * unfortunately no comparison to unrestricted NN
+  * would not be sensible, as NF, NI, NNM would be 1, hence hypervolume = 0, as values along 3 dimensions would be at reference point
 - overview over pareto sets can be accessed on our github repo at ~/code/export/*.csv
@@
   * dhv contributions
+    - measures contribution of an individual to the hypervolume, i.e. difference between hypervolume of entire pareto front vs hypervolume of pareto front without
+      individual $\boldsymbol\lambda$: $\text{CON}_{\mathcal{P}}(\boldsymbol\lambda)=\text{HYP}(\mathcal{P})-\text{HYP}(\mathcal{P}\setminus\{\boldsymbol\lambda\})$,
+      where $\text{HYP}(\mathcal{P})$ denotes the hypervolume induced by the pareto set $\mathcal{P}$ \citep[p. 384]{10.5555/1943267.1943271}
     - predominately low for fitted models
@@
 \section{Conclusion and Future Outlook}
+(Conclusion)
+- tabular data still difficult discipline for neural networks
+- research suggests heavy regularisation to improve performance on tabular data
+- EAGGA proved to be successful in making XGBoost more interpretable while keeping performance on par with unregularised XGBoost
+- we extend EAGGA to neural networks to see if we can utilise the regularisation induced by it to make NNs both interpretable and performant on tabular data
+- as consequence propose new architecture allowing to model equivalence relations of EAGGA
+- found overall performance comparable to that of XGBoost fitted using EAGGA, which is a plus, but no outperformance
+
+(Future Outlook)
 - MO BO on group structure space possible?

### 2025-03-14 paper/main.tex
introduction bullets

@@
 %\blindtext
-- propose a network architecture specifically suited for the EAGGA algorithm
-- achieve comparable performance as training XGBoost with EAGGA
+- tabular data still difficult for NNs, where it's still outperformed by other ML model classes
+- research suggests NN performance benefits from heavy regularisation
+- using EAGGA, we regularise an NN and achieve both comparable performance as XGB on EAGGA as well as interpretability
+- for this, we propose a network architecture specifically suited for the EAGGA algorithm
 \end{abstract}
@@
 - most common type of data
-- still difficult for neural networks  % TODO: reference, or better only in introduction
-- recent research suggests that heavy regularisation is necessary  % TODO: reference, or better only in introduction
+- still difficult for neural networks
+- \citep[p. 7499]{Borisov_2024}
+- recent research suggests that strong regularisation is beneficial to NN performance on tab data \citep[8]{NEURIPS2021_c902b497}
 
-regularisation
-- LM: ridge (L2) + lasso (L1)
-  * ridge: overcome multi-collinearity by pulling coeffs close to 0
-  * lasso: variable selection by setting coeffs = 0
-- NN:
-  * L2 regularisation implicitly included in SGD when using weight decay (pull params towards 0) -> does this combat multi-collinearity as well or serve another purpose?
-  * L1
-    + key claim of DL: feature selection obsolete, network finds important ones itself
-    + still, feature usually always has some impact (back this up), no explicit feature selection
-    + explicit feature selection simply by not inluding a feature
-  * another classical NN regularisation technique: dropout against co-adaptation
-  * early stopping against overfitting when using iterative method
--> all of those require specifying Hyperparameters in advance
+- we know regularisation from linear models can come with improvements in interpretabiltiy
+  * e.g. LM-LASSO (L1) regularisation as feature selection -> reduces \# features used in model by setting some coeffs to 0 \citep[p. 267]{lasso}
+- other forms of regularisation improve performance
+  * e.g. LM-Ridge (L2) on multi-collinear data \citep[p. 55]{ridge}  % TODO: mention in "extension" section that L2 is implemented via weight decay
+  * e.g. NN dropout, reduces co-adaptation \citep[p. 1]{hinton2012improvingneuralnetworkspreventing}
+  * e.g. NN early stopping, reduces overfitting on training data \citep[p. 778f]{early_stopping}
+
+- we want to explore if we can use NN regularisation to tackle both interpretability and improved performance (i.e. ``comparable'' to XGBoost) on tab data
+- using EAGGA framework, which proved it can improve performance while keeping (already high) performance of XGBoost on tab data -> see if interpretabiltiy
+  improvements translate to NN + performance can also be on par
 
@@
 - we don't employ NAS for our extension as research focusses on the NLP and image domains, % TODO: find some more references
-  whose datasets exhibit strong correlation amongst features (i.e. tokens or pixels, respectively), an effect that is much weaker for tabular data \citep[p. 1]{Borisov_2024}
+  whose datasets exhibit strong correlation amongst features (i.e. tokens or pixels, respectively), an effect that is much weaker for tabular data \citep[p. 7499]{Borisov_2024}
 - at least want to give a little overview over most notable approaches

### 2025-03-14 paper/main.tex
add NN architecture + dataset split visualisation

@@
 that in its fully-connected feedforwad form is inherently non-interpretable.
-% alternative
+% alternative formulation
 %As fully-connected feedforward neural networks (NNs) are inherently non-interpretable models, the taxonomy introduced by \citet[chap. 2]{survey_NN_interpretability} proves
@@
       * binary forced choice: human evaluator chooses which of two explanations he finds better
-      * forward simulation: evaluator must correctly simulate model output when presented model input + explanation % perhaps binary forced choice is already enough to include, this (and the third one is a bit awkward to explain why this is good for interpretability imo)
+      * forward simulation: evaluator must correctly simulate model output when presented model input + explanation % TODO: perhaps binary forced choice is already enough to include, this (and the third one is a bit awkward to explain why this is good for interpretability imo)
   * \textbf{functionally-grounded evaluation}
@@
     - for a position $k$ in their HP configs, with certain p, swap their values after $k$, yielding 2 children
-    %use example from book page 12
+    % TODO: use example from book page 12
   * \textbf{mutation}: with certain p, change values of an individual, e.g. by adding Gaussian noise (Gaussian mutation) to real values, or flipping bit for binary values
@@
     - entire group structure is passed to NN, where architecture is built accordingly
+\begin{figure}
+  \centering
+  \includegraphics[width=\linewidth]{./architecture.png}  % scale figure, cf. https://tex.stackexchange.com/a/16584
+  \caption{TODO: caption}
+  \label{fig-nn-architecture}
+\end{figure}
   * neural network
@@
       -> equation how monotonicity is enforced with this
+\begin{figure}
+  \centering
+  \includegraphics[width=\linewidth]{./dataset_split.png}
+  \caption{TODO: caption}
+  \label{fig-dataset-split}
+\end{figure}
   * evaluation, holdout, cv, early stopping
@@
   \centering
-  \includegraphics[width=0.9\linewidth]{../code/export/plot_test_dhvs.png}  % scale figure, cf. https://tex.stackexchange.com/a/16584
-  \caption{caption}
+  \includegraphics[width=0.9\linewidth]{../code/export/plot_test_dhvs.png}
+  \caption{TODO: caption}
   \label{fig-test-dhvs}

### 2025-03-14 paper/main.tex
add interaction equations

@@
       -> equation why the interacting features need to be grouped together when using ReLU, cf. photos
+      * given layer input $\boldsymbol{x}$ and omitting bias
+      * $\text{ReLU}(\boldsymbol{x})=\max(0,\sum_{j=1}^p w_p x_p)$
+      * interaction not of the form $x_j \cdot x_{j'}$ as used to from LMs but of a form where the sum $w_j x_j+w_{j'} x_{j'}$ decides whether $x_j,x_{j'}$
+        pass through the activation or not
+      * in other words: $\max(0,w_j x_j + w_{j'} x_{j'})\neq\max(0,w_j x_j)+\max(w_{j'} x_{j'})$ and thus there is an interaction
     - monotonicty constraint achieved by clipping weights to [0, infty) for restricted equivalence relations (monotonic decrease achieved via dataset object multiplying

### 2025-03-14 paper/main.tex
add monotonicity constraint equations

@@
       -> equation why the interacting features need to be grouped together when using ReLU, cf. photos
-      * given layer input $\boldsymbol{x}$ and omitting bias
+      * given layer input $\boldsymbol{x}$ and w.l.o.g. omitting bias
       * $\text{ReLU}(\boldsymbol{x})=\max(0,\sum_{j=1}^p w_p x_p)$
@@
       -> equation how monotonicity is enforced with this
+      * consider vector form computation of hidden neuron $\boldsymbol{o}_{\text{in}}^{(l)}\in\mathbb{R}^{p^{(l)}}$ of layer $l$
+        - $\boldsymbol{o}_{\text{in}}^{(l)}=\boldsymbol{W} \boldsymbol{o}_{\text{out}}^{(l-1)} + \boldsymbol{b}$
+        - where $\boldsymbol{o}_{\text{in}}^{(l)}$ refers to pre-activation values, $\boldsymbol{o}_{\text{out}}^{(l)}$ to value after applying ReLU
+        - weight matrix $\boldsymbol{W}\in\mathbb{R}^{p^{(l)} \times p^{(l-1)}}$
+        - bias $\boldsymbol{b}\in\mathbb{R}^{p^{(l)}}$
+      * definition of monotonically increasing function $f$: for $x_1\le x_2$ it must hold that $f(x_1)\le f(x_2)$
+      * $\boldsymbol{W} \boldsymbol{o}_{\text{out}}^{(l-1)} + \boldsymbol{b}$ clearly only monotonic if and only if $\boldsymbol{W}\in\mathbb{R}_{0+}{p^{(l-1)}\times p^{(l)}}$
+        - where $\mathbb{R}_{0+}=[0,\infty)$
+        - hence weight clipping necessary, bias clipping not because only constant term
+      * $\boldsymbol{o}_{\text{out}}^{(l)}=ReLU(\boldsymbol{o}_{\text{in}}^{(l)})=\max(0,\boldsymbol{o}_{\text{in}}^{(l)})$ monotonic
 \begin{figure}
@@
       * weight clipping very imperfect way of enforcing monotonicity, but currently in pytorch unfortunately only way to implement this
-- Figure \ref{fig-test-dhvs} suggests comparable performance to XGBoost EAGGA
-  * unfortunately no comparison to unrestricted NN
-  * would not be sensible, as NF, NI, NNM would be 1, hence hypervolume = 0, as values along 3 dimensions would be at reference point
+- Figure \ref{fig-test-dhvs}
+  * suggests comparable performance to XGBoost EAGGA
+  * did not compare to unrestricted NN, would not be sensible, as NF, NI, NNM would be 1, hence hypervolume = 0, as values along 3 dimensions would be at reference point
+  * as XGBoost EAGGA, NN EAGGA consistently outperforms union of competitors as evident from \citep[Figure 2]{EAGGA}
+  * unfortunately due to time constraints no own union comparison with NN instead of XGBoost
 - overview over pareto sets can be accessed on our github repo at ~/code/export/*.csv

### 2025-03-14 paper/main.tex
some changes to bullets, started finalising interpretability wording

@@
 where \textit{explanations} are logical decision rules and \textit{understandable terms} relate to commonly used terms in the domain of the problem,
-as suggested by \citet[chap. 1]{survey_NN_interpretability}.
-Further, we use the term ``explanability'' in an exchangeable manner with ``interpretability'', as is commonly done.
+as suggested by \citet[chap. 1]{survey_NN_interpretability}. Further, we use the term ``explainability'' in an exchangeable manner with ``interpretability'',
+as is commonly done.
 
-(Why Interpretability is desirable?)
-- in many ways desirable, as per \citet[pp. 3-4]{Zach2019InterpretabilityOD}, e.g.
-  * \textbf{gaining trust}: most notable use case, e.g. medical diagnosis: predictions need to be validated by doctors \citep[p. 1]{review_NN_interpretability}
-  * \textbf{model debugging}
-    - model is optimised w.r.t. loss and judged on accuracy
-    - might have unexpected performance drops in certain situations, makes model unreliable \citep[1B]{survey_NN_interpretability}
-    - having an interpretable model + model induction process can help find ways to improve accuracy + reliability
-  * \textbf{scientific understanding}: when only models can make sense of increasingly complex data anymore, interpretability enables extraction of learnt knowledge encoded
-    in the model to make it accessible + reliable to humans (also mentioned here \citep[p. 1]{review_NN_interpretability})
-  * \textbf{subconcious bias}, e.g. loan approval: must ensure that decision is not discriminatory
-  * \textbf{regulatory}
-    - such as EU's "Right to Explanation" warranted by the GDPR \citep[p. 1]{review_NN_interpretability}, \citep[1B]{survey_NN_interpretability} or
-    - drug approval processes in situations where a machine learning model discovered a new drug, process needs to be transparent for the regulator to approve \citep[1B]{survey_NN_interpretability}
+% Why Interpretability is desirable?
+Explainability of a model's reasoning is in many ways desirable. \citet[pp. 3-4]{Zach2019InterpretabilityOD} gives a range of examples,
+amongst which are \textit{gaining trust}, e.g. when doctors rely on medical diagnosis predictions, avoiding \textit{subconcious biases} by
+making sure loan approvals are non-discriminatory, or \textit{regulatory} reasons, most notably the EU's ``Right to Explanation'' warranted by the
+GDPR \citep[p. 1]{review_NN_interpretability} or for approval of drugs discovered using machine learning models \citep[1B]{survey_NN_interpretability}.
+It can further prove helpful for explaining unexpected drops in model performance, which could arise from optimising a model with respect to loss and then judging its
+performance on a different metric such as accuracy, a practice known as \textit{model debugging} \citep[1B]{survey_NN_interpretability} or for
+\textit{scientific understanding} in domains where only models can make sense of increasingly complex data anymore and learnt knowledge encoded in a model needs to be
+made accessible to humans to be used reliably \citep[p. 1]{review_NN_interpretability}.
 
@@
 Amongst post-hoc techniques, we can further divide the space into model-specific (such as analysing GLM coefficients) and model-agnostic
-(e.g. partial depence plots, ALE) methods. \citet[chap. 3.2]{molnar2022}
+(e.g. partial depence plots, ALE) methods \citep[chap. 3.2]{molnar2022}.
 
-(Taxonomy)
-\citet[chap. 2]{survey_NN_interpretability} extend this distinction to a 3-dimensional taxonomy, allowing for better categorisation of neural networks (NN), a model class
-that in its fully-connected feedforwad form is inherently non-interpretable.
-% alternative formulation
-%As fully-connected feedforward neural networks (NNs) are inherently non-interpretable models, the taxonomy introduced by \citet[chap. 2]{survey_NN_interpretability} proves
-%particulary useful to categorise NN explanability methods along 3 dimensions:
+% Taxonomy
+\citet[chap. 2]{survey_NN_interpretability} extend this distinction to a three-dimensional taxonomy, allowing for better categorisation of neural networks (NNs),
+a model class that in its fully-connected feedforwad form is inherently non-interpretable.
 - \textbf{Passive vs Active Approaches}
@@
     - idea: evaluate ``interpretability model'' directly w.r.t. the task -> human experts evaluate the outcome
-    - e.g.: model performing medical diagnosis on patients would be evaluated by based on doctors doing the same
+    - e.g.: model performing medical diagnosis on patients would be evaluated based on doctors doing the same
   * \textbf{human-grounded metrics}
@@
 - based on a population, which iteratively (iteration = generation) generates $\lambda$ offspring via 3 operators (all sub-bullets from \citet[pp. 10-]{genetic_algos})
-  * \textbf{reproduction}: pick an individual to reproduce with p proportional to its fitness
+  * \textbf{reproduction}: pick an individual to reproduce with prob proportional to its fitness
   * \textbf{crossover}
     - select 2 ``parents'' from pool of reproducing individuals
-    - for a position $k$ in their HP configs, with certain p, swap their values after $k$, yielding 2 children
+    - for a position $k$ in their HP configs, with certain prob, swap their values after position $k$, yielding 2 new children
     % TODO: use example from book page 12
-  * \textbf{mutation}: with certain p, change values of an individual, e.g. by adding Gaussian noise (Gaussian mutation) to real values, or flipping bit for binary values
+  * \textbf{mutation}: with certain prob, change values of an individual, e.g. by adding Gaussian noise (Gaussian mutation) to real values, or flipping bit for binary values
 - selection at end of each genertion: keep $\mu$ best individuals (based on some fitness function), either from only offspring (``$(\mu,\lambda)$-selection'')
@@
   whose datasets exhibit strong correlation amongst features (i.e. tokens or pixels, respectively), an effect that is much weaker for tabular data \citep[p. 7499]{Borisov_2024}
-- at least want to give a little overview over most notable approaches
+- at least want to give a little outline on most notable approaches
   * \textbf{cell search space}
@@
     - trains one overall ``fabric'' comprising all architectures of search space
-    - visualisation as DAG, nodes are nodes, edges in-between are operations on a node
+    - visualisation as DAG, nodes are layers, edges in-between are operations on a layer's values
     - each path through DAG (from input to output node) represents one architecture of search space -> train entire DAG, then pick optimal path (architecture)
@@
 - thus aim to find trade-off solutions of non-dominated points
-- we say a point dominates another
+- we say a ``point dominates another''
   * if there is no other point strictly better in at least one dimension and better or equal in the remaining ones
@@
 - if no knowledge over true Pareto-front, volume-based approaches are popular, which measure volume between Pareto front estim and some chosen
-  reference point (usually worst point in objective space)
+  reference point (usually worst point in objective space), e.g. 0 for accuracy (optimum is 1)
 - \citep[pp. 8-10]{10.1145/3610536}
@@
   * $\epsilon$-constraint
-    - translate all but one objective into constraint, then optimise remaining objective under subject to the constraints
+    - translate all but one objective into constraint, then optimise remaining objective subject to the constraints
     - $\text{w.l.o.g.} \argmin_{\boldsymbol\lambda\in\boldsymbol\Lambda} c_1(\boldsymbol\lambda) \text{ s.t. } c_2(\boldsymbol\lambda)\le\epsilon_2,...,c_m(\boldsymbol\lambda)\le\epsilon_m$
@@
 - \textbf{lexicographic method}
-  * define priorisation of objectives
+  * define priority of objectives
   * greedily optimise objectives in order of priority, constraint to the solutions of the already optimised (higher-priority) objectives
@@
   * a-posteriori evaluates configs just as a-priori, but keeps track of multiple ``best'' (non-dominated) solutions -> makes relationship between HP config + objectives visible
-- aside from usualy baselines grid and random seach there are multi-objective BO adaptations, mainly using either of two approaches
+- aside from usual baselines grid and random seach there are multi-objective BO adaptations, mainly using either of two approaches
   (1) fit single surrogate model on scalarised objectives, e.g. ParEGO, which employs the augemented Tchebycheff function as scalarisation to ensure the Pareto front is
@@
     - either one acquisition function per objective -> return set of promising next candidates
-    - or one overall acquisition function aggregating surrogates, e.g. EHI, maximises expected improvement of hypervolume \citep[p. 16]{10.1145/3610536} + \citep[pp. 8f]{EHI}
+    - or one overall acquisition function aggregating surrogates, e.g. \textit{EHI}, maximises expected improvement of hypervolume \citep[p. 16]{10.1145/3610536} + \citep[pp. 8f]{EHI}
 - lastly, there is also multi-objective EA (MOEA) algorithms to explore Pareto front, one of which is NSGA-II (Nondominated Sorting Genetic Algorithm)
@@
   * difference to SOEA: ranking of individuals (SOEA uses scalar fitness, MOEA mutliple objectives)
-    - ranking mechanism based on 2 parts
+    - multi-objective ranking mechanism based on 2 parts
       * non-dominated sorting, ensures elitism
@@
         (2) remove pareto front from population
-        (3) if individuals left, repeat from 1, increment rank \#
+        (3) if any individuals left, repeat from 1, increment rank \#
         - \citep[p. 201]{genetic_algos} + \citep[pp. 183f]{NSGA_II}
@@
         - crowding distance of an individual = mean side length of cuboid spanned by its nearest neighbours as vertices in objective space
-        - individuals without two neighbour in any dimension are assigned infinitely high distance value
+        - individuals without two neighbours in a dimension are assigned infinitely high distance value
         - \citep[p. 185]{NSGA_II}
@@
       -> rank individuals by nds front rank (ascending) + crowding distance (descending) as tie breaker
-  * ranking used for selecting $\mu$ best individuals to keep for next generation + for reproduction: via binary tournament solution: select two random individuals,
-    pick best w.r.t. nds + cd for offspring creation (mutation / crossover)
+  * ranking used for selecting $\mu$ best individuals to keep for next generation + for reproduction: via binary tournament selection: select two random individuals,
+    pick best w.r.t. nds + cd for offspring creation (i.e. mutation / crossover)
   * using nds for ranking / fitness makes intuitive sense, but why cd?
@@
   * NNM: rel. \# of non-monotone feature effects
-- aims to find high performing models (AUC) with low NF, low NI, and NNM -> ensures better interpretability than simply optimising for performance
+- aims to find high performing models (AUC) with low NF, NI, and NNM -> ensures better interpretability than simply optimising for performance
   but creates very large objective space $\check{\boldsymbol\Lambda}$ to be optimised over with
@@
 - thus: augmented search space $\tilde{\boldsymbol\Lambda}=\boldsymbol\Lambda\times\mathcal{G}$ comprising model HP space and group structure space
-  * each group structure $\boldsymbol{G}\in\mathcal{G}$ consists of $g$ groups
+  * each group structure $\boldsymbol{G}\in\mathcal{G}$ consists of $g$ groups:
     - $G_1$ set of excluded features
@@
 - EA on model hyperparameters $\boldsymbol\Lambda$
-- GGA on group structure space $\mathcal{G}$
+- GGA on group structure space $\mathcal{G}$ with adapted mutation and crossover operators
 - \citep[pp. 541f]{EAGGA}
@@
   (general algorithm)
-  * EAGGA algorithm implemented just as described in the paper
+  * EAGGA algorithm implemented largely as described in the paper
   * HP init
-    - NN total layers $\in\{3, ..., 10\}$, i.e. hidden layers $\in\{1, ..., 8\}$, init from trunc geom with p=0.5
-    - NN nodes per hidden layer (for each group) $\in\{3, ..., 20\}$, init from trunc geom with p=0.5
+    - NN total layers $\in\{3, ..., 10\}$, i.e. hidden layers $\in\{1, ..., 8\}$, init from trunc geom with prob=0.5
+    - NN nodes per hidden layer (for each group) $\in\{3, ..., 20\}$, init from trunc geom with prob=0.5
     - NN dropout \% $\in[0, 1]$, init from trunc gamma with shape=2, scale=0.15
@@
       * as described in paper
-      * but instead of fitting 10 trees + taking rel. \# of features used as p for trunc geom, we simply use 0.5 (sklearn dectree examination not straightforward)
+      * but instead of fitting 10 trees + taking rel. \# of features used as prob for trunc geom, we simply use 0.5 (sklearn dectree examination not straightforward)
       * also in preliminary experiments we found the sampled \# of features to be used occassionally to be > \# of non-0 values in normalised information gain filter
-        (e.g. if former is = total \# of features and one feature is indep. from target) in these edge cases we only use the features with non-0 filter values
+        (e.g. if former is = total \# of features and one feature is indep. from target) in these edge cases we simply use all features with non-0 filter values
     - interaction detector
       * as described in paper
-      * same as for feat detect: use p=0.5 to sample \# of interactions used instead of 10 dec trees
+      * same as for feat detect: use prob=0.5 to sample \# of interactions used instead of 10 dec trees
       * also for FAST algorithm don't use RSS (lin reg metric) but mean accuracy, as we fit log reg
@@
       * as described in paper
-      * use default HPs (max depth 30, minsplie = 20) of mlr3 classification tree implementation, as this is the library the paper uses
+      * use default HPs (max depth 30, minsplit = 20) of mlr3 classification tree implementation, as this is the library the paper uses
     -> both interaction + monotonicity detectors fit their models on 80\% of train split from holdout + eval on remaining 20\%
@@
       monotonicity detector
-    - included features are passed to dataset, dataset outputs included features, multiplied by the features' individual monotonicity (-1 / 1)
+    - included features are passed to dataset, dataset outputs included features, multiplied by the features' individual monotonicity (-1 / 1), so that group monotonicity
+      can be encoded with only $\{0,1\}$, as proposed in paper
     - entire group structure is passed to NN, where architecture is built accordingly
@@
       * comprises of "sub-NNs", one for each equivalence relation -> basically non-fully connected NN
-      * hidden layers use ReLU activation and dropout afterwards, then a shared output layer with concatenated sub-NNs' outputs as input and sigmoid activation with
-        (sigmoid implicit in the loss function for better numerical stability, NN itself outputs "logits", which refers to pre-activation output in pytorch)
-      * loss is binary cross entropy loss, optimizer is AdamW with default params
-    - feature sparsity thus achieved by only training on included feature groups -> goes somewhat against of deep learning where model is supposed to decide itself,
+      * hidden layers use ReLU activation and dropout afterwards, then a shared output layer with concatenated sub-NNs' outputs as input and sigmoid activation
+      * optimizer is AdamW with default params, loss is binary cross entropy loss (sigmoid thus implicit in the loss function for better numerical stability,
+        NN itself outputs "logits", which refers to pre-activation output in pytorch)
+    - feature sparsity thus achieved by only training on included feature groups -> goes somewhat against notion of deep learning where model is supposed to decide itself,
       which feature to "use" / put importance on -> ELABORATE
@@
       * in other words: $\max(0,w_j x_j + w_{j'} x_{j'})\neq\max(0,w_j x_j)+\max(w_{j'} x_{j'})$ and thus there is an interaction
-    - monotonicty constraint achieved by clipping weights to [0, infty) for restricted equivalence relations (monotonic decrease achieved via dataset object multiplying
-      features with their individual signs), bias clipping not necessary (constant additive term)
+    - monotonicty constraint achieved by clipping weights to [0, infty) after each epoch for restricted equivalence relations (monotonic decrease achieved
+      via dataset object multiplying features with their individual signs), bias clipping not necessary (constant additive term)
       -> equation how monotonicity is enforced with this
@@
         - where $\mathbb{R}_{0+}=[0,\infty)$
-        - hence weight clipping necessary, bias clipping not because only constant term
+        - hence weight clipping necessary, bias clipping not because only constant additive term
       * $\boldsymbol{o}_{\text{out}}^{(l)}=ReLU(\boldsymbol{o}_{\text{in}}^{(l)})=\max(0,\boldsymbol{o}_{\text{in}}^{(l)})$ monotonic
@@
       * for each fold, always train for min 200 epochs, keep track of model with lowest loss
-      * after that, use patience of 100: if current model's loss is > mean of last 100 epochs' losses
+      * after that, use patience of 100: if current model's loss is > mean of last 100 epochs' losses -> stop early
       * if no early stopping, train each fold for max 10 minutes
       * then stop training, return model with lowest loss
-    -> add graphic of entire dataset split
+    -> refere Figure \ref{fig-dataset-split}
     - then evaluate on last remaining from CV + keep best $\mu$ individuals + generate $\lambda=10$ offspring for new generation + repeat
@@
       * decided on max \# epochs instead of mean after looking at loss graphs in preliminary experiments, also refer Figure \ref{fig-es-losses}
-      * those show that there is no uptick in losses on the early stopping portion (disjunct from training portion) of the set, hence max reasonable
+      * those show that there is no uptick in losses on the early stopping portion (disjunct from training portion) of the set, hence max is reasonable
   * hardware: training on Sagemaker Notebook instance
@@
     -> mlr3 default decision tree max depth 30
-      * i.e. datasets with p < 30 likely use all features, which translates to trunc geom prob = 1
+      * i.e. datasets with p < 30 might use all features, which translates to trunc geom prob = 1
       * vice-versa for p >> 30 datasets relative \# features < 0.5 (our trunc geom prob)
@@
     - this is not by mistake, usually AUC on binary target can simply be inverted (simply predict the opposite class)
-    - this cannot easily be inverted due to monotonicity constraints (weights cannot simply be multiplied by -1 if constraint)
+    - for us this cannot easily be inverted due to monotonicity constraints (weights cannot simply be multiplied by -1 if constraint)
     - could happen e.g. due to AdamW and weight clipping (employed to enforce monotonicity)
@@
         - group structure space much more likely to be exhausted, i.e. more exploration of NN hps
-    - NI, NNM consequently (bounded by NF, can never be more than max \# for respective \# included features) rather low low
+    - NI, NNM consequently (bounded by NF, can never be more than max \# for respective \# included features) rather low
   * dhv contributions
@@
   * suggests models could have benefitted from longer training on some datasets, as some loss curves haven't converged when stopping criterion hit
-  * crit was likely triggered by short-term spike, thus could potentially be resolved by comparing average of last k loss against average of >patience< losses prior
+  * crit was likely triggered by short-term spike, thus could potentially be resolved by comparing average of last k losses against average of >patience< losses prior
     to that to not be as exposed to short-term spikes in loss
@@
 - dhv over generations Figure \ref{fig-val-dhvs}
-  * compute on val sett, i.e. had 5 folds per individual
+  * compute on val set, i.e. had 5 folds per individual
     - NF, NI, NNM always the same for each fold
@@
 - as consequence propose new architecture allowing to model equivalence relations of EAGGA
+- ... conforming to \citet[chap. 2]{survey_NN_interpretability} taxonomy it fits as ..., ..., ... %TODO
 - found overall performance comparable to that of XGBoost fitted using EAGGA, which is a plus, but no outperformance
@@
 (Future Outlook)
-- MO BO on group structure space possible?
+- MO BO on group structure space possible, perhaps via restricted BO?
 

### 2025-03-14 paper/main.tex
interpretability taxonomy

@@
 \citet[chap. 2]{survey_NN_interpretability} extend this distinction to a three-dimensional taxonomy, allowing for better categorisation of neural networks (NNs),
-a model class that in its fully-connected feedforwad form is inherently non-interpretable.
-- \textbf{Passive vs Active Approaches}
-  * passive: post-hoc
-  * active: changing architecture or training process to make model interpretable
-- \textbf{Type of Explanations}
-  * examples: providing examples of what leads to desired output
-  * attribution: attributing effect on output for a specific feature
-  * hidden semantics: examine what type of inputs specific neurons / layers pick up on
-  * rules: logical rules, e.g. if-then, trees
-- \textbf{Local vs Global Interpretability}
-  * local: explanation based on individual samples
-  * semi-local: explanation based on set of samples, e.g. grouped together by some criterion
-  * global: explaining network as a whole
+a model class that in its fully-connected feedforwad form is inherently non-interpretable.\\ % TODO: letztes halbsatz vllt lieber in intro unterbringen, hier ist der awkward und unnötig ausschmückend
+\textbf{Passive vs Active Approaches}, where \textit{passive} are all post-hoc methods and \textit{active} methods actively change either the architecture or
+training process to increase model interpretability.\\
+\textbf{Type of Explanations}, distinguishing between \textit{example} methods, providing concrete examples of what leads to a desired output, \textit{attribution}
+methods that attribute the effect on the output for a specific feature, \textit{hidden semantic} methods, which explain the types of inputs particular neurons or layers
+pick up on, and logical \textit{rules}, such as if-then clauses or tree-induced rules.\\
+\textbf{Local vs Global Interpretability}, ranging from \textit{local} methods providing explanations based on individual samples, \textit{semi-local} methods,
+explaining model behaviour for sets of samples grouped by some criterion, to \textit{global} methods, which explain the network as a whole.
 
-(Evaluation)
+% Evaluation
 - objective evaluation difficult, as no clear definition of interpretability

### 2025-03-14 paper/main.tex
interpretability evaluation

@@
 
-% feels better in front of AutoML, as it feels more qualitative, less quantitative
 \subsection{Interpretability}
@@
 % Evaluation
-- objective evaluation difficult, as no clear definition of interpretability
-- \citet[3]{DoshiVelez2017TowardsAR} propose a taxonomy to categorise possible evaluation methods
-  * \textbf{application-grounded evaluation}
-    - most rigorous method, also most expensive + time consuming
-    - idea: evaluate ``interpretability model'' directly w.r.t. the task -> human experts evaluate the outcome
-    - e.g.: model performing medical diagnosis on patients would be evaluated based on doctors doing the same
-  * \textbf{human-grounded metrics}
-    - similar to application-grounded, but tries to simplify task (while preserving its essence) so that laypersons can do it
-    - cheaper due to larger, less qualified subject pool
-    - especially suitable if only general concepts of the tasks need to be validated
-    - examples for evaluation set-up
-      * binary forced choice: human evaluator chooses which of two explanations he finds better
-      * forward simulation: evaluator must correctly simulate model output when presented model input + explanation % TODO: perhaps binary forced choice is already enough to include, this (and the third one is a bit awkward to explain why this is good for interpretability imo)
-  * \textbf{functionally-grounded evaluation}
-    - assess explanation quality via some formally defined proxy for interpretability
-    - no human time nor cost required beyond initial formulation of proxy
-    - e.g. if we already have an interpretable model class (e.g. identified via human-grounded evaluation), can then rank different models of that class based on proxy
-    - main challenge: what proxies
+Given its unclear definition, evaluating interpretability can be challenging. \citet[3]{DoshiVelez2017TowardsAR} propose a taxonomy to categorise
+possible evaluation methods based on their rigorousness.\\
+\textbf{Application-grounded evaluation} evaluates interpretations directly with respect to the task, by having human experts evaluate the outcome and
+is therefore the most expensive and time-consuming of the three approaches.\\
+\textbf{Human-grounded metrics} is similar to application-grounded evaluation in that a human still evaluates the interpretations, but tries to simplify the task
+so that a layperson can do it. Human-grounded approaches are especially suitable if it's sufficient to validate the general concepts of a task. A typical evaluation
+set-up in this category is binary forced choice, where the human evaluator chooses, which of two generated explanations he prefers.\\
+\textbf{Functionally-grounded evaluation} is the least rigorous, but easiest to implement of the three. It assess explanatory quality according to some
+formally defined proxy for interpretability and is particularly useful in ranking different models if their model-class is already identified
+(e.g. via human-grounded evaluation) to be interpretable. The main challenge for functionally-grounded evaluation is finding a good proxy.
 

### 2025-03-14 paper/main.tex
HPO intro

@@
 \subsection{Hyperparameter Optimization (HPO)}
-- in contrast to model parameters, which are optimised during training, hyperparameters are parameters describing the ML algo + are specified before training
-- still often have great impact on final (trained) model performance -> can be optimised too
-- formal definition
-  * dataset $\mathcal{D}\subseteq\mathcal{X}\times\mathcal{Y}$ consisting of $n$ tuples $(\boldsymbol{x}^{(i)}, y^{(i)})\stackrel{i.i.d.}{\sim}\mathbb{P}_{\boldsymbol{x}y}$
-    (data-generating distribution), i.e. $\mathcal{D}=\{(\boldsymbol{x}^{(i)}, y^{(i)})\}_{i=1}^n$
-  * ML algo $\mathcal{I}(\cdot,\boldsymbol\lambda)$ with HP config $\boldsymbol\lambda\in\boldsymbol\Lambda$ maps dataset to model
-    $\hat{f}:\mathcal{X}\rightarrow\mathbb{R}^g\subseteq\mathcal{Y}$ (where $g\in\mathbb{N}$ is \# of classes of target),
-    i.e. $\mathcal{I}:(\mathbb{D}\times\boldsymbol\Lambda)\rightarrow\mathcal{H}$, with $\mathbb{D}$ space of all finite datasets and $\mathcal{H}$ hypothesis space (space of all models $\hat{f}$)
-  * denote $\mathcal{I}_{\boldsymbol\lambda}$ as (untrained) model with fixed HP config $\boldsymbol\lambda$
-  * loss function $L$
-  * goal of model training: for fixed $\boldsymbol\lambda$, have $\mathcal{I}$ find model $\hat{f}$ minimising expected generalisation error
-    $GE(\mathcal{I},\boldsymbol\lambda,\mathcal{D},L)=\mathbb{E}_{(\boldsymbol{x},y)\sim\mathbb{P}_{\boldsymbol{x}y}}[L(y,\mathcal{I}_{\boldsymbol\lambda}(\mathcal{D})(\boldsymbol{x}))]$
-  * goal of HP optim: find $\boldsymbol\lambda$ minimising expected generalisation error, i.e. $\argmin_{\boldsymbol\lambda\in\boldsymbol\Lambda} GE(\mathcal{I},\boldsymbol\lambda,\mathcal{D},L)$
-- \citep[p. 3]{10.1145/3610536}
-- HPO = black-box optim problem -> we can apply any black box algo to tune
+In contrast to model parameters, which are optimized during training, hyperparameters are those describing the machine learning algorithm and are fixed before training.
+Because they usually have a significant impact on the trained model's performance, hyperparameters are often optimized, too.
+Let $\mathcal{D}\subseteq\mathcal{X}\times\mathcal{Y}$ be a dataset consisting of $n$
+tuples drawn from the data-generating distribution, i.e. $(\boldsymbol{x}^{(i)}, y^{(i)})\stackrel{i.i.d.}{\sim}\mathbb{P}_{\boldsymbol{x}y},\forall i=1,...,n$.
+Furthermore, let $\mathcal{I}:(\mathbb{D}\times\boldsymbol\Lambda)\rightarrow\mathcal{H}, (\mathcal{D},\boldsymbol\lambda)\mapsto\hat{f}$
+be an algorithm that maps a given dataset $\mathcal{D}$ and hyperparameter configuration $\boldsymbol\lambda\in\boldsymbol\Lambda$ to a model $\hat{f}$.
+Denote with $\mathcal{I}_{\boldsymbol\lambda}$ an algorithm with fixed hyperparameter configuration a chosen loss function with $L$.
+The goal of model training is, for fixed $\boldsymbol\lambda$, to have $\mathcal{I}_{\boldsymbol\lambda}$ find model $\hat{f}$ minimising the expected generalisation error
+$GE(\mathcal{I}_{\boldsymbol\lambda},\mathcal{D},L)=\mathbb{E}_{(\boldsymbol{x},y)\sim\mathbb{P}_{\boldsymbol{x}y}}[L(y,\mathcal{I}_{\boldsymbol\lambda}(\mathcal{D})(\boldsymbol{x}))]$
+The goal of HPO, on the other hand, is to find $\boldsymbol\lambda$ minimising the expected generalisation error,
+i.e. $\argmin_{\boldsymbol\lambda\in\boldsymbol\Lambda} GE(\mathcal{I}_{\boldsymbol\lambda},\mathcal{D},L)$.
+For this there is no analytical expression available, making HPO a blackbox optimization problem.  \citep[pp. 2f]{10.1145/3610536}
 

### 2025-03-14 paper/main.tex
HPO model-free

@@
 \subsection{Hyperparameter Optimization (HPO)}
-In contrast to model parameters, which are optimized during training, hyperparameters are those describing the machine learning algorithm and are fixed before training.
-Because they usually have a significant impact on the trained model's performance, hyperparameters are often optimized, too.
+In contrast to model parameters, which are optimized during training, hyperparameters (HPs) are those describing the machine learning algorithm and are fixed before training.
+Because they usually have a significant impact on the trained model's performance, HPs are often optimized, too.
 Let $\mathcal{D}\subseteq\mathcal{X}\times\mathcal{Y}$ be a dataset consisting of $n$
@@
 Furthermore, let $\mathcal{I}:(\mathbb{D}\times\boldsymbol\Lambda)\rightarrow\mathcal{H}, (\mathcal{D},\boldsymbol\lambda)\mapsto\hat{f}$
-be an algorithm that maps a given dataset $\mathcal{D}$ and hyperparameter configuration $\boldsymbol\lambda\in\boldsymbol\Lambda$ to a model $\hat{f}$.
-Denote with $\mathcal{I}_{\boldsymbol\lambda}$ an algorithm with fixed hyperparameter configuration a chosen loss function with $L$.
+be an algorithm that maps a given dataset $\mathcal{D}$ and HP configuration $\boldsymbol\lambda\in\boldsymbol\Lambda$ to a model $\hat{f}$.
+Denote with $\mathcal{I}_{\boldsymbol\lambda}$ an algorithm with fixed HP configuration a chosen loss function with $L$.
 The goal of model training is, for fixed $\boldsymbol\lambda$, to have $\mathcal{I}_{\boldsymbol\lambda}$ find model $\hat{f}$ minimising the expected generalisation error
@@
 
-[model free]
-(grid + random search)
-- most basic strategy: grid search, for all HPs define range of interest, then evaluate cartesian product of those -> scales poorly in \# of HP dimensions + \# of query points per HP range
-- random search
-  * randomly sample value for each HP until it runs out of budget
-  * explorative: for budget $B$, each HP will (likely) be queried with $B$ different values, whereas grid search only $B^{1/N}$ for $N$ HPs
-  * thus better than grid in case of some HPs being non-informative (almost always the case), as it doesn't waste time specifically exploring these (as does grid)
-  * usually useful as baseline, as no assumptions about model, easy to parallelise, high exploration, and in expectation will ``achieve performance arbitrarily close to the optimum''
+We will first outline model-free methods to approach blackbox optimization problems, before presenting a model-based variant.
+
+% model free
+% grid + random search
+The most basic model-free strategies for blackbox optimization are grid and random search.
+In the latter, the user defines a range of interest for each HP and then evaluating points along a grid within the cartesian product of those.
+This has the drawback of scaling extremely poorly in both number of HP dimensions and number of query points per range of interest.
+Random search, on the other hand, randomly samples a value for each HP until it runs out of budget. Its explorative nature, ease of use, and no assumptions about the model
+makes it a good baseline. It is also superior to grid search in cases were one or more HPs have little impact on model performance, as for a given budget $B$,
+each HP will likely be queried with $B$ different values, whereas for grid search each HP will only be queried with $B^{1/N}$ different values for $N$ HPs.
 

### 2025-03-14 paper/main.tex
HPO model-free EA intro

@@
 This has the drawback of scaling extremely poorly in both number of HP dimensions and number of query points per range of interest.
-Random search, on the other hand, randomly samples a value for each HP until it runs out of budget. Its explorative nature, ease of use, and no assumptions about the model
-makes it a good baseline. It is also superior to grid search in cases were one or more HPs have little impact on model performance, as for a given budget $B$,
-each HP will likely be queried with $B$ different values, whereas for grid search each HP will only be queried with $B^{1/N}$ different values for $N$ HPs.
+Random search, on the other hand, randomly samples a value for each HP until it runs out of budget. Its explorative nature, ease of use, and no assumptions
+about the model makes it a good baseline. It is also superior to grid search in cases were one or more HPs have little impact on model performance,
+as for a given budget $B$, each HP will likely be queried with $B$ different values, whereas for grid search each HP will only be queried with $B^{1/N}$
+different values for $N$ HPs. \citep[chap. 1.3]{feurer_hyperparameter_2019}
 
-(evolutionary algorithms)
-- another model-free class of optimisation algorithms, inspired by naturally occuring evolution
-- based on a population, which iteratively (iteration = generation) generates $\lambda$ offspring via 3 operators (all sub-bullets from \citet[pp. 10-]{genetic_algos})
-  * \textbf{reproduction}: pick an individual to reproduce with prob proportional to its fitness
-  * \textbf{crossover}
-    - select 2 ``parents'' from pool of reproducing individuals
-    - for a position $k$ in their HP configs, with certain prob, swap their values after position $k$, yielding 2 new children
+% evolutionary algorithms
+Another popular class of model-free optimizers are evolutionary algorithms (EAs), which are conceptually simple and can handle even complex parameter spaces,
+given appropriate implementation of operators.
+EAs are iteratively evolving a population of individuals (an individual is simply a hyperparameter configuration) of size $\mu$, where in each iteration (generation),
+$\lambda$\footnote{We distinguish between $\lambda\in\mathbb{N}$ when referring to the offspring size of EAs and $\boldsymbol\lambda\in\boldsymbol\Lambda$ when
+referring to a hyperparameter configuration.} offspring are generated via three operators \citep[pp. 10-TODO]{genetic_algos}.\\
+\textbf{Reproduction} samples an individual from the population to reproduce with probability proportional to its fitnes. For HPO, fitness usually refers to the
+performance metric the model resulting from the individual's HP configuration is evaluated on.\\
+\textbf{Crossover} selects two ``parents'' from the pool of reproducing individuals. An index $k$ is sampled at random and all HP values from
+index $k$ on are swapped, yielding two new ``children'' individuals.\\
+\textbf{Mutation} randomly changes single HP values of an individual chosen for reproduction, e.g. by adding Gaussian noise to real values or flipping bits
+on binary values.\\
+At the end of each generation, the EA keeps the best $\mu$ individuals, either only from the offspring (``$(\mu,\lambda)$-selection'') or more commonly from
+population and offspring (``$(\mu+\lambda)$-selection''), which is also called an ``elitist'' strategy, as it's guaranteed to keep the best individual.
+\citep[chap. 1.3]{feurer_hyperparameter_2019}
+
+%- another model-free class of optimisation algorithms, inspired by naturally occuring evolution
+%- based on a population, which iteratively (iteration = generation) generates $\lambda$ offspring via 3 operators (all sub-bullets from \citet[pp. 10-]{genetic_algos})
+%  * \textbf{reproduction}: pick an individual to reproduce with prob proportional to its fitness
+%  * \textbf{crossover}
+%    - select 2 ``parents'' from pool of reproducing individuals
+%    - for a position $k$ in their HP configs, with certain prob, swap their values after position $k$, yielding 2 new children
     % TODO: use example from book page 12
-  * \textbf{mutation}: with certain prob, change values of an individual, e.g. by adding Gaussian noise (Gaussian mutation) to real values, or flipping bit for binary values
-- selection at end of each genertion: keep $\mu$ best individuals (based on some fitness function), either from only offspring (``$(\mu,\lambda)$-selection'')
-  or (more commonly) from population + offspring (``$(\mu+\lambda)$-selection'', guaranteed to keep best individual)
-- pro: conceptually simple + can handle even complex parameter spaces (continuous, discrete, hierarchical, etc.) given appropriate implementation of operators
-all above starting at HPO from \citep[chap. 1.3]{feurer_hyperparameter_2019}
+%  * \textbf{mutation}: with certain prob, change values of an individual, e.g. by adding Gaussian noise (Gaussian mutation) to real values, or flipping bit for binary values
+%- selection at end of each genertion: keep $\mu$ best individuals (based on some fitness function), either from only offspring (``$(\mu,\lambda)$-selection'')
+%  or (more commonly) from population + offspring (``$(\mu+\lambda)$-selection'', guaranteed to keep best individual)
+%- pro: conceptually simple + can handle even complex parameter spaces (continuous, discrete, hierarchical, etc.) given appropriate implementation of operators
 - prominent examples

### 2025-03-14 paper/main.tex
HPO model-free finished

@@
 i.e. $\argmin_{\boldsymbol\lambda\in\boldsymbol\Lambda} GE(\mathcal{I}_{\boldsymbol\lambda},\mathcal{D},L)$.
-For this there is no analytical expression available, making HPO a blackbox optimization problem. \citep[pp. 2f]{10.1145/3610536}
+For this, there is no analytical expression available, making HPO a blackbox optimization problem. \citep[pp. 2f]{10.1145/3610536}
 
@@
 $\lambda$\footnote{We distinguish between $\lambda\in\mathbb{N}$ when referring to the offspring size of EAs and $\boldsymbol\lambda\in\boldsymbol\Lambda$ when
-referring to a hyperparameter configuration.} offspring are generated via three operators \citep[pp. 10-TODO]{genetic_algos}.\\
+referring to a hyperparameter configuration.} offspring are generated via three operators \citep[pp. 10-14]{genetic_algos}.\\
 \textbf{Reproduction} samples an individual from the population to reproduce with probability proportional to its fitnes. For HPO, fitness usually refers to the
@@
 
-%- another model-free class of optimisation algorithms, inspired by naturally occuring evolution
-%- based on a population, which iteratively (iteration = generation) generates $\lambda$ offspring via 3 operators (all sub-bullets from \citet[pp. 10-]{genetic_algos})
-%  * \textbf{reproduction}: pick an individual to reproduce with prob proportional to its fitness
-%  * \textbf{crossover}
-%    - select 2 ``parents'' from pool of reproducing individuals
-%    - for a position $k$ in their HP configs, with certain prob, swap their values after position $k$, yielding 2 new children
-    % TODO: use example from book page 12
-%  * \textbf{mutation}: with certain prob, change values of an individual, e.g. by adding Gaussian noise (Gaussian mutation) to real values, or flipping bit for binary values
-%- selection at end of each genertion: keep $\mu$ best individuals (based on some fitness function), either from only offspring (``$(\mu,\lambda)$-selection'')
-%  or (more commonly) from population + offspring (``$(\mu+\lambda)$-selection'', guaranteed to keep best individual)
-%- pro: conceptually simple + can handle even complex parameter spaces (continuous, discrete, hierarchical, etc.) given appropriate implementation of operators
-- prominent examples
-  * CMA-ES
-    - ``Covariance Matrix Adapation Evolution Strategy''
-    - offspring generation exclusively via multivariate normal \citep[p. 8]{hansen2023cmaevolutionstrategytutorial}
-      * mean = weighted average of previous generation
-      * covariance = weighted covariance of previous generation, with weighting as for mean
-    - weighing scheme done to sample in a way as to reproduce previously successful (i.e. selected) steps \citep[p. 11]{hansen2023cmaevolutionstrategytutorial}
-  * differential evolution \citep[-]{differential_evolution}
-    - init population randomly so that entire param space is covered
-      * $D$-dim vectors $\boldsymbol{x}_{i,G}$ with $i=1,2,...,n$ for gen $G$
-    - mutation
-      * for each $\boldsymbol{x}_{i,G}$, generate \textbf{mutant vector} $\boldsymbol{v}_{i,G+1}=\boldsymbol{x}_{r_1,G}+F\cdot(\boldsymbol{x}_{r_2,G}-\boldsymbol{x}_{r_3,G})$
-      * $r_1,r_2,r_3\in\{1,2,...,n\}$ mutually different random idx + different from $i$
-      * constant $F\in[0,2]$
-    - crossover
-      * \textbf{trial vector} $\boldsymbol{u}_{i,G+1}=(u_{i,G+1}^{(1)},u_{i,G+1}^{(2)},...,u_{i,G+1}^{(D)})$
-        - chose random index $R\in\{1,2,...,D\}$
-        - sample $p\sim U[0,1]$
-        - define crossover constant $\text{CR}\in[0,1]$
-        - $u_{i,G+1} = \begin{dcases}
-          v_{i,G+1}^{(j)} & \text{if } p\le\text{CR}\text{ or } j=R \\
-          x_{i,G}^{(j)} & \text{else}
-         \end{dcases}$
-    - selection: greedy, compare fitness of $\boldsymbol{u}_{i,G+1}$ with $\boldsymbol{x}_{i,G}$, pick better
+One of the most popular EA implementations is the ``Covariance Matrix Adaptation Evolution Strategy'' (CMA-ES).
+It employs a multivariate normal distribution to generate offspring, for which the mean is a weighted average of the previous generation's individuals
+and the covariance matrix similarly is the weighted covariance of the previous generation.
+The weighing scheme for both is done in a way as to reproduce previously successful (i.e. selected) steps \citep[p. 8-11]{hansen2023cmaevolutionstrategytutorial}.
+
+Another popular EA approach is Differential Evolution. Its populaton is randomly initialised so the entire search space is covered.
+In each generation $g$, for each $\boldsymbol\lambda_{i,g}$ with $i=1,...,n$ it creates ``mutant vectors''
+$\boldsymbol\nu_{i,g+1}=\boldsymbol\lambda_{r_1,g}+F\cdot(\boldsymbol\lambda_{r_2,g}-\boldsymbol\lambda_{r_3,g})$,
+where $r_1,r_2,r_3\in\{1,...,n\}$ are mutually different random indices that also different from $i$ and $F\in[0,2]$ is constant.
+For crossover, it then generates a ``trial vector'' $\boldsymbol\upsilon_{i,g+1}=(\upsilon_{i,g+1}^{(1)},\upsilon_{i,g+1}^{(2)},...,\upsilon_{i,g+1}^{(N)})$ with
+\begin{equation}
+  \upsilon_{i,g+1}^{(j)} = \begin{dcases}
+    \nu_{i,g+1}^{(j)} & \text{if } p\le\text{CR}\text{ or } j=R \\
+    \lambda_{i,g}^{(j)} & \text{else}
+  \end{dcases}
+\end{equation}
+for a random index $R\in\{1,2,...,D\}$, sampled $p\sim U[0,1]$, and crossover constant $\text{CR}\in[0,1]$.
+Selection is done by picking the better of $\boldsymbol\upsilon_{i,g+1}$ and $\boldsymbol\lambda_{i,g}$ with respect to fitness. \citep[p. 343]{differential_evolution}
 

### 2025-03-14 paper/main.tex
remove dummy appendix

@@
 
-\section{}
-\label{app:theorem}
-
-% Note: in this sample, the section number is hard-coded in. Following
-% proper LaTeX conventions, it should properly be coded as a reference:
-
-%In this appendix we prove the following theorem from
-%Section~\ref{sec:textree-generalization}:
-
-In this appendix we prove the following theorem from
-Section~6.2:
-
-\noindent
-{\bf Theorem} {\it Let $u,v,w$ be discrete variables such that $v, w$ do
-not co-occur with $u$ (i.e., $u\neq0\;\Rightarrow \;v=w=0$ in a given
-dataset $\dataset$). Let $N_{v0},N_{w0}$ be the number of data points for
-which $v=0, w=0$ respectively, and let $I_{uv},I_{uw}$ be the
-respective empirical mutual information values based on the sample
-$\dataset$. Then
-\[
-	N_{v0} \;>\; N_{w0}\;\;\Rightarrow\;\;I_{uv} \;\leq\;I_{uw}
-\]
-with equality only if $u$ is identically 0.} \hfill\BlackBox
-
-\section{}
-
-\noindent
-{\bf Proof}. We use the notation:
-\[
-P_v(i) \;=\;\frac{N_v^i}{N},\;\;\;i \neq 0;\;\;\;
-P_{v0}\;\equiv\;P_v(0)\; = \;1 - \sum_{i\neq 0}P_v(i).
-\]
-These values represent the (empirical) probabilities of $v$
-taking value $i\neq 0$ and 0 respectively.  Entropies will be denoted
-by $H$. We aim to show that $\fracpartial{I_{uv}}{P_{v0}} < 0$....\\
-
-{\noindent \em Remainder omitted in this sample. See http://www.jmlr.org/papers/ for full paper.}
-
 \vskip 0.2in

### 2025-03-15 paper/main.tex
finished HPO model-based

@@
 \citet[chap. 2]{survey_NN_interpretability} extend this distinction to a three-dimensional taxonomy, allowing for better categorisation of neural networks (NNs),
-a model class that in its fully-connected feedforwad form is inherently non-interpretable.\\ % TODO: letztes halbsatz vllt lieber in intro unterbringen, hier ist der awkward und unnötig ausschmückend
+a model class that in its fully-connected feedforwad form is inherently non-interpretable. % TODO: letztes halbsatz vllt lieber in intro unterbringen, hier ist der awkward und unnötig ausschmückend
+\\
 \textbf{Passive vs Active Approaches}, where \textit{passive} are all post-hoc methods and \textit{active} methods actively change either the architecture or
-training process to increase model interpretability.\\
+training process to increase model interpretability.
+\\
 \textbf{Type of Explanations}, distinguishing between \textit{example} methods, providing concrete examples of what leads to a desired output, \textit{attribution}
 methods that attribute the effect on the output for a specific feature, \textit{hidden semantic} methods, which explain the types of inputs particular neurons or layers
-pick up on, and logical \textit{rules}, such as if-then clauses or tree-induced rules.\\
+pick up on, and logical \textit{rules}, such as if-then clauses or tree-induced rules.
+\\
 \textbf{Local vs Global Interpretability}, ranging from \textit{local} methods providing explanations based on individual samples, \textit{semi-local} methods,
@@
 Given its unclear definition, evaluating interpretability can be challenging. \citet[3]{DoshiVelez2017TowardsAR} propose a taxonomy to categorise
-possible evaluation methods based on their rigorousness.\\
+possible evaluation methods based on their rigorousness.
+\\
 \textbf{Application-grounded evaluation} evaluates interpretations directly with respect to the task, by having human experts evaluate the outcome and
-is therefore the most expensive and time-consuming of the three approaches.\\
+is therefore the most expensive and time-consuming of the three approaches.
+\\
 \textbf{Human-grounded metrics} is similar to application-grounded evaluation in that a human still evaluates the interpretations, but tries to simplify the task
 so that a layperson can do it. Human-grounded approaches are especially suitable if it's sufficient to validate the general concepts of a task. A typical evaluation
-set-up in this category is binary forced choice, where the human evaluator chooses, which of two generated explanations he prefers.\\
+set-up in this category is binary forced choice, where the human evaluator chooses, which of two generated explanations he prefers.
+\\
 \textbf{Functionally-grounded evaluation} is the least rigorous, but easiest to implement of the three. It assess explanatory quality according to some
@@
 $\lambda$\footnote{We distinguish between $\lambda\in\mathbb{N}$ when referring to the offspring size of EAs and $\boldsymbol\lambda\in\boldsymbol\Lambda$ when
-referring to a hyperparameter configuration.} offspring are generated via three operators \citep[pp. 10-14]{genetic_algos}.\\
+referring to a hyperparameter configuration.} offspring are generated via three operators \citep[pp. 10-14]{genetic_algos}.
+\\
 \textbf{Reproduction} samples an individual from the population to reproduce with probability proportional to its fitnes. For HPO, fitness usually refers to the
-performance metric the model resulting from the individual's HP configuration is evaluated on.\\
+performance metric the model resulting from the individual's HP configuration is evaluated on.
+\\
 \textbf{Crossover} selects two ``parents'' from the pool of reproducing individuals. An index $k$ is sampled at random and all HP values from
-index $k$ on are swapped, yielding two new ``children'' individuals.\\
+index $k$ on are swapped, yielding two new ``children'' individuals.
+\\
 \textbf{Mutation} randomly changes single HP values of an individual chosen for reproduction, e.g. by adding Gaussian noise to real values or flipping bits
-on binary values.\\
+on binary values.
+\\
 At the end of each generation, the EA keeps the best $\mu$ individuals, either only from the offspring (``$(\mu,\lambda)$-selection'') or more commonly from
@@
 where $r_1,r_2,r_3\in\{1,...,n\}$ are mutually different random indices that also different from $i$ and $F\in[0,2]$ is constant.
-For crossover, it then generates a ``trial vector'' $\boldsymbol\upsilon_{i,g+1}=(\upsilon_{i,g+1}^{(1)},\upsilon_{i,g+1}^{(2)},...,\upsilon_{i,g+1}^{(N)})$ with
+For crossover, it then generates a ``trial vector'' $\boldsymbol\upsilon_{i,g+1}=(\upsilon_{i,g+1}^{(1)},\upsilon_{i,g+1}^{(2)},...,\upsilon_{i,g+1}^{(N)})^T$ with
 \begin{equation}
   \upsilon_{i,g+1}^{(j)} = \begin{dcases}
-    \nu_{i,g+1}^{(j)} & \text{if } p\le\text{CR}\text{ or } j=R \\
+    \nu_{i,g+1}^{(j)} & \text{if } u\le\text{CR}\text{ or } j=R \\
     \lambda_{i,g}^{(j)} & \text{else}
@@
 \end{equation}
-for a random index $R\in\{1,2,...,D\}$, sampled $p\sim U[0,1]$, and crossover constant $\text{CR}\in[0,1]$.
+for a random index $R\in\{1,2,...,D\}$, sampled $u\sim U[0,1]$, and crossover constant $\text{CR}\in[0,1]$.
 Selection is done by picking the better of $\boldsymbol\upsilon_{i,g+1}$ and $\boldsymbol\lambda_{i,g}$ with respect to fitness. \citep[p. 343]{differential_evolution}
 
-[model based]
-- contrast to model free, fit a surrogate model on blackbox problem + optimise this
-(Bayesian optimisation)
-- iterative algo comprising of
-  * surrogate model for black box problem, needs to be able to model mean + variance
-  * acquisition function, to decide which point to query next
-- in each iteration
-  * fit surrogate model on all data points (posterior distribution)
-  * get highest utility point from acquisition function for newly fitted surrogate model -> add to data points
-  -> hence Bayesian, get new query point given already previously fitted points
-  * acquisition function trades off exploration + exploitation of surrogate
-  * BO up to here from \citep[chap. 1.3.2]{feurer_hyperparameter_2019} + \citep[pp. 2-3]{frazier2018tutorialbayesianoptimization}
-- no optimising of model directly, instead iteratively optimise acquisition function
-- common choices for surrogate models
-  * \textbf{gaussian processes}
-    - pros
-      * fully specified by mean + covariance
-        - covariance function (aka kernel) solely determines quality of GP
-        - kernel as function of two points from search space, yields their covariance
-        % TODO: write down kernel equation, cf. p 10ff https://ieeexplore.ieee.org/document/7352306
-        - usual property of kernels: the closer points are in search space, the more strongly their correlation \citep[p. 5]{frazier2018tutorialbayesianoptimization}
-      * well-calibrated uncertainty estimates
-      * closed-form computability
-    - con: neither scales well in \# of data points nor in \# of HP dimensions BUT workaround: sparse GPs, approx full GP with small subset of original dataset
-  * \textbf{random forests}
-    - can handle complex search spaces (high-dim, categorical, hierarchical), unlike GPs
-    - computational complexity scales far better than GPs
-      * GPs $O(n^3)$ fitting, $O(n^2)$ predicting
-      * RFs $O(n \log n)$ fitting, $O(\log n)$ predicting
-- common choices for acquisition functions
-  % TODO: cf. p. 13 cf. p 10ff https://ieeexplore.ieee.org/document/7352306
-  * \textbf{expected improvement}
-  * \textbf{Thompson sampling}
-- all BO stuff if not specified differently comes from \citep[chap. 1.3.2]{feurer_hyperparameter_2019}
+% model based
+Contrasting to model-free approaches, model-based methods fit a surrogate model on the target function and optimize the surrogate.
+Currently, one of the most popular model-based approaches is Bayesian Optimization (BO).
+BO is an iterative algorithm comprising a probabilistic surrogate model $\tilde{f}$ for the blackbox problem $f$ and an acquisition function.
+It maintains and continually extends a set of points it already evaluated on the target function.
+In each iteration, the posterior predictive distribution is determined by fitting the surrogate model on this set of points.
+The acquisition function then retrieves the highest utility point from the surrogate model and evaluates it on the target function, after which it adds the point
+to the set of queried points \citep[chap. 1.3.2]{feurer_hyperparameter_2019} and \citep[pp. 2f]{frazier2018tutorialbayesianoptimization}.
+Evidently, neither the target function nor the surrogate model are optimized directly. Rather, the acquisition function trades-off exploration and exploitation of
+the surrogate and is maximised to yield the next query point most likely to optimize the two objectives as defined by the acquisition function.
+\\
+Common choices for the acquisition function are Expected Improvement (Eq. \ref{eq-expected-improvement}) and Thompson Sampling (Eq. \ref{eq-thompson-sampling}).
+\begin{equation}
+  \text{EI}_t(\boldsymbol\lambda):=\mathbb{E}_t[(f(\boldsymbol\lambda)-f_t^*)^+]
+  \label{eq-expected-improvement}
+\end{equation}
+\textbf{Expected Improvement} computes the expected minimisation over the incumbent $f_t^*$ for evaluating the target function $f$ at query point $\boldsymbol\lambda$.
+$\mathbb{E}_t$ denotes the expectation under the posterior predictive distribution with $t$ query points, similarly, the current incumbent is the minimum value
+of the target function from previous evaluations $(\boldsymbol\lambda_{1:t},f(\boldsymbol\lambda_{1:t}))$. \citep[p. 7]{frazier2018tutorialbayesianoptimization}
+\begin{equation}
+  f^{(t)}\sim\tilde{f}_t
+  \label{eq-thompson-sampling}
+\end{equation}
+\textbf{Thompson Sampling}, on the other hand, can conceptually be described as drawing a candidate function $f^{(t)}$ from the posterior predictive distribution and
+minimising it to obtain the next query point for the target function. \citep[p. 161]{7352306}
+\\
+Composing the other part of BO, surrogate models need to be able to model mean and variance of its target function estimate.
+Commonly chosen models are Gaussian Processes (Eq. \ref{eq-gaussian-process}) and Random Forests.
+\begin{equation}
+  \text{GP}(m(\boldsymbol\lambda), k(\boldsymbol\lambda,\boldsymbol\lambda'))
+  \label{eq-gaussian-process}
+\end{equation}
+\textbf{Gaussian Processes} (GP) are fully specified by their mean and covariance functions and have closed-form solutions.
+The mean $m(\boldsymbol\lambda)$ fits all points already evaluated on the target function, while for the covariance or
+kernel function $k(\boldsymbol\lambda,\boldsymbol\lambda')$ it is desirable that points close to
+each other in HP space $\boldsymbol\Lambda$ exhibit stronger correlation than those far apart.
+The kernel function can be freely chosen, a common choices is the Gaussian (Eq. \ref{eq-kernel-gaussian}) kernel with
+hyperparameter $\alpha_0$. \citep[p. 5]{frazier2018tutorialbayesianoptimization}
+\begin{equation}
+  k(\boldsymbol\lambda,\boldsymbol\lambda')=\alpha_0 \exp(-\|\boldsymbol\lambda-\boldsymbol\lambda'\|_2^2)
+  \label{eq-kernel-gaussian}
+\end{equation}
+A drawback of Gaussian Processes is its poor scalability in number of data points and number of hyperparameters, although there exist workarounds approximating the full GP
+with only a subset of the samples. \citep[chap. 1.3.2]{feurer_hyperparameter_2019}
+\\
+\textbf{Random Forests}, unlike GPs, can handle complex hyperparameter spaces even including categorical and hierarchical HPs. Their computational complexity is also far
+less, making them a popular alternative to Gaussian Processes. \citep[chap. 1.3.2]{feurer_hyperparameter_2019}
 

### 2025-03-15 paper/main.tex
NAS

@@
 \subsection{Hyperparameter Optimization (HPO)}
-In contrast to model parameters, which are optimized during training, hyperparameters (HPs) are those describing the machine learning algorithm and are fixed before training.
-Because they usually have a significant impact on the trained model's performance, HPs are often optimized, too.
+In contrast to model parameters, which are optimized during training, hyperparameters (HPs) are those describing the machine learning algorithm and are
+fixed before training. Because they usually have a significant impact on the trained model's performance, HPs are often optimized, too.
 Let $\mathcal{D}\subseteq\mathcal{X}\times\mathcal{Y}$ be a dataset consisting of $n$
@@
 of the target function from previous evaluations $(\boldsymbol\lambda_{1:t},f(\boldsymbol\lambda_{1:t}))$. \citep[p. 7]{frazier2018tutorialbayesianoptimization}
-\begin{equation}
+\begin{equation}  % TODO: take out
   f^{(t)}\sim\tilde{f}_t
@@
 Commonly chosen models are Gaussian Processes (Eq. \ref{eq-gaussian-process}) and Random Forests.
-\begin{equation}
+\begin{equation}  % TODO: make inline or provide further information
   \text{GP}(m(\boldsymbol\lambda), k(\boldsymbol\lambda,\boldsymbol\lambda'))
@@
 
-\subsection{Neural Architecture Search (NAS)}
-- approaches specifically for NN HPO due to flexible structure of NNs and thus very large search space (each layer can have different \# nodes, different activations, for CNN different pooling, etc.
-operations)
-- we don't employ NAS for our extension as research focusses on the NLP and image domains, % TODO: find some more references
-  whose datasets exhibit strong correlation amongst features (i.e. tokens or pixels, respectively), an effect that is much weaker for tabular data \citep[p. 7499]{Borisov_2024}
-- at least want to give a little outline on most notable approaches
-  * \textbf{cell search space}
-    - modularise NN architecture into cells, NN as chain-structure of those
-    - cell = basic building block, fixed component of an NN
-      * e.g. linear layer with specific \# of neurons in feedforward
-      * or convolutional / pooling / etc layer for CNN
-    - then optimise sequential placement of the cells as HPO problem
-      * e.g. via random search or BO
-      * \citet[p. 3]{zoph2017neuralarchitecturesearchreinforcement} + \citet[pp. 2-4]{Zoph_2018_CVPR} use RNN to recursively optimise HPs of a cell given already determined
-        previous cells (+ HPs of current cell), which is trained via RL with the different HPs the RNN predicts being the actions and performance on held-out data being reward
-    - \citep[chap. 3.2]{elsken_neural_2019}
-  * \textbf{one-shot model}
-    - trains one overall ``fabric'' comprising all architectures of search space
-    - visualisation as DAG, nodes are layers, edges in-between are operations on a layer's values
-    - each path through DAG (from input to output node) represents one architecture of search space -> train entire DAG, then pick optimal path (architecture)
-    - pro: very efficient training, individual architectures share operations along edges they share, training one-shot model trains all subsumed models (more expensive
-      than training a single model but less expensive than trying out all configurations included in the fabric)
-    - \citep[pp. 1-2, p.8]{saxena2017convolutionalneuralfabrics}
+\subsection{Neural Architecture Search}
+A special case of HPO is Neural Architecture Search (NAS). Because of the flexibility of neural network architectures (number of hidden layers, strength of dropout,
+each layer can have a different numer of neurons, each layer can have a different activation function, etc.), traditional HPO methods cannot efficiently explore the
+entire space.
+In our extension of the EAGGA approach we don't employ NAS as research focusses on the NLP and image domains, % TODO: find some more references
+where features (i.e. tokens or pixels, respectively) exhibit strong correlation amongst themselves, which is usually not the case for tabular data \citep[p. 7499]{Borisov_2024}.
+Still, we want to outline two notable approaches for neural architecture search in this paper, one of which are \textbf{cell search spcaces}.
+These modularise an NN into a chain-structure of cells, where each cell is a basic building block of the respective architecture. For regular feedforwad neural networks,
+this could be a linear layer of fixed size and with a specific activation function,
+for CNNs this could be a specific convolutional operation or pooling function, for instance.
+Instead of tuning every hyperparameter, the sequential placement of the cells is then optimzed during NAS, e.g. via random search or Bayesian
+Optimization. \citep[chap. 3.2]{elsken_neural_2019}
+\\
+Treating a neural network as a sequence of cells instead of one fixed entity further allows exploiting this structure with special sequential optimization techniques.
+\citet[p. 3]{zoph2017neuralarchitecturesearchreinforcement} and \citet[pp. 2-4]{Zoph_2018_CVPR} propose a cell search method using RNNs and reinforcement learning.
+The cells in their approach are not fixed, but have hyperparameters themselves, such as the filter size and the stride size of a convolutional layer.
+These hyperparameters are then predicted sequentially, given all the already predicted hyperparameters of previous layers. The RNN is trained using reinforcement
+learning, where the actions are the different values the network can predict for a given hyperparameter and the reward function is the RNN's performance on held-out
+validation data.
+\\
+The second approach to NAS is the \textbf{one-shot model}, which trains a ``fabric'', which can be understood as a DAG comprising all architectures of a given
+search space. Each path through the DAG is one specific network architecture, where the nodes represent a layer of neurons and the edges in-between are operations
+on the layers' values. The entire DAG has one designated input and one output node.
+It is then trained just like a regular NN, after which the optimal path, i.e. network architecture, is selected from the fabric.
+This has the advantage of, albeit being slower than training a single model, being far more efficient and less expensive than training all architectures encoded
+in the fabric. \citep[pp. 1-2, p.8]{saxena2017convolutionalneuralfabrics}
 % don't necessarily talk about co-adaptation problem

### 2025-03-15 paper/main.tex
MOO pareto front + a priori

@@
 \subsection{Multi-Objective Optimization}
-- in most practical use cases one doesn't just want to optimise for performance but e.g. also for interpretability using some proxy metrics
-- formal definition
-  * vector of objectives $c_1,c_2,...,c_m=\boldsymbol{c}:\boldsymbol\Lambda\rightarrow\mathbb{R}^m$
-  * goal: minimise vector
-  * \citep[p. 11]{10.1145/3610536}
+In most applications, practicioners don't just want to optimize for performance exclusively but, for instance, also desire an interpretable model,
+which they may measure via some proxy metric.
+Let $c_1,c_2,...,c_m = \boldsymbol{c}:\boldsymbol\Lambda\rightarrow\mathbb{R}^m$ be the vector of $m$ objectives, the goal of multi-objective optimization (MOO)
+is to minimise this vector \citep[p. 11]{10.1145/3610536}.
+
 \subsubsection{Pareto-optimality}
-- problem: usually conflicting objectives, i.e. minimisation of $\boldsymbol{c}$ not possible along all dimensions
-- thus aim to find trade-off solutions of non-dominated points
-- we say a ``point dominates another''
-  * if there is no other point strictly better in at least one dimension and better or equal in the remaining ones
-  * formally: $\boldsymbol\lambda$ dominates $\boldsymbol\lambda'$ ($\boldsymbol\lambda\prec\boldsymbol\lambda'$) if and only if
-    $\forall i\in\{1,...,m\}:c_i(\boldsymbol\lambda) \le c_i(\boldsymbol\lambda') \wedge \exists j\in\{1,...,m\}:c_j(\boldsymbol\lambda) < c_j(\boldsymbol\lambda')$
-  * \citep[pp. 7f]{10.1145/3610536} + \citep[pp. 198f]{genetic_algos}
-- Pareto set: set of nondominated points $\mathcal{P}:=\{\boldsymbol\lambda\in\boldsymbol\Lambda|\nexists\boldsymbol\lambda'\in\boldsymbol\Lambda\text{ s.t.
-}\boldsymbol\lambda'\prec\boldsymbol\lambda\}$
-- Pareto front: image of nondominated points
-- goal: find set of nondominated points $\hat{\mathcal{P}}$ approximating true Pareto set $\mathcal{P}$ well
-(evaluation)
-- if knowledge over true Pareto-front, evaluating set of individuals can be done based on distance of estimated to true Pareto front
-- if no knowledge over true Pareto-front, volume-based approaches are popular, which measure volume between Pareto front estim and some chosen
-  reference point (usually worst point in objective space), e.g. 0 for accuracy (optimum is 1)
-- \citep[pp. 8-10]{10.1145/3610536}
+Optimizing $\boldsymbol{c}$ usually comes with the challenge of conflicting objectives: improving one objective means deteriorating another.
+We thus seek to find a set of trade-off solutions, so called \textit{non-dominated} points.
+A point $\boldsymbol\lambda$ dominates another point $\boldsymbol\lambda'$ ($\boldsymbol\lambda\prec\boldsymbol\lambda'$) if there is no other point that is
+strictly better in at least one objective and better or equal in the remaining ones. Formally,
+\begin{equation}
+  \forall i\in\{1,...,m\}:c_i(\boldsymbol\lambda) \le c_i(\boldsymbol\lambda') \wedge \exists j\in\{1,...,m\}:c_j(\boldsymbol\lambda) < c_j(\boldsymbol\lambda')
+  \label{eq-pareto-domination}
+\end{equation}
+as defined in \citep[pp. 7f]{10.1145/3610536} and \citep[pp. 198f]{genetic_algos}.
+\\
+The set of non-dominated or \textit{Pareto-optimal}
+points $\mathcal{P}:=\{\boldsymbol\lambda\in\boldsymbol\Lambda|\nexists\boldsymbol\lambda'\in\boldsymbol\Lambda\text{ s.t. }\boldsymbol\lambda'\prec\boldsymbol\lambda\}$
+is commonly referred to as the \textit{Pareto set} and its image as the \textit{Pareto front}.
+The goal of MOO is to find a set of non-dominated points $\hat{\mathcal{P}}$ approximating the true Pareto set $\mathcal{P}$ well.
+
+% evaluation
+There are different techniques to evaluate an estimated Pareto-front. If we have knowledge over the true Pareto-front, we can compute the distance between the approximation
+and the true front. In most cases, this knowledge cannot be assumed.
+Then, usually volume-based approaches are used for evaluation, most notably the hypervolume. The hypervolume, or S-metric, of a Pareto-front, computes the volume between
+the estimated front and some chosen reference point, which is usually the worst point in objective space.
+The larger the hypervolume, the closer the estimate is to the true Pareto-front. \citep[pp. 8-10]{10.1145/3610536}
+
 \subsubsection{A-priori}
-- requires specifying trade-off between objectives a-priori, will outline 2 popular approaches
-- e.g. different versions of \textbf{scalarization}
-  * weighted sum of objective functions
-    - $\argmin_{\boldsymbol\lambda\in\boldsymbol\Lambda} w_i c_i(\boldsymbol\lambda)$ with $\sum_{i=1}^k w_i=1$ and $w_i>0,\forall i=1,...,k$
-    - drawbacks
-      * solution sensitive to weights
-      * different users might have different opinions on weights \citep[chap. 3.1]{NSGA}
-    - \citep[p. 11]{10.1145/3610536}
-  * $\epsilon$-constraint
-    - translate all but one objective into constraint, then optimise remaining objective subject to the constraints
-    - $\text{w.l.o.g.} \argmin_{\boldsymbol\lambda\in\boldsymbol\Lambda} c_1(\boldsymbol\lambda) \text{ s.t. } c_2(\boldsymbol\lambda)\le\epsilon_2,...,c_m(\boldsymbol\lambda)\le\epsilon_m$
-    - conceputally similar to weighted sum: also sensitive to constraints, must be chosen sensibly
-    - \citep[p. 12]{10.1145/3610536}
-- \textbf{lexicographic method}
-  * define priority of objectives
-  * greedily optimise objectives in order of priority, constraint to the solutions of the already optimised (higher-priority) objectives
-  * again very dependent on user-defined priorisation
-  * \citep[p. 13749]{lexicographic_MOO}  % TODO: perhaps find a different source more related to ML
+For optimizing a multi-objective problem, there are largely two general approaches.
+One of those are \textit{a-priori} methods, of which we will briefly outline two popular techniques.
+
+\textbf{Scalarization} approaches impose an implicit order of priority on the objectives.
+The simplest one is optimizing a weighted sum of objective functions
+\begin{equation}
+  \argmin_{\boldsymbol\lambda\in\boldsymbol\Lambda} \sum_{i=1}^m w_i c_i(\boldsymbol\lambda)
+  \text{ s.t. } \sum_{i=1}^m w_i=1 \wedge w_i>0,\forall i=1,...,m
+  \label{eq-a-priori-scalarization-weighted-sum}
+\end{equation}
+Weights are chosen a-priori by the user and the solution is very sensitive to them, making this method very reliant on different
+users' preferences. \citep[p. 11]{10.1145/3610536} and \citep[chap. 3.1]{NSGA}
+\\
+Another form of scalarization is the $\epsilon$-constraint, which translates all but one objective into constraints and then optimizes the remaining
+objective subject to the constraints. Without loss of generality, the first constraint can be optimized
+\begin{equation}
+  \argmin_{\boldsymbol\lambda\in\boldsymbol\Lambda} c_1(\boldsymbol\lambda)
+  \text{ s.t. } c_2(\boldsymbol\lambda)\le\epsilon_2,...,c_m(\boldsymbol\lambda)\le\epsilon_m
+  \label{eq-a-priori-scalarization-epsilon}
+\end{equation}
+This method is, similarly to the weighted sum, conceptually simple yet also very sensitive to the chosen constraints. \citep[p. 12]{10.1145/3610536}
+\\
+Alternative to scalarization is the \textbf{lexicographic method}.
+For it, the user assigns a priority to each objective and then greedily optimizes each objective in order of priority, constrained to the solutions of
+the already optimized, higher-priority objectives. \citep[p. 13749]{lexicographic_MOO}
+Again, solutions are very dependent on the user-defined priorisation.
+
 \subsubsection{A-posteriori}

### 2025-03-15 paper/main.tex
finish MOO

@@
 \label{sec-moo-post}
-- problem a priori
-  * either restrict search space to ``enforce our will'' (e.g. only use 50\% of features), optimise only prediction performance + use this
-  * or leave search space unrestricted but adjust loss function to incorporate multiple objectives + take optimum from there
-  * in either case: no knowledge of interplay between HP config + performance on all objectives
-  * in practical applications, it is oftentimes useful to make the decision of which point of the Pareto set to use a-posteriori, e.g. if only a slight decrease in
-    one objective translates to a significant improvement in another that would have been missed if the problem was optimised using a-priori methods
-  * a-posteriori evaluates configs just as a-priori, but keeps track of multiple ``best'' (non-dominated) solutions -> makes relationship between HP config + objectives visible
-- aside from usual baselines grid and random seach there are multi-objective BO adaptations, mainly using either of two approaches
-  (1) fit single surrogate model on scalarised objectives, e.g. ParEGO, which employs the augemented Tchebycheff function as scalarisation to ensure the Pareto front is
-      explored sufficiently \citep[pp. 15f]{10.1145/3610536} + \citep[pp. 54-56]{ParEGO}
-  (2) fit one surrogate per objective, then
-    - either one acquisition function per objective -> return set of promising next candidates
-    - or one overall acquisition function aggregating surrogates, e.g. \textit{EHI}, maximises expected improvement of hypervolume \citep[p. 16]{10.1145/3610536} + \citep[pp. 8f]{EHI}
-- lastly, there is also multi-objective EA (MOEA) algorithms to explore Pareto front, one of which is NSGA-II (Nondominated Sorting Genetic Algorithm)
-  * improves upon predecessor NSGA \citep{NSGA} by making it parameterless, ensuring elitism, and reducing computational complexity of ranking of individuals \citep[p. 182]{NSGA_II}
-  * uses all the regular operators, i.e. reproduction, mutation, crossover can be used from single-objective EA (SOEA)
-  * difference to SOEA: ranking of individuals (SOEA uses scalar fitness, MOEA mutliple objectives)
-    - multi-objective ranking mechanism based on 2 parts
-      * non-dominated sorting, ensures elitism
-        - iterative procedure to rank individuals in population by their fronts
-        (1) determine pareto front, assign rank 0
-        (2) remove pareto front from population
-        (3) if any individuals left, repeat from 1, increment rank \#
-        - \citep[p. 201]{genetic_algos} + \citep[pp. 183f]{NSGA_II}
-      * crowding distance, ensures sufficient diversity, i.e. exploration of pareto front
-        - assigns score depending on how crowded area around individual in objective space is
-        - crowding distance of an individual = mean side length of cuboid spanned by its nearest neighbours as vertices in objective space
-        - individuals without two neighbours in a dimension are assigned infinitely high distance value
-        - \citep[p. 185]{NSGA_II}
-        -> the less the crowding distance, the more ``crowded'' an area is by other individuals
-      -> rank individuals by nds front rank (ascending) + crowding distance (descending) as tie breaker
-  * ranking used for selecting $\mu$ best individuals to keep for next generation + for reproduction: via binary tournament selection: select two random individuals,
-    pick best w.r.t. nds + cd for offspring creation (i.e. mutation / crossover)
-  * using nds for ranking / fitness makes intuitive sense, but why cd?
-    - goal of MOO: not simply optimise HV, but approx true Pareto front well
-    - only having individuals representing one area in objective space makes Pareto front estimate very ``unstable'': removing this area (i.e. solutions form that area)
-      would ``collapse'' entire front
-    - instead having multiple areas be represented makes the estimate ``stable'': removing one area wouldn't impact Pareto front estimate a lot
-    - \citep[p. 185, pp. 189-192]{genetic_algos}
+The disadvantage of a-priori methods is the missing knowledge of the interplay between a hyperparameter configuration and the trained model's
+performance across objectives:
+we can either restrict the search space to enforce some will, such as only using a maximum of 50\% of features, optimize for performance and use the
+resulting model, or leave the search space unrestricted, adjust the loss function to incorporate all objectives and take its optimum.
+In neither case do we know the impact our a-priori trade-off has on the final performance.
+In practical applications, the impact on performance across all objectives is what matters.
+This is the main advantage of a-posteriori methods: instead of implicitly defining a trade-off prior to training, we still evaluate multiple HP configurations
+but keep the set of non-dominated solutions instead of just one that satisfies the trade-off criterion.
+This is beneficial as it makes the relationship visible, a practicioner can now see, for instance, that a slight decrease in one objective might translate to
+a significant improvement in another that would have been missed if the problem was optimized with a-priori methods.
+\\
+Aside from the usual baselines grid and random search there are multi-objetive Bayesian Optimization adaptations, mainly using one of two approaches:
+\begin{enumerate}
+  \item Fitting a single surrogate model on scalarized objectives. \citet[pp. 54-56]{ParEGO} propose \textit{ParEGO}, which employs the augemented Tchebycheff function as
+        scalarization to ensure the Pareto front is explored sufficiently.
+  \item Fitting one surrogate per objective, then either using one acquisition function per surrogate to return a set of promising candidate query points, or using
+        one overall acquisition function to aggregate the surrogates, such as \textit{EHI} as proposed by \citet[pp. 8f]{EHI}, which maximises the expected improvement
+        of hypervolume.
+\end{enumerate}
+
+Lastly, there is also a multitude of multi-objective evolutionary algorithms (MOEA) to explore the Pareto front.
+A key challenge for MOEAs is determining the fitness of an individual, as there are now multiple objectives that need to be incorporated instead of
+one as in single-objective EA (SOEA).
+We will outline the popular NSGA-II (Nondominated Sorting Genetic Algorithm II), which improves upon its predecessor NSGA \citep{NSGA} by making it
+parameterless, ensuring elitism, and reducing the computational complexity of ranking of individuals \citep[p. 182]{NSGA_II}.
+NSGA-II uses all the regular operators reproduction, mutation, and crossover that are already known from SOEA.
+The difference is in the fitness function, which for NSGA-II comprises two parts: non-dominated sorting and crowding distance.
+\\
+\textit{Non-dominated sorting}, as described by \citet[p. 201]{genetic_algos} and \citet[pp. 183f]{NSGA_II} is an iterative procedure to ensure
+elitist selection through ranking individuals by their fronts.
+First, it assigns all individuals of the Pareto front a rank of 0. It then temporarily removes these individuals from the population and repeats
+this procedure with an incremented rank number until no individuals are left.
+\\
+The second part, \textit{crowding distance} ensures sufficient diversity, i.e. exploration of the Pareto front.
+The algorithm assigns each individual a score depending on how ``crowded'' the area in objective space around it is. Crowding distance is computed
+as the mean side length of the cuboid spanned by its nearest neighbours as vertices in the objective space, where individuals without two neighbours
+along some dimension are assigned an infinitely high distance value. The less the crowding distance, the more an area is by other individuals. \citep[p. 185]{NSGA_II}
+\\
+NSGA-II then ranks individuals by their non-dominated sorting front ranks in ascending order and by their crowding distance in descending order as tie breaker.
+This ranking is then used for selecting the $\mu$ best individuals to keep for the next generation.
+Reproduction follows a binary tournament selection, where two individuals are randomly drawn from the population and the better one according to the ranking gets
+included in the reproduction pool for mutation and crossover.
+\\
+Conceptually, the reasoning behind using non-dominated sorting to rank individuals is intuitive: the closer an individuals front is to the Pareto front, i.e.
+the smaller its front rank, the better it approaches the true Pareto front.
+The usage of crowding distance, on the other hand, is not as straightforward. To understand it, we recall the goal of MOO, which is not merely optimizing
+hypervolume, but to approximate the true Pareto front well.
+Considering the scenario of only having individuals of one area in the objective space included in the Pareto front estimate makes it very unstable: removing this
+area (i.e. the solutions from this area) would collapse the entire front estimate.
+Considering now the opposite, having multiple areas making up the Pareto front estimate, makes it stable: removing one area would not impact the estimate a lot,
+as it's still held up by other individuals. \citep[p. 185, pp. 189-192]{genetic_algos}
 
@@
       * $\boldsymbol{o}_{\text{out}}^{(l)}=ReLU(\boldsymbol{o}_{\text{in}}^{(l)})=\max(0,\boldsymbol{o}_{\text{in}}^{(l)})$ monotonic
+  (experimental setup)
 \begin{figure}
@@
       * so far only happened for madeline in after gen-8.json after 5hrs 45mins (at start of gen-9, which wasn't exported yet) -> loaded this and ran for another 2hrs 15mins
+  * only 18 datasets
 

### 2025-03-15 paper/main.tex
EAGGA

@@
 \usepackage{bm}
+
+\usepackage{enumitem}  % nested enumerations, cf. https://tex.stackexchange.com/a/126756
 
@@
 Aside from the usual baselines grid and random search there are multi-objetive Bayesian Optimization adaptations, mainly using one of two approaches:
-\begin{enumerate}
+\begin{enumerate}[label*=\arabic*.]
   \item Fitting a single surrogate model on scalarized objectives. \citet[pp. 54-56]{ParEGO} propose \textit{ParEGO}, which employs the augemented Tchebycheff function as
         scalarization to ensure the Pareto front is explored sufficiently.
-  \item Fitting one surrogate per objective, then either using one acquisition function per surrogate to return a set of promising candidate query points, or using
-        one overall acquisition function to aggregate the surrogates, such as \textit{EHI} as proposed by \citet[pp. 8f]{EHI}, which maximises the expected improvement
-        of hypervolume.
+  \item Fitting one surrogate per objective, then either 
+  \begin{enumerate}[label*=\arabic*.]
+    \item using one acquisition function per surrogate to return a set of promising candidate query points or
+    \item using one overall acquisition function to aggregate the surrogates, such as \textit{EHI} as proposed by \citet[pp. 8f]{EHI}, which maximises
+          the expected improvement of hypervolume.
+  \end{enumerate}
 \end{enumerate}
@@
 \section{EAGGA}
-- NSGA-II inspired evolutionary + genetic algorithm
-- uses AUC-ROC as performance metric and 3 interpretation metrics
-  * NF: rel. \# of features used in model
-  * NI: rel. \# of pairwise feature interaction effects
-  * NNM: rel. \# of non-monotone feature effects
-- aims to find high performing models (AUC) with low NF, NI, and NNM -> ensures better interpretability than simply optimising for performance
-  but creates very large objective space $\check{\boldsymbol\Lambda}$ to be optimised over with
+EAGGA, as proposed by \citet{EAGGA}, is an NSGA-II inspired algorithm that optimizes performance along with three interpretability objectives.
+The three interpretability proxies are
+\begin{itemize}
+  \item NF: the relative number of features used in the model
+  \item NI: the relative number of pairwise interaction effects in the model
+  \item NNM: the relative number of non-monotone feature effects in the model
+\end{itemize}
+The authors argue that a sparse model as indicated by a low NF score helps interpretability by reducing the number of e.g. ALE plots necessary in post-hoc
+analysis \citep[p. 540]{EAGGA}. Similarly, a low NI score reduces the interplay between features and thus the dimensionality of post-hoc analysis plots
+and a low NNM avoids complex non-monotone feature effects, which is often desirable in practical use-cases such as loan approval, where a higher income should
+translate to a higher chance of approval.
+Lastly, performance is measured by the area under the receiver operating characteristic curve (AUC), which trades-off the true against the false positives rate.
+\\
+\citet[pp. 540f]{EAGGA} acknowledge that this set-up creates the large objective space $\check{\boldsymbol\Lambda}$ to be optimized over, containing
 tuples $(\boldsymbol\lambda, \boldsymbol{s}, \boldsymbol{I}_s, \boldsymbol{m}_{\boldsymbol{I}_{\boldsymbol{s}}})=\check{\boldsymbol\lambda}\in\check{\boldsymbol\Lambda}$
 with
-  * $\boldsymbol\lambda\in\boldsymbol\Lambda$ ``regular'' model HP config
-  * $\boldsymbol{s}\in\{0,1\}^p$ vector denoting feature usage
-  * $\boldsymbol{I}_s\in\{0,1\}^{p\times p}$ matrix denoting pairwise interactions
-  * $\boldsymbol{m}_{\boldsymbol{I}_{\boldsymbol{s}}}\in\{-1,0,1\}^p$ vector denoting monotonicity constraint of each feature (-1 decreasing, 0 none, 1 increasing)
-- thus would require to solve $\argmin_{\check{\boldsymbol\lambda}\in\check{\boldsymbol\Lambda}} (GE(\mathcal{I}_{\check{\boldsymbol\lambda}},\mathcal{D}),NF(\hat{f}_{\mathcal{D},\check{\boldsymbol\lambda}}),NI(\hat{f}_{\mathcal{D},\check{\boldsymbol\lambda}}),NNM(\hat{f}_{\mathcal{D},\check{\boldsymbol\lambda}}))$
-- \citep[pp. 540f]{EAGGA}
-- introduces equivalence relation $R$ ``allowed to interact'' on features that significantly reduces search space for more efficient optimization
-- thus: augmented search space $\tilde{\boldsymbol\Lambda}=\boldsymbol\Lambda\times\mathcal{G}$ comprising model HP space and group structure space
-  * each group structure $\boldsymbol{G}\in\mathcal{G}$ consists of $g$ groups:
-    - $G_1$ set of excluded features
-    - $\forall k=2,...,g:G_k=(E_k,M_{E_k})$ tuple
-      * $E_k$ set of features allowed to interact with each other
-      * $M_{E_k}$ monotonicity constraint of entire group $k$
-- EA on model hyperparameters $\boldsymbol\Lambda$
-- GGA on group structure space $\mathcal{G}$ with adapted mutation and crossover operators
-- \citep[pp. 541f]{EAGGA}
-- futher introduces special initialisation of group structures to increase sample-efficiency of EAGGA algorithm compared to random init
-- \citep[pp. 542f]{EAGGA}
+\begin{itemize}
+  \item $\boldsymbol\lambda\in\boldsymbol\Lambda$ being the model's hyperparameter configuration
+  \item $\boldsymbol{s}\in\{0,1\}^p$ being a binary vector denoting which features are used in the model
+  \item $\boldsymbol{I}_s\in\{0,1\}^{p\times p}$ being a binary matrix denoting included pairwise interactions between features
+  \item $\boldsymbol{m}_{\boldsymbol{I}_{\boldsymbol{s}}}\in\{-1,0,1\}^p$ being the vector denoting the monotonicity constraint of a feature (-1 decreasing, 0 none, 1 increasing)
+\end{itemize}
+Consequently, the optimization problem over this space would be
+\begin{equation}
+  \argmin_{\check{\boldsymbol\lambda}\in\check{\boldsymbol\Lambda}} \big(
+    GE(\mathcal{I}_{\check{\boldsymbol\lambda}},\mathcal{D}),
+    NF(\hat{f}_{\mathcal{D},\check{\boldsymbol\lambda}}),
+    NI(\hat{f}_{\mathcal{D},\check{\boldsymbol\lambda}}),
+    NNM(\hat{f}_{\mathcal{D},\check{\boldsymbol\lambda}})
+  \big)
+  \label{eq-eagga-extended-space}
+\end{equation}
+The authors thus introduce the equivalence relation $R$ ``allowed to interact'' on the set of features $C$. They further denote all features included in the model with $C_s$.
+This induces a group structure of equivalence classes, where features allowed to interact belong to the same class and share the same monotonicity attribute.
+Using this group structure, they define a much simpler augmented serach space $\tilde{\boldsymbol\Lambda}=\boldsymbol\Lambda\times\mathcal{G}$ comprising of
+the cartesian product of the model hyperparameter space $\boldsymbol\Lambda$ and the group structure space $\mathcal{G}$.
+Each group structure $\boldsymbol{G}\in\mathcal{G}$ consists of $g$ groups:
+\begin{itemize}
+  \item $G_1=C\setminus C_s$ being the set of excluded features
+  \item $G_k=(E_k,M_{E_k}), \forall k=2,...,g$ being tuples with
+  \begin{itemize}
+    \item $E_k\subseteq C_s$ being the set of features in group $k$, i.e. features allowed to interact with each other
+    \item $M_{E_k}$ being the monotonicity constraint of group $k$
+  \end{itemize}
+\end{itemize}
+To generate new offspring, they use a regular evolutionary algorithm on the model hyperparameters $\boldsymbol\Lambda$ and a grouping genetic algorithm (GGA) on
+the group structure space $\mathcal{G}$, with adapted mutation and crossover operators.
+The authors further introduce a special initialisation of the group structures to increase the sample-efficiency of their algorithm compared to random initialisation.
+\citep[pp. 541-543]{EAGGA}
 

### 2025-03-15 paper/main.tex
own extension, overall algorithm

@@
 \\
-\citet[pp. 540f]{EAGGA} acknowledge that this set-up creates the large objective space $\check{\boldsymbol\Lambda}$ to be optimized over, containing
+\citet[pp. 540f]{EAGGA} acknowledge that this set-up creates the complex ``extended search space'' $\check{\boldsymbol\Lambda}$ to be optimized over, containing
 tuples $(\boldsymbol\lambda, \boldsymbol{s}, \boldsymbol{I}_s, \boldsymbol{m}_{\boldsymbol{I}_{\boldsymbol{s}}})=\check{\boldsymbol\lambda}\in\check{\boldsymbol\Lambda}$
@@
 This induces a group structure of equivalence classes, where features allowed to interact belong to the same class and share the same monotonicity attribute.
-Using this group structure, they define a much simpler augmented serach space $\tilde{\boldsymbol\Lambda}=\boldsymbol\Lambda\times\mathcal{G}$ comprising of
+Using this group structure, they define a much simpler ``augmented serach space'' $\tilde{\boldsymbol\Lambda}=\boldsymbol\Lambda\times\mathcal{G}$ comprising of
 the cartesian product of the model hyperparameter space $\boldsymbol\Lambda$ and the group structure space $\mathcal{G}$.
@@
 
-\section{Extension to neural networks}
-- original EAGGA applied to XGBoost model proved to vastly outperform union of competitor models w.r.t. dominated hypervolume and comparable or better than ParEGO on
-  extended search space $\check{\boldsymbol\Lambda}$ \citep[pp. 543-545]{EAGGA}
-- why extend it to NNs?
-  * NNs notoriously uninterpretable due to complex transformation of the feature space
-  * due to (lin alg) non-linear activation functions, interactions can be modelled
-  * no monotonicity guarantees
-- our approach
-  (general algorithm)
-  * EAGGA algorithm implemented largely as described in the paper
-  * HP init
-    - NN total layers $\in\{3, ..., 10\}$, i.e. hidden layers $\in\{1, ..., 8\}$, init from trunc geom with prob=0.5
-    - NN nodes per hidden layer (for each group) $\in\{3, ..., 20\}$, init from trunc geom with prob=0.5
-    - NN dropout \% $\in[0, 1]$, init from trunc gamma with shape=2, scale=0.15
-  * group structure init
-    - feature detector
-      * as described in paper
-      * but instead of fitting 10 trees + taking rel. \# of features used as prob for trunc geom, we simply use 0.5 (sklearn dectree examination not straightforward)
-      * also in preliminary experiments we found the sampled \# of features to be used occassionally to be > \# of non-0 values in normalised information gain filter
-        (e.g. if former is = total \# of features and one feature is indep. from target) in these edge cases we simply use all features with non-0 filter values
-    - interaction detector
-      * as described in paper
-      * same as for feat detect: use prob=0.5 to sample \# of interactions used instead of 10 dec trees
-      * also for FAST algorithm don't use RSS (lin reg metric) but mean accuracy, as we fit log reg
-    - monotonicity detector
-      * as described in paper
-      * use default HPs (max depth 30, minsplit = 20) of mlr3 classification tree implementation, as this is the library the paper uses
-    -> both interaction + monotonicity detectors fit their models on 80\% of train split from holdout + eval on remaining 20\%
-  (implementation details)
+\section{Extension to Neural Networks}
+EAGGA as implemented by \citet{EAGGA} applies the optimizer to XGBoost, which proved to vastly outperform a union of competitor models with respect to
+dominated hypervolume and achieved comparable or better performance than ParEGO optimizing over the
+extended search space $\check{\boldsymbol\Lambda}$ \citep[pp. 543-545]{EAGGA}.
+Our work extends EAGGA to neural networks. We chose this model class for several reasons:
+\begin{enumerate}[label*=\arabic*.]
+  \item NNs are notoriously uninterpretable due to complex transformations of the feature spaces
+  \item Non-linear activation functions implicitly model interactions without direct control of which features interact with each other
+  \item There are no monotonicty guarantees
+\end{enumerate}
+
+\subsection{Overall Algorithm}
+We implement EAGGA largely as described by \citet{EAGGA} with only minor changes due to no straightforward transferability either due to different implementations
+of some classes and functions in R compared to Python or differences in model properties between XGBoost and neural nets.\\
+Our NNs use the following hyperparameters initialised as described
+\begin{itemize}
+  \item Total number of layers $\in\{3, ..., 10\}$, initialised from a truncated Geometric distribution with $\pi=0.5$.
+        This includes input and output layers, the number of hidden layers is thus $\in\{1, ..., 8\}$.
+  \item Number of nodes per hidden layer $\in\{3, ..., 20\}$, initialisation from a truncated Geometric distribution with $\pi=0.5$.
+  \item Dropout probability $\in[0, 1]$, initialised from a truncated Gamma distribution with shape 2 and scale 0.15.
+\end{itemize}
+For the group structure initialization we also employ a feature, interaction, and monotonicity detector.
+\\
+The feature detector is as described by \citet[p. 542]{EAGGA}, but instead of fitting ten trees and using the relative number of features used as
+probability for the truncated Geometric distribution, we use a constant value of 0.5, as the sklearn DecisionTree examination is not straightforward.
+Further, in preliminary experiments we found the sampled number of features to occassionally be larger than the number of non-zero values in the
+normalised information gain filter. This can happen if the former is equal to the total number of features and one or more features are independent form the
+target. In these edge cases we use all features with non-zero filter-values instead.
+\\
+The interaction detector similarly uses a constant probability of 0.5 to sample the number of interactions to be used instead of the relative number of interactions
+from ten decision trees, for the same reasoning as for the feature detector. Also, for the FAST algorithm computing the interaction scores we don't use the residual
+sum of squares but the mean accuracy, as the datasets all have binary targets and FAST thus fits a logistic instead of a linear regression.
+\\
+The monotonicty detector is implemented as described by \citet[p. 543]{EAGGA}, using the default decision tree hyperparameters from mlr3 in Python.
+\\
+Both the models in the interaction and the monotonicity detector (FAST and an ensemble of decision trees, respectively) use 80\% of the data for model estimation and
+the remaining 20\% to compute their scores.
+
+\subsection{Architectural Details}
   * group structures + datasets
@@
     - NN
-      * comprises of "sub-NNs", one for each equivalence relation -> basically non-fully connected NN
+      * comprises "sub-NNs", one for each equivalence relation -> basically non-fully connected NN
       * hidden layers use ReLU activation and dropout afterwards, then a shared output layer with concatenated sub-NNs' outputs as input and sigmoid activation
@@
 - preliminary
-  * our implementation of feature + interaction detectors likely tend to include
+  * our implementation of feature + interaction detectors likely tend to include  % belongs in "general algorithm"
     - more features + interactions for high p datasets
@@
     - this is not by mistake, usually AUC on binary target can simply be inverted (simply predict the opposite class)
-    - for us this cannot easily be inverted due to monotonicity constraints (weights cannot simply be multiplied by -1 if constraint)
+    - for us this cannot easily be inverted due to monotonicity constraints (weights cannot simply be multiplied by -1 if constrained)
     - could happen e.g. due to AdamW and weight clipping (employed to enforce monotonicity)
@@
   * dhv contributions
+    % TODO: put explanation in line 400, don't introduce new topic here
     - measures contribution of an individual to the hypervolume, i.e. difference between hypervolume of entire pareto front vs hypervolume of pareto front without
@@
 
-\newpage
+%\newpage  % we are not to add page breaks
 

### 2025-03-15 paper/main.tex
appendix after references
own extension architectural details

@@
 \subsection{Architectural Details}
-  * group structures + datasets
-    - group structures implemented as described in the paper with additional list encoding sign of monotonicity of the individual features as detected by the
-      monotonicity detector
-    - included features are passed to dataset, dataset outputs included features, multiplied by the features' individual monotonicity (-1 / 1), so that group monotonicity
-      can be encoded with only $\{0,1\}$, as proposed in paper
-    - entire group structure is passed to NN, where architecture is built accordingly
+We implement the group structures as described by \citet[p. 541]{EAGGA} with an additional list encoding whether the signs of individual features should
+be inverted (-1) or not (1), as detected by the monotonicity detector.
+The features included in the group structure are then passed to a custom pytorch Dataset implementation, which outputs only the included features, multiplied with their
+individual sign. This allows the groups' monotonicity constraints to be encoded with only $\{0,1\}$, as proposed by \citet[p. 543]{EAGGA}.
+The entire group structure is then passed to the neural network, where the architecture is built accordingly.
 \begin{figure}
@@
 \end{figure}
-  * neural network
-    - "instead of XGB as in the original paper, we apply the method to NNs to examine whether this type of regularisation can achieve interpretability on NNs while
-      outperforming EAGGA on XGB (original paper), ..., this requires special architecture, etc. pp."
-    - NN
-      * comprises "sub-NNs", one for each equivalence relation -> basically non-fully connected NN
-      * hidden layers use ReLU activation and dropout afterwards, then a shared output layer with concatenated sub-NNs' outputs as input and sigmoid activation
-      * optimizer is AdamW with default params, loss is binary cross entropy loss (sigmoid thus implicit in the loss function for better numerical stability,
-        NN itself outputs "logits", which refers to pre-activation output in pytorch)
-    - feature sparsity thus achieved by only training on included feature groups -> goes somewhat against notion of deep learning where model is supposed to decide itself,
-      which feature to "use" / put importance on -> ELABORATE
-    - feature interaction achieved by grouping, max-operation in ReLU induces interaction effect (different kind of interaction than e.g. in LM with multiplication)
-      -> equation why the interacting features need to be grouped together when using ReLU, cf. photos
-      * given layer input $\boldsymbol{x}$ and w.l.o.g. omitting bias
-      * $\text{ReLU}(\boldsymbol{x})=\max(0,\sum_{j=1}^p w_p x_p)$
-      * interaction not of the form $x_j \cdot x_{j'}$ as used to from LMs but of a form where the sum $w_j x_j+w_{j'} x_{j'}$ decides whether $x_j,x_{j'}$
-        pass through the activation or not
-      * in other words: $\max(0,w_j x_j + w_{j'} x_{j'})\neq\max(0,w_j x_j)+\max(w_{j'} x_{j'})$ and thus there is an interaction
-    - monotonicty constraint achieved by clipping weights to [0, infty) after each epoch for restricted equivalence relations (monotonic decrease achieved
-      via dataset object multiplying features with their individual signs), bias clipping not necessary (constant additive term)
-      -> equation how monotonicity is enforced with this
-      * consider vector form computation of hidden neuron $\boldsymbol{o}_{\text{in}}^{(l)}\in\mathbb{R}^{p^{(l)}}$ of layer $l$
-        - $\boldsymbol{o}_{\text{in}}^{(l)}=\boldsymbol{W} \boldsymbol{o}_{\text{out}}^{(l-1)} + \boldsymbol{b}$
-        - where $\boldsymbol{o}_{\text{in}}^{(l)}$ refers to pre-activation values, $\boldsymbol{o}_{\text{out}}^{(l)}$ to value after applying ReLU
-        - weight matrix $\boldsymbol{W}\in\mathbb{R}^{p^{(l)} \times p^{(l-1)}}$
-        - bias $\boldsymbol{b}\in\mathbb{R}^{p^{(l)}}$
-      * definition of monotonically increasing function $f$: for $x_1\le x_2$ it must hold that $f(x_1)\le f(x_2)$
-      * $\boldsymbol{W} \boldsymbol{o}_{\text{out}}^{(l-1)} + \boldsymbol{b}$ clearly only monotonic if and only if $\boldsymbol{W}\in\mathbb{R}_{0+}{p^{(l-1)}\times p^{(l)}}$
-        - where $\mathbb{R}_{0+}=[0,\infty)$
-        - hence weight clipping necessary, bias clipping not because only constant additive term
-      * $\boldsymbol{o}_{\text{out}}^{(l)}=ReLU(\boldsymbol{o}_{\text{in}}^{(l)})=\max(0,\boldsymbol{o}_{\text{in}}^{(l)})$ monotonic
-  (experimental setup)
+Applying EAGGA to NNs requires defining a special architecture to realise interaction and monotonicity constraints.
+The neural network consists of a group of fully-connected sub-networks that are not connected amongst each other, as visualised in Figure \ref{fig-nn-architecture}.
+Each equivalence class in the group structure has its own sub-network, so that it is guaranteed that only these features can interact with each other.
+\\
+For the hidden layers, we use ReLU activations with dropout afterwards, after which a shared output layer transforms the concatenated output of the sub-networks
+to a probability using the Sigmoid activation function.
+We use AdamW with default parameters, optimizing a binary cross-entropy loss with Sigmoid implicit in the loss function for better numerical stability. The network
+itself thus outputs only a ``logit'', which is the pre-activation output in pytorch.
+\\
+Feature sparsity is achieved by only training on the included features that are passed to the network from the dataset.
+\\
+Grouping the network into sub-networks enables restricting feature interactions to only those features included in the respective group.
+The $\max$-operation in the ReLU induces a certain interaction effect on its inputs. It is important to note that this is a different kind of interaction
+than the multiplicative one happening for instance in a linear model.
+Given a layer's input $\boldsymbol{x}$ and without loss of generality omitting the bias, the ReLU activation is defined as follows
+\begin{equation}
+  \text{ReLU}(\boldsymbol{x})=\max(0,\sum_{j=1}^p w_j x_j)
+  \label{eq-our-extension-interaction-1}
+\end{equation}
+We see that the interaction is not of the form $x_j \cdot x_{j'}$ but rather of a form where the sum $w_j x_j+w_{j'} x_{j'}$ decides whether $x_j,x_{j'}$
+pass through the activation or not.
+In other words, the interaction is given by the fact that
+\begin{equation}
+  \max(0,w_j x_j + w_{j'} x_{j'})\neq\max(0,w_j x_j)+\max(w_{j'} x_{j'})
+  \label{eq-our-extension-interaction-2}
+\end{equation}
+Finally, monotonicty is achieved by clipping the respective sub-network's weights to $[0,\infty)$ after each epoch. Whether a feature's effect is
+increasing or decreasing depends on the individual feature's sign encoded in the custom dataset implementation.
+Consider the vector form computation of a pre-activation hidden neuron $\boldsymbol{o}_{\text{in}}^{(l)}\in\mathbb{R}^{p^{(l)}}$ of layer $l$
+\begin{equation}
+  \boldsymbol{o}_{\text{in}}^{(l)}=\boldsymbol{W} \boldsymbol{o}_{\text{out}}^{(l-1)} + \boldsymbol{b}
+  \label{eq-our-extension-monotonicity}
+\end{equation}
+where $\boldsymbol{o}_{\text{out}}^{(l)}$ denotes the value after applying ReLU, $\boldsymbol{W}\in\mathbb{R}^{p^{(l)} \times p^{(l-1)}}$ refers to
+the weight matrix and $\boldsymbol{b}\in\mathbb{R}^{p^{(l)}}$ to the bias.
+A function $f$ is monotonically increasing if and only if $\forall x_1\le x_2$ it holds that $f(x_1)\le f(x_2)$ and thus
+$\boldsymbol{W} \boldsymbol{o}_{\text{out}}^{(l-1)} + \boldsymbol{b}$ is clearly only monotonically increasing if and only if
+$\boldsymbol{W}\in\mathbb{R}_{0+}^{p^{(l-1)}\times p^{(l)}}$ where $\mathbb{R}_{0+}=[0,\infty)$.
+$\boldsymbol{o}_{\text{out}}^{(l)}=ReLU(\boldsymbol{o}_{\text{in}}^{(l)})=\max(0,\boldsymbol{o}_{\text{in}}^{(l)})$ Is therefore a monotonically increasing
+function if and only if the weights are clipped as described. Bias clipping is not necessary as it is a constant additive term.
+
+\subsection{Experimental Setup}
 \begin{figure}
@@
 
+\vskip 0.2in
+\bibliography{references}
+
 \appendix
@@
 
-\vskip 0.2in
-\bibliography{references}
-
 \end{document}

### 2025-03-15 paper/main.tex
finished own extension

@@
 of some classes and functions in R compared to Python or differences in model properties between XGBoost and neural nets.\\
+% TODO: after architectural details:
 Our NNs use the following hyperparameters initialised as described
@@
 \end{figure}
+\\
 Applying EAGGA to NNs requires defining a special architecture to realise interaction and monotonicity constraints.
 The neural network consists of a group of fully-connected sub-networks that are not connected amongst each other, as visualised in Figure \ref{fig-nn-architecture}.
-Each equivalence class in the group structure has its own sub-network, so that it is guaranteed that only these features can interact with each other.
+Each equivalence class in the group structure has its own sub-network, so that it is guaranteed that only the features within that group can interact with each other.
 \\
@@
 The $\max$-operation in the ReLU induces a certain interaction effect on its inputs. It is important to note that this is a different kind of interaction
-than the multiplicative one happening for instance in a linear model.
+than the multiplicative one happening, for instance, in a linear model.
 Given a layer's input $\boldsymbol{x}$ and without loss of generality omitting the bias, the ReLU activation is defined as follows
@@
 \end{equation}
-where $\boldsymbol{o}_{\text{out}}^{(l)}$ denotes the value after applying ReLU, $\boldsymbol{W}\in\mathbb{R}^{p^{(l)} \times p^{(l-1)}}$ refers to
+where $\boldsymbol{o}_{\text{out}}^{(l)}$ denotes the value of layer $l$ after applying ReLU, $\boldsymbol{W}\in\mathbb{R}^{p^{(l)} \times p^{(l-1)}}$ refers to
 the weight matrix and $\boldsymbol{b}\in\mathbb{R}^{p^{(l)}}$ to the bias.
@@
 $\boldsymbol{W} \boldsymbol{o}_{\text{out}}^{(l-1)} + \boldsymbol{b}$ is clearly only monotonically increasing if and only if
-$\boldsymbol{W}\in\mathbb{R}_{0+}^{p^{(l-1)}\times p^{(l)}}$ where $\mathbb{R}_{0+}=[0,\infty)$.
-$\boldsymbol{o}_{\text{out}}^{(l)}=ReLU(\boldsymbol{o}_{\text{in}}^{(l)})=\max(0,\boldsymbol{o}_{\text{in}}^{(l)})$ Is therefore a monotonically increasing
+$\boldsymbol{W}\in\mathbb{R}_{0+}^{p^{(l)}\times p^{(l-1)}}$, where $\mathbb{R}_{0+}=[0,\infty)$.
+Therefore, $\boldsymbol{o}_{\text{out}}^{(l)}=\text{ReLU}(\boldsymbol{o}_{\text{in}}^{(l)})=\max(0,\boldsymbol{o}_{\text{in}}^{(l)})$ is a monotonically increasing
 function if and only if the weights are clipped as described. Bias clipping is not necessary as it is a constant additive term.
@@
 \end{figure}
-  * evaluation, holdout, cv, early stopping
-    - evaluation via dominated hypervolume along AUC-ROC, NF, NI, NNM as defined in original paper
-      * NF simply rel. \# of included features
-      * NI = sum of all possible pairwise interactions in each group over all possible pairwise interactions among all features = $\frac{\sum_g^G {p_g \choose 2}}{{p \choose 2}}$
-      * NNM = rel. \# unconstrained features
-    - outer holdout split: 2/3 train, 1/3 test (as in paper)
-      * run EAGGA on holdout train split, i.e. train + select best $\mu=100$ individuals based on non-dom-sorting (ascending) + crowding distance (descending) as tie breaker
-    - inner CV split on outer train portion: 5-fold (as in paper)
-      * in each fold fit model on 80\% of CV-train portion, use remaining 20\% of CV-train for early stopping
-    - early stopping criterion
-      * for each fold, always train for min 200 epochs, keep track of model with lowest loss
-      * after that, use patience of 100: if current model's loss is > mean of last 100 epochs' losses -> stop early
-      * if no early stopping, train each fold for max 10 minutes
-      * then stop training, return model with lowest loss
-    -> refere Figure \ref{fig-dataset-split}
-    - then evaluate on last remaining from CV + keep best $\mu$ individuals + generate $\lambda=10$ offspring for new generation + repeat
-    - final evaluation of individuals of pareto set
-      * for each individual, train model on training set of hold-out for max \# of epochs taken during CV-training
-      * decided on max \# epochs instead of mean after looking at loss graphs in preliminary experiments, also refer Figure \ref{fig-es-losses}
-      * those show that there is no uptick in losses on the early stopping portion (disjunct from training portion) of the set, hence max is reasonable
-  * hardware: training on Sagemaker Notebook instance
-    - initial experiments on ml.g4dn.xlarge instance (2 vCPUs, 16GiB RAM, 1 NVIDIA T4, cf. https://aws.amazon.com/de/ec2/instance-types/g4/)
-      using cuda not much faster than on ml.t3.medium (2 Intel Xeon 8000 vCPUs, 4GiB RAM, cf. https://aws.amazon.com/de/ec2/instance-types/t3/)
-    - thus decided for more economic + ressourcen-schonend t3.medium
-  * did not train on philippine and gina datasets (308, 970 features, respectively), as they ran out of memory when computing interaction detectors
-  * known bugs
-    - in rare cases (anecdotally once every ~5-10 datasets), gga\_mutate seems to be generating -1 as monotonicity attribute, despite np.random.randint(low=0, high=2, size=1)
-      * loop crashes at group\_structure creation, for these cases subtract previous runtime from 8hrs + load from last generation in output via load\_population
-      * so far only happened for madeline in after gen-8.json after 5hrs 45mins (at start of gen-9, which wasn't exported yet) -> loaded this and ran for another 2hrs 15mins
-  * only 18 datasets
+We evaluate our models just like \citet[pp. 543f]{EAGGA} by computing the dominated hypervolume along AUC, NF, NI, NNM with reference point $(0, 1, 1, 1)^T$.
+Internally, this is implemented as minimising $1-\text{AUC}$ and a reference point $(1, 1, 1, 1)^T$.
+For NF, we simply count the relative number of included features.
+NI sums the numbers of all possible pairwise interactions in each group $\sum_g^G {p_g \choose 2}$ and divides them by the number of all possible pairwise interactions
+on the entire dataset ${p \choose 2}$.
+NNM counts the relative number of features in unconstrained equivalence classes.
+\\
+Our dataset is split as visualised in Figure \ref{fig-dataset-split}.
+We apply an outer holdout train-test split with 1/3 of the dataset being held for final testing, the 2/3 are used for 5-fold cross-validated training of each individual,
+as described by \citet[p. 543]{EAGGA}.
+What is different for our implementation is that on each training portion of the CV split, we further apply a holdout split, where 80\% are used for model training and
+20\% are reserved for early stopping.
+\\
+We use a conservative early stopping stratetgy, where each fold is always trained for a minimum of 200 epochs, keeping track of the fold's model with the lowest loss.
+After 200 epochs, our criterion with a patience of 100 epochs is applied: now, if the current epoch's model loss is greater than the mean of the last 100 epochs'
+losses, training in this fold stops early and returns the model with the lowest loss over the entire fold's training period.
+If no early stopping is triggered, we train each fold for a maximum of 10 minutes and return the model with the lowest loss on the early stopping portion.
+\\
+The model is then evaluated on the remaining portion from the CV split and after each generation, we keep the best $\mu=100$ individuals and generate $\lambda=10$
+offspring for the next generation and repeat this for 8 hours.
+\\
+The final evaluation of the individuals from the cross-validated Pareto set is then conducted on the held 1/3 of the dataset. Each individual
+is trained on the 2/3 training portion for the maximum number of epochs the individual required in the cross-validated training.
+Oftentimes, for cross-validation with early stopping, the final model is only trained on the mean or median number of epochs from CV.
+We decided specifically on the maximum after examining the loss over epochs in preliminary experiments, as plotted in Figure \ref{fig-es-losses}
+for the final experimental setup. Those show that there is no uptick in losses on the early stopping portion indicating that training too long does not increase
+the validation error.
+\\
+We trained in Sagemaker Notebooks on AWS and did initial training runs on
+ml.g4dn.xlarge\footnote{2 vCPUs, 16GiB RAM, 1 NVIDIA T4, cf. https://aws.amazon.com/de/ec2/instance-types/g4/} instances with CUDA.
+There was, however, no noticeable speed-up in model training compared to the much more economical
+ml.t3.medium\footnote{2 Intel Xeon 8000 vCPUs, 4GiB RAM, cf. https://aws.amazon.com/de/ec2/instance-types/t3/} instances, after which we decided to run the
+final experimental setup on those.
+\\
+In total, we run our models on the same datasets as \citet[p. 544]{EAGGA}, except for the 2 largest ones ``philippine'' and ``gina'' (308 and 970 features, respectively),
+as those ran out of memory when initializing the interaction effects.
+\\
+In rare cases (anecdotally once every ~10 datasets), we observed our monotonicity detector generating a group monotonicty attribute of -1, which crashed the computation.
+For these cases, we implemented a function to import a previous generation of individuals to be able to pick up from where the training crashed and not train anew.
+This happened once for the dataset ``madeline'' after generation 9 and training for 5 hours 45 minutes, and hence we imported the generation and ran it for the remaining
+2 hours 15 minutes.
 

### 2025-03-15 paper/main.tex
put Overall Algorithm after Architectural Details

@@
 \begin{enumerate}[label*=\arabic*.]
-  \item NNs are notoriously uninterpretable due to complex transformations of the feature spaces
+  \item NNs are notoriously uninterpretable due to complex transformations of the feature space
   \item Non-linear activation functions implicitly model interactions without direct control of which features interact with each other
-  \item There are no monotonicty guarantees
+  \item There are no monotonicty guarantees in neural networks
 \end{enumerate}
-
-\subsection{Overall Algorithm}
-We implement EAGGA largely as described by \citet{EAGGA} with only minor changes due to no straightforward transferability either due to different implementations
-of some classes and functions in R compared to Python or differences in model properties between XGBoost and neural nets.\\
-% TODO: after architectural details:
-Our NNs use the following hyperparameters initialised as described
-\begin{itemize}
-  \item Total number of layers $\in\{3, ..., 10\}$, initialised from a truncated Geometric distribution with $\pi=0.5$.
-        This includes input and output layers, the number of hidden layers is thus $\in\{1, ..., 8\}$.
-  \item Number of nodes per hidden layer $\in\{3, ..., 20\}$, initialisation from a truncated Geometric distribution with $\pi=0.5$.
-  \item Dropout probability $\in[0, 1]$, initialised from a truncated Gamma distribution with shape 2 and scale 0.15.
-\end{itemize}
-For the group structure initialization we also employ a feature, interaction, and monotonicity detector.
-\\
-The feature detector is as described by \citet[p. 542]{EAGGA}, but instead of fitting ten trees and using the relative number of features used as
-probability for the truncated Geometric distribution, we use a constant value of 0.5, as the sklearn DecisionTree examination is not straightforward.
-Further, in preliminary experiments we found the sampled number of features to occassionally be larger than the number of non-zero values in the
-normalised information gain filter. This can happen if the former is equal to the total number of features and one or more features are independent form the
-target. In these edge cases we use all features with non-zero filter-values instead.
-\\
-The interaction detector similarly uses a constant probability of 0.5 to sample the number of interactions to be used instead of the relative number of interactions
-from ten decision trees, for the same reasoning as for the feature detector. Also, for the FAST algorithm computing the interaction scores we don't use the residual
-sum of squares but the mean accuracy, as the datasets all have binary targets and FAST thus fits a logistic instead of a linear regression.
-\\
-The monotonicty detector is implemented as described by \citet[p. 543]{EAGGA}, using the default decision tree hyperparameters from mlr3 in Python.
-\\
-Both the models in the interaction and the monotonicity detector (FAST and an ensemble of decision trees, respectively) use 80\% of the data for model estimation and
-the remaining 20\% to compute their scores.
+The optimizer is implemented largely as described by \citet{EAGGA} with only minor changes due to there being no straightforward transferability of some functions
+from R to Python or differences in model properties between XGBoost and neural nets.\\
 
@@
 \\
-Grouping the network into sub-networks enables restricting feature interactions to only those features included in the respective group.
+Grouping the network into sub-networks enables restricting feature interactions to only those features that are included in the respective group.
 The $\max$-operation in the ReLU induces a certain interaction effect on its inputs. It is important to note that this is a different kind of interaction
-than the multiplicative one happening, for instance, in a linear model.
+than the multiplicative one known, for instance, from the linear model.
 Given a layer's input $\boldsymbol{x}$ and without loss of generality omitting the bias, the ReLU activation is defined as follows
@@
 
+\subsection{Overall Algorithm}
+Our NNs use the following hyperparameters initialised as described
+\begin{itemize}
+  \item Total number of layers $\in\{3, ..., 10\}$, initialised from a truncated Geometric distribution with $\pi=0.5$.
+        This includes input and output layers, the number of hidden layers is thus $\in\{1, ..., 8\}$
+  \item Number of nodes per hidden layer $\in\{3, ..., 20\}$, initialisation from a truncated Geometric distribution with $\pi=0.5$
+  \item Dropout probability $\in[0, 1]$, initialised from a truncated Gamma distribution with shape 2 and scale 0.15
+\end{itemize}
+For the group structure initialization we also employ feature, interaction, and monotonicity detectors.
+\\
+The feature detector is as described by \citet[p. 542]{EAGGA}, but instead of fitting ten trees and using the relative number of features as
+probability for the truncated Geometric distribution, we use a constant value of 0.5, as the sklearn DecisionTree examination is not straightforward.
+Further, in preliminary experiments we found the sampled number of features to occassionally be larger than the number of non-zero values in the
+normalised information gain filter. This can happen if the former is equal to the total number of features and one or more features are independent form the
+target. In these edge cases we use all features with non-zero filter-values instead.
+\\
+The interaction detector similarly uses a constant probability of 0.5 to sample the number of interactions to be used instead of the relative number of interactions
+from ten decision trees, for the same reasoning as for the feature detector. Also, for the FAST algorithm, when computing the interaction scores we don't
+use the residual sum of squares but the mean accuracy, as the datasets all have binary targets and FAST thus fits a logistic instead of a linear regression.
+\\
+The monotonicty detector is implemented as described by \citet[p. 543]{EAGGA}, using the default decision tree hyperparameters from mlr3 in Python.
+\\
+Both the models in the interaction and the monotonicity detector (FAST and an ensemble of decision trees, respectively) use 80\% of the data for model estimation and
+the remaining 20\% to compute their scores.
+
 \subsection{Experimental Setup}
@@
 For NF, we simply count the relative number of included features.
-NI sums the numbers of all possible pairwise interactions in each group $\sum_g^G {p_g \choose 2}$ and divides them by the number of all possible pairwise interactions
+NI sums the numbers of all possible pairwise interactions in each group ${p_k \choose 2}$ and divides them by the number of all possible pairwise interactions
 on the entire dataset ${p \choose 2}$.

### 2025-03-15 paper/main.tex
put two figures next to each other

@@
 We trained in Sagemaker Notebooks on AWS and did initial training runs on
-ml.g4dn.xlarge\footnote{2 vCPUs, 16GiB RAM, 1 NVIDIA T4, cf. https://aws.amazon.com/de/ec2/instance-types/g4/} instances with CUDA.
+ml.g4dn.xlarge\footnote{2 vCPUs, 16GiB RAM, 1 NVIDIA T4, cf. \url{https://aws.amazon.com/ec2/instance-types/g4/}} instances with CUDA.
 There was, however, no noticeable speed-up in model training compared to the much more economical
-ml.t3.medium\footnote{2 Intel Xeon 8000 vCPUs, 4GiB RAM, cf. https://aws.amazon.com/de/ec2/instance-types/t3/} instances, after which we decided to run the
+ml.t3.medium\footnote{2 Intel Xeon 8000 vCPUs, 4GiB RAM, cf. \url{https://aws.amazon.com/ec2/instance-types/t3/}} instances, after which we decided to run the
 final experimental setup on those.
@@
 \section{Experimental Results and Discussion}
+
+% have two figures next to each other
 \begin{figure}
   \centering
-  \includegraphics[width=0.9\linewidth]{../code/export/plot_test_dhvs.png}
-  \caption{TODO: caption}
+  % make figure wider than textwidth: https://tex.stackexchange.com/a/16584
+  \makebox[\textwidth][c]{\begin{minipage}{0.6\linewidth}
+      \centering
+      \includegraphics[width=\linewidth]{../code/export/plot_test_dhvs.png}
+      \caption{Dominated Hypervolume of NNs trained using XGBoost.}
       \label{fig-test-dhvs}
+  \end{minipage}\hfill
+  \begin{minipage}{0.6\linewidth}
+      \centering
+      \includegraphics[width=\linewidth]{./EAGGA-XGB-dhv.PNG}
+      \caption{Mean dominated Hypervolume of $EAGGA_{\text{XGBoost}}$, $EAGGA_{\text{XGBoost}_{\text{md2}}}$, and union of competitors averaged over
+      10 replications. Bars represent standard errors. \citep[Figure 2, p. 544]{EAGGA}}
+      \label{fig-eagga-xgb-dhvs}
+  \end{minipage}}
 \end{figure}
+
 - preliminary

### 2025-03-15 paper/main.tex
start results + discussion

@@
 as it's still held up by other individuals. \citep[p. 185, pp. 189-192]{genetic_algos}
+A possible proxy for exploration of the Pareto front is the hypervolume contribution. The hypervolume contribution of individual
+$\boldsymbol\lambda\in\hat{\mathcal{P}}$ is the difference between the hypervolume of the entire estimated Pareto front vs the hypervolume of the front without the
+individual: $\text{CON}_{\hat{\mathcal{P}}}(\boldsymbol\lambda)=\text{HYP}(\hat{\mathcal{P}})-\text{HYP}(\hat{\mathcal{P}}\setminus\{\boldsymbol\lambda\})$,
+where $\text{HYP}(\hat{\mathcal{P}})$ denotes the hypervolume induced by the Pareto set $\hat{\mathcal{P}}$ \citep[p. 384]{10.5555/1943267.1943271}.
 
@@
 \end{figure}
+As evident from Figures \ref{fig-test-dhvs}, \ref{fig-eagga-xgb-dhvs}, applying EAGGA to neural networks yields comparable performance to applying it to XGBoost,
+both with and without restricting the maximum tree depth.
+A direct comparison to unrestricted neural networks trained with regular stochastic gradient descent has not been made, as this would have led to a hypervolume of 0
+due to NF, NI, and NNM all being 1.
+Similar to $EAGGA_{\text{XGBoost}}$ and $EAGGA_{\text{XGBoost}_{\text{md2}}}$, our extension consistently outperforms the union of competitors.
+Unfortunately, due to time constraints, we were not able to compare the performance against a union of competitors including an unrestricted neural network
+instead of XGBoost, although we hypothesise that, given unrestricted NNs' struggle on tabular data compared to XGBoost, the outperformance over the union would
+be even greater.
+\\
+Examining the Pareto sets returned from training, which are accessible on our GitHub repository at ~/code/export/*.csv, we can observe a clear trend in the
+Pareto sets' models' hyperparameters:
+\begin{itemize}
+  \item the total number of layers is mostly 3 or 4, but goes as high as 7 for the Satellite (36) and diabetes (8) datasets
+  \item the nodes per hidden layer are mostly in the range between 3 and 6, but go as high as 12 on the diabetes (8) dataset
+  \item the dropout probability goes as high as 0.7 on the climate (18) and spambase (57) dataset, but is mostly in the 0.1 to 0.3 range
+\end{itemize}
+The returned group structures are more diverse:
+NF goes up to 1 for individuals in the Pareto sets of phoneme (5) and blood (4). Further datasets with low feature cardinality exhibt similarly high
+NF scores, with bankote (4) up to 0.75, climate (18) up to 0.67, and diabetes (8) up to 0.5. For the remainder, NF is much lower. The heightened NF for
+datasets with less features can be a possible consequence of shorter training times: on average, low feature cardinality datasets train quicker and thus
+for more generations than high cardinality datasets. This in turn means that their group structure space is far more likely to be exhaustively searched,
+with then most of the performance improvements coming from changes to the model's hyperparameters.
+As a consequence of the consistently low NF (except for low cardinality datasets), NI and NNM are equally very low, as they are directly dependent on the
+number of features included in the model.
+\\
+The contributions to the dominated hypervolumes are predominately low for the models in the Pareto set, often being below 0.1, with the most contribution
+often coming from the featureless learner. This is a sign for good and stable exploration of the Pareto front, as outlined in Section \ref{sec-moo-post}
+\\
+The loss graphs in Figure \ref{fig-es-losses} suggest that some models could have benefitted from longer training on some datasets, as some losses have not
+coverged when the early stopping criterion was triggered.
+This indicates that the criterion was likely triggered by a short-term spike, which could potentially be resolved by not comparing the current epoch's loss
+to the mean of the previous 100 epochs but, for instance, the mean of a range of more recent epochs to the last 100.
+This strategy, on the other hand, would then likely yield to even longer training periods on those datasets that clearly have converged, as evident by
+horizontal loss graphs.
+As the loss graphs remain horizontal and there is no uptick again, this would likely not be detrimental to model performance and the main ``risk'' would be
+more a expensive computation from longer trainings.
+\\
+The dominated hypervolume over generations in Figure \ref{fig-val-dhvs}
+
+- dhv over generations Figure \ref{fig-val-dhvs}
+  * compute on val set, i.e. had 5 folds per individual
+    - NF, NI, NNM always the same for each fold
+    - but 5 different AUCs
+    -> hypervolume of (mean(AUC 1, AUC 2, ..., AUC 5), NF, NI, NNM)
+  * artifacts / drops along y likely due to inconsistent computation of Pareto front by third-party library
+    - preliminary experiments on dummy pareto fronts with 10x4 metrics: noticed nds function returning different rankings for same front
+    - returned ranking switched back and forth between two only slightly different options (only 1 or 2 indices were swapped)
+    - not sure what caused this, as made same observation using another library that was planned as alternative
+    - thus unfortunately not fixable for me
+  * all but 3 datasets (2 of which only trained for 1 generation, aynway) show improvement of dominated hypervolume over generations
+  * BUT: absolute as well as relative improvement almost negligible -> no large final dhv decrease would we just have evaluated the models gotten from the detectors
+    - this was also the reason we didn't run EAGGA on philippine and gina simply without the detectors (as that's where they crashed) -> evidence suggested that detectors are vital to initial performance
+    - original paper supports this assumption \citep[Fig. 4, p. 545]{EAGGA}
 
@@
       * weight clipping very imperfect way of enforcing monotonicity, but currently in pytorch unfortunately only way to implement this
-- Figure \ref{fig-test-dhvs}
-  * suggests comparable performance to XGBoost EAGGA
-  * did not compare to unrestricted NN, would not be sensible, as NF, NI, NNM would be 1, hence hypervolume = 0, as values along 3 dimensions would be at reference point
-  * as XGBoost EAGGA, NN EAGGA consistently outperforms union of competitors as evident from \citep[Figure 2]{EAGGA}
-  * unfortunately due to time constraints no own union comparison with NN instead of XGBoost
-- overview over pareto sets can be accessed on our github repo at ~/code/export/*.csv
-  * NN HPs
-    - total layers mostly 3-4, most commonly 3, goes as high as 7 (Satellite (36), diabetes (8))
-    - nodes per hidden layer mostly 3-6, goes as high as 12 (diabetes (8))
-    - p dropout goes as high as 0.7 (climate (18), spambase (57)), but mostly in the 0.1 to 0.3 range
-  * group structures
-    - great diversity in NF across datasets
-      * phoneme (5) up to 1
-      * blood (4) up to 1
-      * banknote (4) up to 0.75
-      * diabetes (8) up to 0.5
-      * climate (18) up to 0.67
-      -> low p datasets in the benchmark tend to have higher NF
-        - possible consequence of shorter evaluation time
-        - low p datasets have much higher \# generations
-        - group structure space much more likely to be exhausted, i.e. more exploration of NN hps
-    - NI, NNM consequently (bounded by NF, can never be more than max \# for respective \# included features) rather low
-  * dhv contributions
-    % TODO: put explanation in line 400, don't introduce new topic here
-    - measures contribution of an individual to the hypervolume, i.e. difference between hypervolume of entire pareto front vs hypervolume of pareto front without
-      individual $\boldsymbol\lambda$: $\text{CON}_{\mathcal{P}}(\boldsymbol\lambda)=\text{HYP}(\mathcal{P})-\text{HYP}(\mathcal{P}\setminus\{\boldsymbol\lambda\})$,
-      where $\text{HYP}(\mathcal{P})$ denotes the hypervolume induced by the pareto set $\mathcal{P}$ \citep[p. 384]{10.5555/1943267.1943271}
-    - predominately low for fitted models
-    - mostly highest for featureless learner predicting majority class
-    -> sign of good exploration of pareto front + stable estimate, refer Section \ref{sec-moo-post}
-- loss graphs Figure \ref{fig-es-losses}
-  * suggests models could have benefitted from longer training on some datasets, as some loss curves haven't converged when stopping criterion hit
-  * crit was likely triggered by short-term spike, thus could potentially be resolved by comparing average of last k losses against average of >patience< losses prior
-    to that to not be as exposed to short-term spikes in loss
-  * on other datasets, graphs suggest earlier stop would have been totally fine as the networks have long converged, but longer training likely not an issue as loss
-    graphs come from early stopping dataset portion, which is disjunct from training set
-- dhv over generations Figure \ref{fig-val-dhvs}
-  * compute on val set, i.e. had 5 folds per individual
-    - NF, NI, NNM always the same for each fold
-    - but 5 different AUCs
-    -> hypervolume of (mean(AUC 1, AUC 2, ..., AUC 5), NF, NI, NNM)
-  * artifacts / drops along y likely due to inconsistent computation of Pareto front by third-party library
-    - preliminary experiments on dummy pareto fronts with 10x4 metrics: noticed nds function returning different rankings for same front
-    - returned ranking switched back and forth between two only slightly different options (only 1 or 2 indices were swapped)
-    - not sure what caused this, as made same observation using another library that was planned as alternative
-    - thus unfortunately not fixable for me
-  * all but 3 datasets (2 of which only trained for 1 generation, aynway) show improvement of dominated hypervolume over generations
-  * BUT: absolute as well as relative improvement almost negligible -> no large final dhv decrease would we just have evaluated the models gotten from the detectors
-    - this was also the reason we didn't run EAGGA on philippine and gina simply without the detectors (as that's where they crashed) -> evidence suggested that detectors are vital to initial performance
-    - original paper supports this assumption \citep[Fig. 4, p. 545]{EAGGA}
 

### 2025-03-15 paper/main.tex
experimental results + discussion

@@
 \\
-The dominated hypervolume over generations in Figure \ref{fig-val-dhvs}
+The AUC used in computing the dominated hypervolume over generations in Figure \ref{fig-val-dhvs} is the mean AUC over all folds on the validation set.
+All but 3 datasets, 2 of which were only trained for 1 generation, exhibit improvement of their dominated hypervolume over generations. The absolute as
+well as relative improvement, however, is almost negligible, suggesting there to be no large decrease in dominated hypervolume, would we have just evaluated
+the group structure initialised by the detectors.
+This was also the reason behind not running EAGGA on the philipine and gina datasets on randomly initialised detectors (as the detectors for those datasets
+ran out of memory due to the high feature cardinality), as the detectors prove vital for the initial performance. This finding is also in
+line with \citep[Fig. 4, p. 545]{EAGGA}.
+\\
+Clearly, there are some artifacts causing drops along the y-axis in most plots. This is likely due to the inconsistent computation of the Pareto front
+by a third-party library: in preliminary experiments we observed slightly varying front ranks on a synthetic sample of 10 (AUC, NF, NI, NNM) vectors.
+Unfortunately, this behaviour was still present when switching to another library for front rank computation, and thus not fixable for us.
+From our anecdotal evidence, the variations in rankings in these preliminary experiments only concerned few samples that were assigned to alternating
+neighbouring fronts, which seemed reasonable not to be concerned over.  % ok formulation?
+\\
+Another observation from the logs during training is AUC occassionally being less than 0.5.
+In general, AUC can then simply be inverted, as the model can usually always simply predict the opposite.
+In our case, however, this is not possible as easily due to weight clipping for groups with a monotonicty constraint.
+A potential scenario were an AUC less than 0.5 could happen is when AdamW has high momentum, leading to negative weights that are then clipped to the
+non-negative domain at the end of the epoch.
+We acknowledge that weight clipping is a very imperfect way of enforcing monotonicity, but currently seems to be the only way to implement this type of
+constrained optimisation in pytorch.
 
-- dhv over generations Figure \ref{fig-val-dhvs}
-  * compute on val set, i.e. had 5 folds per individual
-    - NF, NI, NNM always the same for each fold
-    - but 5 different AUCs
-    -> hypervolume of (mean(AUC 1, AUC 2, ..., AUC 5), NF, NI, NNM)
-  * artifacts / drops along y likely due to inconsistent computation of Pareto front by third-party library
-    - preliminary experiments on dummy pareto fronts with 10x4 metrics: noticed nds function returning different rankings for same front
-    - returned ranking switched back and forth between two only slightly different options (only 1 or 2 indices were swapped)
-    - not sure what caused this, as made same observation using another library that was planned as alternative
-    - thus unfortunately not fixable for me
-  * all but 3 datasets (2 of which only trained for 1 generation, aynway) show improvement of dominated hypervolume over generations
-  * BUT: absolute as well as relative improvement almost negligible -> no large final dhv decrease would we just have evaluated the models gotten from the detectors
-    - this was also the reason we didn't run EAGGA on philippine and gina simply without the detectors (as that's where they crashed) -> evidence suggested that detectors are vital to initial performance
-    - original paper supports this assumption \citep[Fig. 4, p. 545]{EAGGA}
-
-    - preliminary
-  * our implementation of feature + interaction detectors likely tend to include  % belongs in "general algorithm"
-    - more features + interactions for high p datasets
-    - less for low p datasets
-    - remember: original paper samples \# of features included from truncated geometric distribution, what is not mentioned is probability of this distribution
-      is determined from fitting 10 trees + looking at relative \# of features used (cf. their github repo ~/R/TunerEAGGA.R function get\_n\_selected\_rpart())
-    -> mlr3 default decision tree max depth 30
-      * i.e. datasets with p < 30 might use all features, which translates to trunc geom prob = 1
-      * vice-versa for p >> 30 datasets relative \# features < 0.5 (our trunc geom prob)
-      * similar reasoning for pairwise interactions
-  * individuals not in Pareto sets occassionally exhibit AUC < 0.5
-    - this is not by mistake, usually AUC on binary target can simply be inverted (simply predict the opposite class)
-    - for us this cannot easily be inverted due to monotonicity constraints (weights cannot simply be multiplied by -1 if constrained)
-    - could happen e.g. due to AdamW and weight clipping (employed to enforce monotonicity)
-      * AdamW uses momentum
-      * possible scenario: large momentum would yield negative weights, but after epoch they're clipped to [0,infty) + training stops due to early stopping
-      * weight clipping very imperfect way of enforcing monotonicity, but currently in pytorch unfortunately only way to implement this
+% belongs in "general algorithm"
+%    - preliminary
+%  * our implementation of feature + interaction detectors likely tend to include
+%    - more features + interactions for high p datasets
+%    - less for low p datasets
+%    - remember: original paper samples \# of features included from truncated geometric distribution, what is not mentioned is probability of this distribution
+%      is determined from fitting 10 trees + looking at relative \# of features used (cf. their github repo ~/R/TunerEAGGA.R function get\_n\_selected\_rpart())
+%    -> mlr3 default decision tree max depth 30
+%      * i.e. datasets with p < 30 might use all features, which translates to trunc geom prob = 1
+%      * vice-versa for p >> 30 datasets relative \# features < 0.5 (our trunc geom prob)
+%      * similar reasoning for pairwise interactions
 
@@
   \caption{Pareto set loss of all datasets, evaluated on early stopping set. Same colours denote losses coming from folds of the same individual.
-            x-axis portrait epochs, y-axis binary cross entropy loss.}
+            x-axes portrait epochs, y-axes binary cross entropy loss.}
   \label{fig-es-losses}
@@
   \caption{Dominated hypervolume over generations, evaluated on validation set.
-            x-axis portrait generations, y-axis dominated hypervolume using mean AUC over folds.}
+            x-axes portrait generations, y-axes dominated hypervolume using mean AUC over folds.}
   \label{fig-val-dhvs}

### 2025-03-15 paper/main.tex
conclusion

@@
 \section{EAGGA}
+\label{sec-eagga}
 EAGGA, as proposed by \citet{EAGGA}, is an NSGA-II inspired algorithm that optimizes performance along with three interpretability objectives.
@@
 From our anecdotal evidence, the variations in rankings in these preliminary experiments only concerned few samples that were assigned to alternating
-neighbouring fronts, which seemed reasonable not to be concerned over.  % ok formulation?
+neighbouring fronts, hence we considered it tolerable.
 \\
@@
 \section{Conclusion and Future Outlook}
-(Conclusion)
-- tabular data still difficult discipline for neural networks
-- research suggests heavy regularisation to improve performance on tabular data
-- EAGGA proved to be successful in making XGBoost more interpretable while keeping performance on par with unregularised XGBoost
-- we extend EAGGA to neural networks to see if we can utilise the regularisation induced by it to make NNs both interpretable and performant on tabular data
-- as consequence propose new architecture allowing to model equivalence relations of EAGGA
-- ... conforming to \citet[chap. 2]{survey_NN_interpretability} taxonomy it fits as ..., ..., ... %TODO
-- found overall performance comparable to that of XGBoost fitted using EAGGA, which is a plus, but no outperformance
+% Conclusion
+The EAGGA framework has been shown to be able to preserve performance while increasing interpretabiltiy of XGBoost models on tabular data, which has long been 
+a challenging discipline for neural networks. Following recent research suggesting that regularisation can improve
+performance \citep[8]{NEURIPS2021_c902b497}, we extended EAGGA \citep{EAGGA} to neural networks, utilising its regularising properties to achieve similar
+performance of neural networks as performance achieved with XGBoost and simultaneously increase interpretability of this notoriously uninterpretable model class.
+As a consequence, we propose a new architecture allowing to model the equivalence relations of the EAGGA framework.
+We create a model whose interpretation is active, attributional, and global in the framework of \citet[chap. 2]{survey_NN_interpretability}, and find
+overall performance, as measured on the dominated hypervolume of (AUC, NF, NI, NNM) to be comparable to that of XGBoost fitted usign EAGGA.
 
 (Future Outlook)
-- MO BO on group structure space possible, perhaps via restricted BO?
+- due to high complexity of EAs, a MOBO implementation on augmented space (cf section \ref{sec-eagga}) would be very nice
+- theoretical idea: use surrogate model with prior computed based on feasible combinations of equivalence classes
+  * high value, 0 uncertainty around combinations that make no sense
+    - e.g. those with a feature in multiple groups
+    - e.g. those with two features in one group, one having monot incr, one monot decr
+  * high uncertainty around combinations that would make sense to encourage exploring these regions as with GPs
+  * this would likely prove to be computationally very expensive for higher numbers of features
+- as more practical alternative: unrestricted MO BO on extended search space, cf section \ref{sec-eagga}, then acquisiton function taking into
+  account the constraints, e.g. as presented by \citep{10.5555/3020751.3020778}, \citep{perrone2019constrainedbayesianoptimizationmaxvalue}
 
@@
 % needed in general before appendices/bibliography.
-
-%\newpage  % we are not to add page breaks
 

### 2025-03-15 paper/main.tex
future outlook

@@
 \begin{enumerate}[label*=\arabic*.]
-  \item Fitting a single surrogate model on scalarized objectives. \citet[pp. 54-56]{ParEGO} propose \textit{ParEGO}, which employs the augemented Tchebycheff function as
+  \item Fitting a single surrogate model on scalarized objectives. \citet[pp. 54-56]{ParEGO} propose \textit{ParEGO}, which employs the augmented Tchebycheff function as
         scalarization to ensure the Pareto front is explored sufficiently.
@@
 The EAGGA framework has been shown to be able to preserve performance while increasing interpretabiltiy of XGBoost models on tabular data, which has long been 
-a challenging discipline for neural networks. Following recent research suggesting that regularisation can improve
-performance \citep[8]{NEURIPS2021_c902b497}, we extended EAGGA \citep{EAGGA} to neural networks, utilising its regularising properties to achieve similar
-performance of neural networks as performance achieved with XGBoost and simultaneously increase interpretability of this notoriously uninterpretable model class.
+a challenging discipline for neural networks. Following recent research suggesting that regularisation can improve NN performance on
+tabular data \citep[8]{NEURIPS2021_c902b497}, we extend EAGGA \citep{EAGGA} to neural networks, utilising its regularising properties to achieve similar
+performance with neural networks as with XGBoost and simultaneously increase interpretability of this notoriously uninterpretable model class.
 As a consequence, we propose a new architecture allowing to model the equivalence relations of the EAGGA framework.
 We create a model whose interpretation is active, attributional, and global in the framework of \citet[chap. 2]{survey_NN_interpretability}, and find
-overall performance, as measured on the dominated hypervolume of (AUC, NF, NI, NNM) to be comparable to that of XGBoost fitted usign EAGGA.
+overall performance, as measured on the dominated hypervolume of (AUC, NF, NI, NNM) to be comparable to that of XGBoost fitted using EAGGA.
+\\
 
-(Future Outlook)
-- due to high complexity of EAs, a MOBO implementation on augmented space (cf section \ref{sec-eagga}) would be very nice
-- theoretical idea: use surrogate model with prior computed based on feasible combinations of equivalence classes
-  * high value, 0 uncertainty around combinations that make no sense
-    - e.g. those with a feature in multiple groups
-    - e.g. those with two features in one group, one having monot incr, one monot decr
-  * high uncertainty around combinations that would make sense to encourage exploring these regions as with GPs
-  * this would likely prove to be computationally very expensive for higher numbers of features
-- as more practical alternative: unrestricted MO BO on extended search space, cf section \ref{sec-eagga}, then acquisiton function taking into
-  account the constraints, e.g. as presented by \citep{10.5555/3020751.3020778}, \citep{perrone2019constrainedbayesianoptimizationmaxvalue}
+% Future Outlook
+Due to the low sample-efficiency of evolutionary algorithms, a multi-objective Bayesian Optimization variant on the augmented search space
+(refer Section \ref{sec-eagga}) could deliver the strongest training time improvement.
+There are two avenues to explore this, either via a change to the surrogate model or by adapting the acquisition function.
+A surrogate model using a custom computed prior with high values and no uncertainty in unfeasible regions and regular uncertainty in feasible regions could be a solution.
+An unfeasible region is for instance one, where a feature is in multiple equivalence classes, or where two features in the same equivalence class have opposing
+monotonicity constraints.
+The second option, adapting the acquisition function, may prove to be computationally less complex for datasets with many features.
+This would mean using an unrestricted surrogate model and then taking account the constraints in the acquisition function. Similar approaches have been
+tried on unrelated problems by \citep{10.5555/3020751.3020778} and \citep{perrone2019constrainedbayesianoptimizationmaxvalue}.
 

### 2025-03-15 paper/main.tex
abstract

@@
 \begin{abstract}%   <- trailing '%' for backward compatibility of .sty file
-%\blindtext
-- tabular data still difficult for NNs, where it's still outperformed by other ML model classes
-- research suggests NN performance benefits from heavy regularisation
-- using EAGGA, we regularise an NN and achieve both comparable performance as XGB on EAGGA as well as interpretability
-- for this, we propose a network architecture specifically suited for the EAGGA algorithm
+  Tabular data proves to be difficult for neural networks, where they are still outperformed by other model classes.
+  Following recent research, regularisation can improve their performance. Incorporating the EAGGA framework designed to
+  enforce explainability, we improve interpretability and achieve performance comparable to that of XGBoost on the
+  metrics AUC, number of features, number of pairwise interactions, and number of non-monotone feature effects.
+  For this, we propose a sparse, regularised network architecture specifically suited for EAGGA's constraints on
+  interaction and monotonicity.
 \end{abstract}
@@
 \begin{keywords}
-  tabular data, multi-objective optimization, interpretability, deep learning
+  tabular data, multi-objective optimization, interpretability, neural networks
 \end{keywords}
@@
 \section{Introduction}
-Tabular Data
-- most common type of data
-- still difficult for neural networks
-- \citep[p. 7499]{Borisov_2024}
-- recent research suggests that strong regularisation is beneficial to NN performance on tab data \citep[8]{NEURIPS2021_c902b497}
-
-- we know regularisation from linear models can come with improvements in interpretabiltiy
-  * e.g. LM-LASSO (L1) regularisation as feature selection -> reduces \# features used in model by setting some coeffs to 0 \citep[p. 267]{lasso}
-- other forms of regularisation improve performance
-  * e.g. LM-Ridge (L2) on multi-collinear data \citep[p. 55]{ridge}  % TODO: mention in "extension" section that L2 is implemented via weight decay
-  * e.g. NN dropout, reduces co-adaptation \citep[p. 1]{hinton2012improvingneuralnetworkspreventing}
-  * e.g. NN early stopping, reduces overfitting on training data \citep[p. 778f]{early_stopping}
-
-- we want to explore if we can use NN regularisation to tackle both interpretability and improved performance (i.e. ``comparable'' to XGBoost) on tab data
-- using EAGGA framework, which proved it can improve performance while keeping (already high) performance of XGBoost on tab data -> see if interpretabiltiy
-  improvements translate to NN + performance can also be on par
-
-%\blindmathpaper
+Regularisation has been extensively shown to improve performance from linear models to neural networks.
+One of the most popular types of regularisation, Ridge or L2-regularisation is still the preferred method for handling
+multicollinear data in linear models \citep[p. 55]{ridge}.
+Practicioners training neural networks regularly use regularisation methods such as dropout to reduce
+co-adaptation \citep[p. 1]{hinton2012improvingneuralnetworkspreventing} or early stopping to avoid overfitting on the training data \citep[p. 778f]{early_stopping}.
+\\
+Some regularisation can even increase interpretability, such as LASSO, or L1-regularisation for linear models, which reduces the number of
+features used to fit the model by setting some coefficients to 0 \citep[p. 267]{lasso}.
+\\
+Neural networks as a model class have achieved impressive performance on a multitude of problems, yet still struggle with the most common format --
+tabular data \citep[p. 7499]{Borisov_2024}.
+Recent research, however, suggest that strong regularisation is beneficial to a neural network's performance on tabular data \citep[8]{NEURIPS2021_c902b497}.
+\citet{EAGGA} propose a framework to increase the interpretability of XGBoost using regularisation.
+In this report, we explore if the same framework can be extended to neural networks to tackle both interpretability and achieve performance on tabular data
+that is on par with XGBoost.
 
@@
 GDPR \citep[p. 1]{review_NN_interpretability} or for approval of drugs discovered using machine learning models \citep[1B]{survey_NN_interpretability}.
-It can further prove helpful for explaining unexpected drops in model performance, which could arise from optimising a model with respect to loss and then judging its
+It can further prove helpful for explaining unexpected drops in model performance, which could arise from optimizing a model with respect to loss and then judging its
 performance on a different metric such as accuracy, a practice known as \textit{model debugging} \citep[1B]{survey_NN_interpretability} or for
@@
 
-Another popular EA approach is Differential Evolution. Its populaton is randomly initialised so the entire search space is covered.
+Another popular EA approach is Differential Evolution. Its populaton is randomly initialized so the entire search space is covered.
 In each generation $g$, for each $\boldsymbol\lambda_{i,g}$ with $i=1,...,n$ it creates ``mutant vectors''
@@
 the group structure space $\mathcal{G}$, with adapted mutation and crossover operators.
-The authors further introduce a special initialisation of the group structures to increase the sample-efficiency of their algorithm compared to random initialisation.
+The authors further introduce a special initialization of the group structures to increase the sample-efficiency of their algorithm compared to random initialization.
 \citep[pp. 541-543]{EAGGA}
@@
 \subsection{Overall Algorithm}
-Our NNs use the following hyperparameters initialised as described
+Our NNs use the following hyperparameters initialized as described
 \begin{itemize}
-  \item Total number of layers $\in\{3, ..., 10\}$, initialised from a truncated Geometric distribution with $\pi=0.5$.
+  \item Total number of layers $\in\{3, ..., 10\}$, initialized from a truncated Geometric distribution with $\pi=0.5$.
         This includes input and output layers, the number of hidden layers is thus $\in\{1, ..., 8\}$
-  \item Number of nodes per hidden layer $\in\{3, ..., 20\}$, initialisation from a truncated Geometric distribution with $\pi=0.5$
-  \item Dropout probability $\in[0, 1]$, initialised from a truncated Gamma distribution with shape 2 and scale 0.15
+  \item Number of nodes per hidden layer $\in\{3, ..., 20\}$, initialization from a truncated Geometric distribution with $\pi=0.5$
+  \item Dropout probability $\in[0, 1]$, initialized from a truncated Gamma distribution with shape 2 and scale 0.15
 \end{itemize}
@@
 \\
-Examining the Pareto sets returned from training, which are accessible on our GitHub repository at ~/code/export/*.csv, we can observe a clear trend in the
+Examining the Pareto sets returned from training, which are accessible on our GitHub repository\footnote{\url{https://github.com/simon29st/ws2425-tab-ml/}}
+at ~/code/export/*.csv, we can observe a clear trend in the
 Pareto sets' models' hyperparameters:
@@
 well as relative improvement, however, is almost negligible, suggesting there to be no large decrease in dominated hypervolume, would we have just evaluated
-the group structure initialised by the detectors.
-This was also the reason behind not running EAGGA on the philipine and gina datasets on randomly initialised detectors (as the detectors for those datasets
+the group structure initialized by the detectors.
+This was also the reason behind not running EAGGA on the philipine and gina datasets on randomly initialized detectors (as the detectors for those datasets
 ran out of memory due to the high feature cardinality), as the detectors prove vital for the initial performance. This finding is also in
@@
 We acknowledge that weight clipping is a very imperfect way of enforcing monotonicity, but currently seems to be the only way to implement this type of
-constrained optimisation in pytorch.
+constrained optimization in pytorch.
 

### 2025-03-15 paper/main.tex
remove todos

@@
 \citet{EAGGA} propose a framework to increase the interpretability of XGBoost using regularisation.
-In this report, we explore if the same framework can be extended to neural networks to tackle both interpretability and achieve performance on tabular data
-that is on par with XGBoost.
+In this report, we explore if the same framework can be extended to neural networks, a model class that in its fully-connected feedforwad form is
+inherently non-interpretable, to tackle both explainability and achieve performance on tabular data that is on par with XGBoost.
 
@@
 % Taxonomy
-\citet[chap. 2]{survey_NN_interpretability} extend this distinction to a three-dimensional taxonomy, allowing for better categorisation of neural networks (NNs),
-a model class that in its fully-connected feedforwad form is inherently non-interpretable. % TODO: letztes halbsatz vllt lieber in intro unterbringen, hier ist der awkward und unnötig ausschmückend
+\citet[chap. 2]{survey_NN_interpretability} extend this distinction to a three-dimensional taxonomy, allowing for better categorisation of neural networks (NNs).
 \\
@@
 of the target function from previous evaluations $(\boldsymbol\lambda_{1:t},f(\boldsymbol\lambda_{1:t}))$. \citep[p. 7]{frazier2018tutorialbayesianoptimization}
-\begin{equation}  % TODO: take out
+\begin{equation}
   f^{(t)}\sim\tilde{f}_t
@@
 Commonly chosen models are Gaussian Processes (Eq. \ref{eq-gaussian-process}) and Random Forests.
-\begin{equation}  % TODO: make inline or provide further information
+\begin{equation}
   \text{GP}(m(\boldsymbol\lambda), k(\boldsymbol\lambda,\boldsymbol\lambda'))
@@
 entire space.
-In our extension of the EAGGA approach we don't employ NAS as research focusses on the NLP and image domains, % TODO: find some more references
+In our extension of the EAGGA approach we don't employ NAS as research focusses on the NLP and
+image domains \citep{baker2017designingneuralnetworkarchitectures},  \citep{10.5555/3504035.3504375},  \citep{10.1145/3071178.3071229},
 where features (i.e. tokens or pixels, respectively) exhibit strong correlation amongst themselves, which is usually not the case for tabular data \citep[p. 7499]{Borisov_2024}.
@@
   \includegraphics[width=\linewidth]{./architecture.png}  % scale figure, cf. https://tex.stackexchange.com/a/16584
-  \caption{TODO: caption}
+  \caption{The EAGGA network architecture for a dataset with features A, B, C, D, E, F, G. The group structure might look as follows: $(\{D,G\}, (\{A,B,C\}, 1), (\{E,F\}, 0))$}
   \label{fig-nn-architecture}
@@
   \includegraphics[width=\linewidth]{./dataset_split.png}
-  \caption{TODO: caption}
+  \caption{Visualisation of the data set split. Outer holdout split (2/3), mid 5-fold CV split, inner holdout split (80\%)}
   \label{fig-dataset-split}
@@
 \section{Software used}
-for implementation we used
-openml \cite{OpenML}, \cite{OpenMLPython},
-numpy \cite{numpy},
-pandas \cite{pandas1}, \cite{pandas2},
-pytorch \cite{PyTorch},
-scikit-learn \cite{scikit-learn},
-scipy \cite{SciPy},
-pymoo \cite{pymoo}, and
-tqdm \cite{tqdm}
+For the Python implementation of EAGGA we use OpenML \citep{OpenML}, \citep{OpenMLPython},
+Numpy \citep{numpy},
+pandas \citep{pandas1}, \citep{pandas2},
+PyTorch \citep{PyTorch},
+scikit-learn \citep{scikit-learn},
+SciPy \citep{SciPy},
+pymoo \citep{pymoo}, and
+tqdm \citep{tqdm}.
 

### 2025-03-15 paper/main.tex
check references

@@
 tabular data \citep[p. 7499]{Borisov_2024}.
-Recent research, however, suggest that strong regularisation is beneficial to a neural network's performance on tabular data \citep[8]{NEURIPS2021_c902b497}.
+Recent research, however, suggest that strong regularisation is beneficial to a neural network's performance on tabular data \citep[p. 8]{NEURIPS2021_c902b497}.
 \citet{EAGGA} propose a framework to increase the interpretability of XGBoost using regularisation.
@@
 where \textit{explanations} are logical decision rules and \textit{understandable terms} relate to commonly used terms in the domain of the problem,
-as suggested by \citet[chap. 1]{survey_NN_interpretability}. Further, we use the term ``explainability'' in an exchangeable manner with ``interpretability'',
+as suggested by \citet[p. 726]{survey_NN_interpretability}. Further, we use the term ``explainability'' in an exchangeable manner with ``interpretability'',
 as is commonly done.
@@
 % Why Interpretability is desirable?
-Explainability of a model's reasoning is in many ways desirable. \citet[pp. 3-4]{Zach2019InterpretabilityOD} gives a range of examples,
+Explainability of a model's reasoning is in many ways desirable. \citet[pp. 3f]{Zach2019InterpretabilityOD} gives a range of examples,
 amongst which are \textit{gaining trust}, e.g. when doctors rely on medical diagnosis predictions, avoiding \textit{subconcious biases} by
 making sure loan approvals are non-discriminatory, or \textit{regulatory} reasons, most notably the EU's ``Right to Explanation'' warranted by the
-GDPR \citep[p. 1]{review_NN_interpretability} or for approval of drugs discovered using machine learning models \citep[1B]{survey_NN_interpretability}.
+GDPR \citep[p. 1]{review_NN_interpretability} or for approval of drugs discovered using machine learning models \citep[pp. 727f]{survey_NN_interpretability}.
 It can further prove helpful for explaining unexpected drops in model performance, which could arise from optimizing a model with respect to loss and then judging its
-performance on a different metric such as accuracy, a practice known as \textit{model debugging} \citep[1B]{survey_NN_interpretability} or for
+performance on a different metric such as accuracy, a practice known as \textit{model debugging} \citep[pp. 727f]{survey_NN_interpretability} or for
 \textit{scientific understanding} in domains where only models can make sense of increasingly complex data anymore and learnt knowledge encoded in a model needs to be
@@
 % Taxonomy
-\citet[chap. 2]{survey_NN_interpretability} extend this distinction to a three-dimensional taxonomy, allowing for better categorisation of neural networks (NNs).
+\citet[pp. 728f]{survey_NN_interpretability} extend this distinction to a three-dimensional taxonomy, allowing for better categorisation of neural networks (NNs).
 \\
@@
 % Evaluation
-Given its unclear definition, evaluating interpretability can be challenging. \citet[3]{DoshiVelez2017TowardsAR} propose a taxonomy to categorise
+Given its unclear definition, evaluating interpretability can be challenging. \citet[pp. 4-6]{DoshiVelez2017TowardsAR} propose a taxonomy to categorise
 possible evaluation methods based on their rigorousness.
@@
 as for a given budget $B$, each HP will likely be queried with $B$ different values, whereas for grid search each HP will only be queried with $B^{1/N}$
-different values for $N$ HPs. \citep[chap. 1.3]{feurer_hyperparameter_2019}
+different values for $N$ HPs. \citep[pp. 7f]{feurer_hyperparameter_2019}
 
@@
 population and offspring (``$(\mu+\lambda)$-selection''), which is also called an ``elitist'' strategy, as it's guaranteed to keep the best individual.
-\citep[chap. 1.3]{feurer_hyperparameter_2019}
+\citep[p. 188]{simon2013evolutionary}
 
@@
 and the covariance matrix similarly is the weighted covariance of the previous generation.
-The weighing scheme for both is done in a way as to reproduce previously successful (i.e. selected) steps \citep[p. 8-11]{hansen2023cmaevolutionstrategytutorial}.
+The weighing scheme for both is done in a way as to reproduce previously successful (i.e. selected) steps \citep[pp. 8-11]{hansen2023cmaevolutionstrategytutorial}.
 
@@
 The acquisition function then retrieves the highest utility point from the surrogate model and evaluates it on the target function, after which it adds the point
-to the set of queried points \citep[chap. 1.3.2]{feurer_hyperparameter_2019} and \citep[pp. 2f]{frazier2018tutorialbayesianoptimization}.
+to the set of queried points \citep[p. 9]{feurer_hyperparameter_2019} and \citep[pp. 2f]{frazier2018tutorialbayesianoptimization}.
 Evidently, neither the target function nor the surrogate model are optimized directly. Rather, the acquisition function trades-off exploration and exploitation of
@@
 A drawback of Gaussian Processes is its poor scalability in number of data points and number of hyperparameters, although there exist workarounds approximating the full GP
-with only a subset of the samples. \citep[chap. 1.3.2]{feurer_hyperparameter_2019}
+with only a subset of the samples. \citep[pp. 10f]{feurer_hyperparameter_2019}
 \\
 \textbf{Random Forests}, unlike GPs, can handle complex hyperparameter spaces even including categorical and hierarchical HPs. Their computational complexity is also far
-less, making them a popular alternative to Gaussian Processes. \citep[chap. 1.3.2]{feurer_hyperparameter_2019}
+less, making them a popular alternative to Gaussian Processes. \citep[p. 12]{feurer_hyperparameter_2019}
 
@@
 where features (i.e. tokens or pixels, respectively) exhibit strong correlation amongst themselves, which is usually not the case for tabular data \citep[p. 7499]{Borisov_2024}.
-Still, we want to outline two notable approaches for neural architecture search in this paper, one of which are \textbf{cell search spcaces}.
+Still, we want to outline two notable approaches for neural architecture search in this paper, one of which are \textbf{cell search spaces}.
 These modularise an NN into a chain-structure of cells, where each cell is a basic building block of the respective architecture. For regular feedforwad neural networks,
@@
 Instead of tuning every hyperparameter, the sequential placement of the cells is then optimzed during NAS, e.g. via random search or Bayesian
-Optimization. \citep[chap. 3.2]{elsken_neural_2019}
+Optimization. \citep[p. 67]{elsken_neural_2019}
 \\
@@
 This has the advantage of, albeit being slower than training a single model, being far more efficient and less expensive than training all architectures encoded
-in the fabric. \citep[pp. 1-2, p.8]{saxena2017convolutionalneuralfabrics}
+in the fabric. \citep[pp. 1-2, p. 8]{saxena2017convolutionalneuralfabrics}
 % don't necessarily talk about co-adaptation problem
@@
 Weights are chosen a-priori by the user and the solution is very sensitive to them, making this method very reliant on different
-users' preferences. \citep[p. 11]{10.1145/3610536} and \citep[chap. 3.1]{NSGA}
+users' preferences. \citep[p. 11]{10.1145/3610536} and \citep[pp. 223f]{NSGA}
 \\
@@
       \caption{Mean dominated Hypervolume of $EAGGA_{\text{XGBoost}}$, $EAGGA_{\text{XGBoost}_{\text{md2}}}$, and union of competitors averaged over
-      10 replications. Bars represent standard errors. \citep[Figure 2, p. 544]{EAGGA}}
+      10 replications. Bars represent standard errors. Figure taken from \citep[Figure 2, p. 544]{EAGGA}}
       \label{fig-eagga-xgb-dhvs}
@@
 \end{figure}
-As evident from Figures \ref{fig-test-dhvs}, \ref{fig-eagga-xgb-dhvs}, applying EAGGA to neural networks yields comparable performance to applying it to XGBoost,
-both with and without restricting the maximum tree depth.
+As evident from Figures \ref{fig-test-dhvs}, \ref{fig-eagga-xgb-dhvs}, applying EAGGA to neural networks yields similar or slightly worse performance compared to
+to applying it to XGBoost, both with and without restricting the maximum tree depth.
 A direct comparison to unrestricted neural networks trained with regular stochastic gradient descent has not been made, as this would have led to a hypervolume of 0
@@
 
-% belongs in "general algorithm"
-%    - preliminary
-%  * our implementation of feature + interaction detectors likely tend to include
-%    - more features + interactions for high p datasets
-%    - less for low p datasets
-%    - remember: original paper samples \# of features included from truncated geometric distribution, what is not mentioned is probability of this distribution
-%      is determined from fitting 10 trees + looking at relative \# of features used (cf. their github repo ~/R/TunerEAGGA.R function get\_n\_selected\_rpart())
-%    -> mlr3 default decision tree max depth 30
-%      * i.e. datasets with p < 30 might use all features, which translates to trunc geom prob = 1
-%      * vice-versa for p >> 30 datasets relative \# features < 0.5 (our trunc geom prob)
-%      * similar reasoning for pairwise interactions
-
 \section{Conclusion and Future Outlook}
@@
 As a consequence, we propose a new architecture allowing to model the equivalence relations of the EAGGA framework.
-We create a model whose interpretation is active, attributional, and global in the framework of \citet[chap. 2]{survey_NN_interpretability}, and find
+We create a model whose interpretation is active, attributional, and global in the framework of \citet[pp. 3f]{survey_NN_interpretability}, and find
 overall performance, as measured on the dominated hypervolume of (AUC, NF, NI, NNM) to be comparable to that of XGBoost fitted using EAGGA.

### 2025-03-15 paper/main.tex
last improvements

@@
   Following recent research, regularisation can improve their performance. Incorporating the EAGGA framework designed to
-  enforce explainability, we improve interpretability and achieve performance comparable to that of XGBoost on the
-  metrics AUC, number of features, number of pairwise interactions, and number of non-monotone feature effects.
+  enforce explainability, we improve interpretability and achieve performance comparable to that of XGBoost on
+  dominated hypervolume on the metrics AUC, number of features, number of pairwise interactions, and number of non-monotone feature effects.
   For this, we propose a sparse, regularised network architecture specifically suited for EAGGA's constraints on
@@
 \section{Introduction}
-Regularisation has been extensively shown to improve performance from linear models to neural networks.
+Regularisation has been extensively shown to improve performance from linear models to neural networks (NNs).
 One of the most popular types of regularisation, Ridge or L2-regularisation is still the preferred method for handling
@@
 \textit{scientific understanding} in domains where only models can make sense of increasingly complex data anymore and learnt knowledge encoded in a model needs to be
-made accessible to humans to be used reliably \citep[p. 1]{review_NN_interpretability}.
+made accessible to humans in order to be used reliably \citep[p. 1]{review_NN_interpretability}.
 
@@
 % Taxonomy
-\citet[pp. 728f]{survey_NN_interpretability} extend this distinction to a three-dimensional taxonomy, allowing for better categorisation of neural networks (NNs).
+\citet[pp. 728f]{survey_NN_interpretability} extend this distinction to a three-dimensional taxonomy, allowing for better categorisation of neural networks.
 \\
@@
 \\
-\textbf{Human-grounded metrics} is similar to application-grounded evaluation in that a human still evaluates the interpretations, but tries to simplify the task
+\textbf{Human-grounded metrics} are similar to application-grounded evaluation in that a human still evaluates the interpretations, but tries to simplify the task
 so that a layperson can do it. Human-grounded approaches are especially suitable if it's sufficient to validate the general concepts of a task. A typical evaluation
-set-up in this category is binary forced choice, where the human evaluator chooses, which of two generated explanations he prefers.
+set-up in this category is \textit{binary forced choice}, where the human evaluator chooses, which of two generated explanations they prefer.
 \\
-\textbf{Functionally-grounded evaluation} is the least rigorous, but easiest to implement of the three. It assess explanatory quality according to some
+\textbf{Functionally-grounded evaluation} is the least rigorous, but easiest to implement of the three. It assesses explanatory quality according to some
 formally defined proxy for interpretability and is particularly useful in ranking different models if their model-class is already identified
@@
 be an algorithm that maps a given dataset $\mathcal{D}$ and HP configuration $\boldsymbol\lambda\in\boldsymbol\Lambda$ to a model $\hat{f}$.
-Denote with $\mathcal{I}_{\boldsymbol\lambda}$ an algorithm with fixed HP configuration a chosen loss function with $L$.
+Denote with $\mathcal{I}_{\boldsymbol\lambda}$ an algorithm with fixed HP configuration $\boldsymbol\lambda$ and a chosen loss function with $L$.
 The goal of model training is, for fixed $\boldsymbol\lambda$, to have $\mathcal{I}_{\boldsymbol\lambda}$ find model $\hat{f}$ minimising the expected generalisation error
@@
 The most basic model-free strategies for blackbox optimization are grid and random search.
-In the latter, the user defines a range of interest for each HP and then evaluating points along a grid within the cartesian product of those.
+In the latter, the user defines a range of interest for each HP and then evaluates points along a grid within the cartesian product of those.
 This has the drawback of scaling extremely poorly in both number of HP dimensions and number of query points per range of interest.
 Random search, on the other hand, randomly samples a value for each HP until it runs out of budget. Its explorative nature, ease of use, and no assumptions
-about the model makes it a good baseline. It is also superior to grid search in cases were one or more HPs have little impact on model performance,
+about the model make it a good baseline. It is also superior to grid search in cases were one or more HPs have little impact on model performance,
 as for a given budget $B$, each HP will likely be queried with $B$ different values, whereas for grid search each HP will only be queried with $B^{1/N}$
@@
 \textbf{Crossover} selects two ``parents'' from the pool of reproducing individuals. An index $k$ is sampled at random and all HP values from
-index $k$ on are swapped, yielding two new ``children'' individuals.
+index $k$ on are swapped, yielding two new ``child'' individuals.
 \\
-\textbf{Mutation} randomly changes single HP values of an individual chosen for reproduction, e.g. by adding Gaussian noise to real values or flipping bits
+\textbf{Mutation} randomly changes some HP values of an individual chosen for reproduction, e.g. by adding Gaussian noise to real values or flipping bits
 on binary values.
@@
 $\boldsymbol\nu_{i,g+1}=\boldsymbol\lambda_{r_1,g}+F\cdot(\boldsymbol\lambda_{r_2,g}-\boldsymbol\lambda_{r_3,g})$,
-where $r_1,r_2,r_3\in\{1,...,n\}$ are mutually different random indices that also different from $i$ and $F\in[0,2]$ is constant.
+where $r_1,r_2,r_3\in\{1,...,n\}$ are mutually different random indices that also differ from $i$ and $F\in[0,2]$ is constant.
 For crossover, it then generates a ``trial vector'' $\boldsymbol\upsilon_{i,g+1}=(\upsilon_{i,g+1}^{(1)},\upsilon_{i,g+1}^{(2)},...,\upsilon_{i,g+1}^{(N)})^T$ with
@@
 \end{equation}
-for a random index $R\in\{1,2,...,D\}$, sampled $u\sim U[0,1]$, and crossover constant $\text{CR}\in[0,1]$.
+for a random index $R\in\{1,2,...,N\}$, sampled $u\sim U[0,1]$, and crossover constant $\text{CR}\in[0,1]$.
 Selection is done by picking the better of $\boldsymbol\upsilon_{i,g+1}$ and $\boldsymbol\lambda_{i,g}$ with respect to fitness. \citep[p. 343]{differential_evolution}
@@
 each other in HP space $\boldsymbol\Lambda$ exhibit stronger correlation than those far apart.
-The kernel function can be freely chosen, a common choices is the Gaussian (Eq. \ref{eq-kernel-gaussian}) kernel with
+The kernel function can be freely chosen, a common choice is the Gaussian (Eq. \ref{eq-kernel-gaussian}) kernel with
 hyperparameter $\alpha_0$. \citep[p. 5]{frazier2018tutorialbayesianoptimization}
@@
 \end{equation}
-A drawback of Gaussian Processes is its poor scalability in number of data points and number of hyperparameters, although there exist workarounds approximating the full GP
+A drawback of Gaussian Processes is their poor scalability in number of data points and number of hyperparameters, although there exist workarounds approximating the full GP
 with only a subset of the samples. \citep[pp. 10f]{feurer_hyperparameter_2019}
@@
 Still, we want to outline two notable approaches for neural architecture search in this paper, one of which are \textbf{cell search spaces}.
-These modularise an NN into a chain-structure of cells, where each cell is a basic building block of the respective architecture. For regular feedforwad neural networks,
+These modularise an NN into a chain-structure of cells, where each cell is a basic building block of the respective architecture. For feedforwad neural networks,
 this could be a linear layer of fixed size and with a specific activation function,
@@
 In neither case do we know the impact our a-priori trade-off has on the final performance.
-In practical applications, the impact on performance across all objectives is what matters.
+In practical applications, however, the impact on performance across all objectives is what matters.
 This is the main advantage of a-posteriori methods: instead of implicitly defining a trade-off prior to training, we still evaluate multiple HP configurations
@@
 as the mean side length of the cuboid spanned by its nearest neighbours as vertices in the objective space, where individuals without two neighbours
-along some dimension are assigned an infinitely high distance value. The less the crowding distance, the more an area is by other individuals. \citep[p. 185]{NSGA_II}
+along some dimension are assigned an infinitely high distance value. The less the crowding distance, the more an area is crowded with other individuals. \citep[p. 185]{NSGA_II}
 \\
@@
 \\
-Conceptually, the reasoning behind using non-dominated sorting to rank individuals is intuitive: the closer an individuals front is to the Pareto front, i.e.
-the smaller its front rank, the better it approaches the true Pareto front.
+Conceptually, the reasoning behind using non-dominated sorting to rank individuals is intuitive: the closer an individual's front is to the Pareto front, i.e.
+the smaller its front rank, the closer it is to the true Pareto front.
 The usage of crowding distance, on the other hand, is not as straightforward. To understand it, we recall the goal of MOO, which is not merely optimizing
@@
 Considering now the opposite, having multiple areas making up the Pareto front estimate, makes it stable: removing one area would not impact the estimate a lot,
-as it's still held up by other individuals. \citep[p. 185, pp. 189-192]{genetic_algos}
+as it's still held up by other individuals \citep[p. 185, pp. 189-192]{genetic_algos}.
 A possible proxy for exploration of the Pareto front is the hypervolume contribution. The hypervolume contribution of individual
@@
 This induces a group structure of equivalence classes, where features allowed to interact belong to the same class and share the same monotonicity attribute.
-Using this group structure, they define a much simpler ``augmented serach space'' $\tilde{\boldsymbol\Lambda}=\boldsymbol\Lambda\times\mathcal{G}$ comprising of
+Using this group structure, they define a much simpler ``augmented search space'' $\tilde{\boldsymbol\Lambda}=\boldsymbol\Lambda\times\mathcal{G}$ comprising of
 the cartesian product of the model hyperparameter space $\boldsymbol\Lambda$ and the group structure space $\mathcal{G}$.
@@
 \\
-The feature detector is as described by \citet[p. 542]{EAGGA}, but instead of fitting ten trees and using the relative number of features as
+The feature detector is as described by \citet[p. 542]{EAGGA}, but instead of fitting 10 trees and using the relative number of features as
 probability for the truncated Geometric distribution, we use a constant value of 0.5, as the sklearn DecisionTree examination is not straightforward.
@@
 The interaction detector similarly uses a constant probability of 0.5 to sample the number of interactions to be used instead of the relative number of interactions
-from ten decision trees, for the same reasoning as for the feature detector. Also, for the FAST algorithm, when computing the interaction scores we don't
+from 10 decision trees, for the same reasoning as for the feature detector. Also, for the FAST algorithm, when computing the interaction scores we don't
 use the residual sum of squares but the mean accuracy, as the datasets all have binary targets and FAST thus fits a logistic instead of a linear regression.
 \\
-The monotonicty detector is implemented as described by \citet[p. 543]{EAGGA}, using the default decision tree hyperparameters from mlr3 in Python.
+The monotonicty detector is implemented as described by \citet[p. 543]{EAGGA}, using the default decision tree hyperparameters from mlr3.
 \\
@@
 \\
-In rare cases (anecdotally once every ~10 datasets), we observed our monotonicity detector generating a group monotonicty attribute of -1, which crashed the computation.
+In rare cases (anecdotally, once every $\sim$10 datasets), we observed our monotonicity detector generating a group monotonicty attribute of -1, which crashed the computation.
 For these cases, we implemented a function to import a previous generation of individuals to be able to pick up from where the training crashed and not train anew.
@@
 Examining the Pareto sets returned from training, which are accessible on our GitHub repository\footnote{\url{https://github.com/simon29st/ws2425-tab-ml/}}
-at ~/code/export/*.csv, we can observe a clear trend in the
+at $\sim$/code/export/*.csv, we can observe a clear trend in the
 Pareto sets' models' hyperparameters:
@@
 As the loss graphs remain horizontal and there is no uptick again, this would likely not be detrimental to model performance and the main ``risk'' would be
-more a expensive computation from longer trainings.
+a more expensive computation from longer trainings.
 \\
@@
 the group structure initialized by the detectors.
-This was also the reason behind not running EAGGA on the philipine and gina datasets on randomly initialized detectors (as the detectors for those datasets
+This was also the reason behind not running EAGGA on the philipine and gina datasets on randomly initialized group structures (as the detectors for those datasets
 ran out of memory due to the high feature cardinality), as the detectors prove vital for the initial performance. This finding is also in
@@
 non-negative domain at the end of the epoch.
-We acknowledge that weight clipping is a very imperfect way of enforcing monotonicity, but currently seems to be the only way to implement this type of
+We acknowledge that weight clipping is a very imperfect way of enforcing monotonicity, but it currently seems to be the only way to implement this type of
 constrained optimization in pytorch.
